```{r}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(stringr)
library(dplyr)
library(batchelor)
library(BiocParallel)
library(tidyr)
library(MultiAssayExperiment)
library(lme4)
library(lmerTest)
library(ggrepel)
n_threads <- 32

```

# path to files

```{r}

# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# load each plate into a separate bioconductor object

# standard panel
files_standard <- c(
  paste0(base, "250710 cytokines_cellpainting/MyExpt_Cells_plate1.csv"), 
  paste0(base, "250710 cytokines_cellpainting/MyExpt_Cells_plate2.csv"), 
  paste0(base, "250717 cytokines_cellpainting lot2/MyExpt_Cells_plate1.csv"), 
  paste0(base, "250717 cytokines_cellpainting lot2/MyExpt_Cells_plate2.csv")
  )
sce_standard <- Reduce(cbind, lapply(files_standard, loadData))
sce_standard$Panel <- "standard"

# metabolism panel
files_metabolism <- c(
  paste0(base, "250709 cytokines_metabolism/MyExpt_Cells-plate1.csv"), 
  paste0(base, "250709 cytokines_metabolism/MyExpt_Cells-plate2.csv"), 
  paste0(base, "250717 cytokines_metabolism lot2/MyExpt_Cells_plate1.csv"), 
  paste0(base, "250717 cytokines_metabolism lot2/MyExpt_Cells_plate2.csv")
  )
sce_metabolism <- Reduce(cbind, lapply(files_metabolism, loadData))
sce_metabolism$Panel <- "metabolism"

# stress panel
files_stress <- c(
  paste0(base, "250709 cytokines_stress/MyExpt_Cells.csv"), 
  paste0(base, "250717 cytokines_stress lot2/MyExpt_Cells.csv")
  )
sce_stress <- Reduce(cbind, lapply(files_stress, loadData))
sce_stress$Panel <- "stress"

# merge se objects on common features
sce_list <- list(standard = sce_standard, 
                 metabolism = sce_metabolism, 
                 stress = sce_stress)

```

## NEW Data Transformation for batch effect

Prepare the data for machine learning.

```{r}
data_transformation <- function(sce) {

  # remove missing values in cells
  mat <- assay(sce, name = "features")
  miss_cells <- apply(mat, 2, function(x) sum(is.na(x)))
  cell_ids <- which(miss_cells > 0)
  sce <- sce[, -cell_ids]
  
  # normalize features
  sce <- removeLowVariance(sce)
  sce <- transformLogScale(sce)
  
  # remove outlier images
  sce <- removeOutliers(sce, min = 10, max = 500)
  
  # remove outlier cells
  stats <- perCellQCMetrics(sce, assay.type = "tfmfeatures")
  sce <- sce[, !isOutlier(stats$sum)]
  
  # batch correct
  sce_corrected <- regressBatches(
    sce, batch = sce$Plate, d = 50,
    assay.type = "tfmfeatures",
    BPPARAM = MulticoreParam(workers = n_threads),
    BSPARAM = BiocSingular::RandomParam()
    )
  assay(sce, "tfmfeatures") <- assay(sce_corrected, "corrected")
  
  # fix factors
  sce$Treatment <- as.factor(sce$Treatment)
  sce$Treatment <- relevel(sce$Treatment, ref = "control")
  sce
  
}
sce_list <- lapply(sce_list, data_transformation)
lapply(sce_list, plotCellsPerImage)

```

#Check batch effect for just one feature.

```{r}
tb_sce <- makePerCellDF(sce_list$stress, 
                        assay.type = "tfmfeatures",
                        features = "Intensity_IntegratedIntensity_CorrCytoID"
                        )
tb_sce |> 
  filter(Treatment == "TNF") |>
  ggplot(aes(Intensity_IntegratedIntensity_CorrCytoID, 
             group = Patient, color = as.factor(Plate))) + 
  geom_density() + 
  facet_grid(Disease ~ Treatment)

```

For Multiple model

```{r}

interest <- setdiff(unique(sce$Treatment), c("control"))
result_list <- lapply(interest, function(level) {
  fitModel(sce, 
           target = "Treatment", 
           interest_level = level, reference_level = "control", 
           group = "Patient", strata = c("Disease"), 
           n_folds = 10, n_threads = 8)
})

plotAUC(result_list)


```
Only selected features

````{r}

# overview of CellProfiler feature categories
#   include: Texture, Intensity, AreaShape, 
#   maybe include: RadialDistribution,  Neighbors, Location
#   exclude because many corner cases: Granularity, Correlation

sce_select <- sce_list$stress[str_detect(names(sce_list$stress), "Texture|Intensity|AreaShape"), ]

interest <- setdiff(unique(sce_select$Treatment), c("control"))

result_list <- lapply(interest, function(level) {
  fitModel(sce_select,
                   target = "Treatment", 
                   interest_level = level, reference_level = "control", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 10, n_threads = 8)
  })

plotAUC(result_list)

ggsave("stress-batch correction.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)

````
#Filter one function at a time

```{r}
#TMRM <- mitochondrial health
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrTMRM"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Mitochondrial Health (TMRM)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

ggsave("AUC_TMRM-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)
````
#run each function with selected features

```{r}

#CorrCytoID <- autophagy
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce_list$stress[str_detect(names(sce_list$stress), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrCytoID"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


 

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Autophagy (CytoID)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

ggsave("AUC_CytoID-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)



#CorrLyso <- Lysosome
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrLyso"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Lysosome"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))


ggsave("AUC_Lysosome-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)

````
#explore each markers in details (only Intensity data)

````{r}
feature_names <- rownames(sce_list$stress)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce_list$stress, "tfmfeatures")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrTMRM", 
                                    "Intensity_MeanIntensity_CorrCytoID", 
                                    "Intensity_MeanIntensity_CorrLyso", 
                                    "Intensity_IntegratedIntensity_CorrTMRM",
                                    "Intensity_IntegratedIntensity_CorrCytoID",
                                    "Intensity_IntegratedIntensity_CorrLyso" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce_list$stress)[, c("Patient", "Treatment", "Disease", "Joint")])  # Replace with correct column names

# Combine metadata with intensity data
df <- cbind(metadata, intensity_data)

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

````
#plot TMRM

````{r}

df_filtered <- df_long %>%
  filter(Feature == "Intensity_MeanIntensity_CorrTMRM")  #Intensity_IntegratedIntensity_CorrTMRM , Intensity_MeanIntensity_CorrTMRM

#Summarize to get mean intensity per patient
df_per_patient <- df_filtered %>%
  group_by(Patient, Disease, Treatment) %>%
  summarise(mean_intensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


#Create summary data for bar plot (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(mean_intensity = mean(mean_intensity, na.rm = TRUE), .groups = "drop")

# 4. Plot
ggplot() +
  # Bars: mean across patients
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_intensity, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # One dot per patient
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = mean_intensity),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +
  labs(
    title = "Mean CorrTMRM Intensity per Patient",  #change to Mean if use
    x = "Disease",
    y = "Mean Intensity per Patient"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

````
#CytoID

````{r}
# Plot histogram for Intensity_MeanIntensity_CorrCytoID
# to select cut off

df_plot <- df_long %>%
  filter(
    Feature == "Intensity_MeanIntensity_CorrCytoID",
    !is.na(Intensity),
    is.finite(Intensity)  # remove Inf and NaN too
  )

ggplot(df_plot, aes(x = Intensity)) +
  geom_histogram(bins = 100, fill = "steelblue", color = "black") +
  coord_cartesian(
    xlim = c(min(df_plot$Intensity, na.rm = TRUE), 
             quantile(df_plot$Intensity, 0.99, na.rm = TRUE))
  ) +
  labs(
    title = "Histogram of Intensity_MeanIntensity_CorrCytoID",
    x = "Mean Intensity",
    y = "Cell Count"
  ) +
  theme_minimal()


# Calculate x between the 2 peaks of histogram
x <- df_corrCytoID$Intensity
x <- x[is.finite(x)]  # remove NAs/Infs

find_valley_between_main_peaks <- function(x, adjust = 1.5) {
  d <- density(x, adjust = adjust)
  y <- d$y; xx <- d$x
  
  # Find peaks
  sc <- diff(sign(diff(y)))
  peaks <- which(sc == -2) + 1
  peak_heights <- y[peaks]
  
  # Get two tallest peaks
  top2 <- order(peak_heights, decreasing = TRUE)[1:2]
  pidx <- sort(peaks[top2])  # ensure left then right
  
  # Find valleys between those peaks
  valleys <- which(sc == 2) + 1
  valley_candidates <- valleys[valleys > pidx[1] & valleys < pidx[2]]
  
  if (length(valley_candidates) == 0) return(NA_real_)
  
  # Pick the lowest valley
  v <- valley_candidates[which.min(y[valley_candidates])]
  return(xx[v])
}

cutoff <- find_valley_between_main_peaks(x, adjust = 1.5)
cat("Peak-minimum cutoff between main peaks:", cutoff, "\n")

# Plot
ggplot(df_corrCytoID, aes(x = Intensity)) +
  geom_histogram(binwidth = 2*IQR(x)/length(x)^(1/3), fill = "steelblue", color = "black") +
  geom_vline(xintercept = cutoff, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = cutoff, y = 0, label = sprintf("cutoff = %.4f", cutoff),
           angle = 90, vjust = -0.5, color = "red") +
  labs(title = "Histogram with cutoff between main peaks") +
  theme_minimal()



# Define the cutoff value => according from x calculation (x=0.046)
cutoff <- 0.01

df_corrCytoID <- df_long %>%
  filter(Feature == "Intensity_MeanIntensity_CorrCytoID", !is.na(Intensity))

# Create the histogram with adjusted axes
ggplot(df_corrCytoID, aes(x = Intensity)) +
  geom_histogram(
    binwidth = 0.02,
    fill = "blue",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = cutoff,  
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Histogram of Intensity_MeanIntensity_CorrCytoID",
    x = "Intensity",
    y = "Count"
  ) +
  facet_wrap(~ Disease + Treatment) +
  coord_cartesian(
    xlim = c(min(df_corrCytoID$Intensity, na.rm = TRUE),
             0.3)  # keep 0.3 as your right limit
  ) +
  scale_x_continuous(
    breaks = seq(floor(min(df_corrCytoID$Intensity, na.rm = TRUE)), 0.3, 0.03)
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

####
#Calculate % of cells above cutoff

df_summary <- df_corrCytoID %>%
  group_by(Disease, Treatment) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

# Print the result
print(df_summary)

#Plot — compare Treatments, facet by Disease

ggplot(df_summary, aes(x = Treatment, y = percent_above, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    y = "% Cells Above Cutoff",
    x = "Treatment"
  ) +
  facet_wrap(~ Disease) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

#facet by both Disease and Treatment
##Calculate per-patient percent above cutoff
df_per_patient <- df_corrCytoID %>%
  group_by(Disease, Treatment, Joint, Patient) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100,
    .groups = "drop"
  )

#Summarize for bars (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(
    mean_percent = mean(percent_above, na.rm = TRUE),
    .groups = "drop"
  )

#Plot with bar + dot (each patient) and facet grid
ggplot() +
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_percent, fill = Disease),
    width = 0.6,
    alpha = 0.5
  ) +
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = percent_above, color = Joint),
    width = 0.2,
    size = 3
  ) +
  geom_text_repel(
    data = df_per_patient,
    aes(x = Disease, y = percent_above, label = Patient, color = Joint),
    size = 3,
    box.padding = 0.4,
    point.padding = 0.3,
    max.overlaps = Inf,
    segment.color = "grey50"
  ) +
  facet_wrap(~ Treatment) +
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    x = "Disease",
    y = "% Cells Above Cutoff",
    color = "Joint"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14),
    legend.position = "bottom"
  )

````
#cytoID but plot directly from intensity

````{r}


df_filtered <- df_long %>%
  filter(Feature == "Intensity_MeanIntensity_CorrCytoID")  #Intensity_IntegratedIntensity_CorrCytoID ,         Intensity_MeanIntensity_CorrCytoID

#Summarize to get mean intensity per patient
df_per_patient <- df_filtered %>%
  group_by(Patient, Disease, Treatment) %>%
  summarise(mean_intensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


#Create summary data for bar plot (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(mean_intensity = mean(mean_intensity, na.rm = TRUE), .groups = "drop")

# 4. Plot
ggplot() +
  # Bars: mean across patients
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_intensity, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # One dot per patient
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = mean_intensity),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +
  labs(
    title = "Mean CorrCytoID Intensity per Patient",  #change to Mean if use
    x = "Disease",
    y = "Mean Intensity per Patient"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )
````
#Lysosome

````{r}
