GSEA analysis
Author: Mim & Ege
User and modify : Mim
18/07/2025

```{r}

#install.packages("HGNChelper")
library(dplyr)
library(tibble)
library(clusterProfiler)
library(enrichplot)
library(ggplot2)
library(forcats)
library(tidyverse)
library(DESeq2)
library(biomaRt)
library(HGNChelper) #they want citation
library(fgsea)
library(envalysis)
library(ggpubr)
library(org.Hs.eg.db) 
library(readxl)
library(eulerr)
library(grid)
library(ggvenn)
library(VennDiagram)
library(pheatmap)

```

###For enrichment analysis to autophagy, lipid metabolism etc. using deg from Tsuchiya et al. paper

Prepare the data for biological terms

```{r}
# for TNF stimulation
  setwd("/data/plapha/TNF DEG-hand_knee/analysis from Ege")
  TNF_data <- readRDS("tnf_vs_nontnf_non_filtered_patientwise.rds")
  TNF_data <- readRDS("nrm_counts_be_removed_mean.RDS")

# for cytokine stimulation Tsuchiya et al.
    
    setwd("/data/plapha/Cytokine stimulation gene Tsuchiya et al")

#Load raw file    
    cytokine_data <- readRDS("ra_de_list_only_expressed_genes.rds")
    names(cytokine_data)
    
# Load autophagy gene from https://autophagy.lu/genes/
   autophagy <-  read_csv("/data/plapha/Cytokine stimulation gene Tsuchiya et al/autophagy genes/HADB.csv")

   # for autophagy after filter fibroblast expressing gene
    autophagy  <- readRDS("/data/plapha/Cytokine stimulation gene Tsuchiya et al/autophagy genes/autophagy_gene_sets_filtered.rds")
   
   
# for healthy vs RA analysis
    setwd("~/data/Gene expression_healthy RA OA")

    df <- read_excel("/home/plapha/data/Gene expression_healthy RA OA/results_vs_healthy_vs_ra.xlsx")

    names(df)
    
# load IL1-exclusive gene 
    il1_genes <- read.csv("/home/plapha/data/Gene expression_healthy RA OA/IL1b_exclusive_genes.csv",
                      header = TRUE, stringsAsFactors = FALSE)
    # The gene list may be in first column → extract it
il1_gene_list <- il1_genes[[1]]

```
#Load new gene sets --> only express in Fibroblast (cutoff from expression file)

```{r}
Fib_gene_sets <- readRDS("fib_gene_sets_cut10.rds")
#Fib_gene_sets <- readRDS("fib_gene_sets_cut100.rds")

```
#read the deg from stimulations and cross-check if all the genes in our gene sets are represented in the deg file

```{r}

#for a specific cytokine:
IFNg_data <- cytokine_data$IFNg 
TNFa_data <- cytokine_data$TNFa
IL1b_data <- cytokine_data$IL1b
IL6_data <- cytokine_data$IL6
  
# check the coverage of deg genes to gene list we have
deg_genes <- unique(IFNg_data$genename)
coverage_check <- lapply(Fib_gene_sets, function(genes) {
  present <- genes %in% deg_genes
  list(
    present_genes = genes[present],
    missing_genes = genes[!present],
    n_present = sum(present),
    n_missing = sum(!present)
  )
})


#check the coverage
sapply(coverage_check, function(x) c(present = x$n_present, missing = x$n_missing))
#there is a huge discrepancy between degs and gene sets

#we replace the the names with the hgnc codes from deg and check the coverage again
IFNg_data$hgnc_name <- checkGeneSymbols(IFNg_data$genename)$Suggested.Symbol
deg_genes_updated <- unique(IFNg_data$hgnc_name)
coverage_check <- lapply(Fib_gene_sets, function(genes) {
  present <- genes %in% deg_genes_updated
  list(
    present_genes = genes[present],
    missing_genes = genes[!present],
    n_present = sum(present),
    n_missing = sum(!present)
  )
})
sapply(coverage_check, function(x) c(present = x$n_present, missing = x$n_missing))

#now it looks so much better!

```
#check coverage
##for healthy vs RA analysis

```{r}

deg_genes <- unique(df$gene)
coverage_check <- lapply(Fib_gene_sets, function(genes) {
  present <- genes %in% deg_genes
  list(
    present_genes = genes[present],
    missing_genes = genes[!present],
    n_present = sum(present),
    n_missing = sum(!present)
  )
})
#we replace the the names with the hgnc codes from deg and check the coverage again
df$hgnc_name <- checkGeneSymbols(df$gene)$Suggested.Symbol
deg_genes_updated <- unique(df$hgnc_name)
coverage_check <- lapply(Fib_gene_sets, function(genes) {
  present <- genes %in% deg_genes_updated
  list(
    present_genes = genes[present],
    missing_genes = genes[!present],
    n_present = sum(present),
    n_missing = sum(!present)
  )
})
sapply(coverage_check, function(x) c(present = x$n_present, missing = x$n_missing))


```

#prepare gsea input

```{r}
#we will filter the ones that doesnt have hgnc names (undefined transcripts) and the ones that dont have lfc 
valid_genes <- !is.na(IFNg_data$hgnc_name) & !is.na(IFNg_data$log2FoldChange)

IFNg_data_updated <- IFNg_data[valid_genes_x,]

unique_genes <- unique(IFNg_data_updated$hgnc_name) 
cat("IFNg dataframe after filtration number of genes:", nrow(IFNg_data_updated), 
    "\nNumber of unique hgnc symbols in the filtered dataframe:", length(unique_genes), "\n",
    "The difference between them:", nrow(IFNg_data_updated)- length(unique_genes))

#now we will check if these genes with more than one entry is represented in our gene set
duplicated_genes <- IFNg_data_updated$hgnc_name[duplicated(IFNg_data_updated$hgnc_name)]
duplicated_genes <- unique(duplicated_genes)

coverage_check <- lapply(Fib_gene_sets, function(genes) {
  present <- genes %in% duplicated_genes
  list(
    present_genes = genes[present],
    missing_genes = genes[!present],
    n_present = sum(present),
    n_missing = sum(!present)
  )
})
sapply(coverage_check, function(x) c(present = x$n_present, missing = x$n_missing))
#they are not covered in gene_set so we dont care about them

#for GSEA we can use use both approaches -> signed p values or shrunken LFC (I have already shrunken them before doing deseq2 deg analysis)

#the lfc rank
IFNg_ranks_shr <- setNames(IFNg_data_updated$log2FoldChange, IFNg_data_updated$hgnc_name)
IFNg_ranks_shr <- sort(IFNg_ranks_shr, decreasing = TRUE)
IFNg_ranks_shr <- IFNg_ranks_shr[!duplicated(names(IFNg_ranks_shr))]

#signed pval rank
IFNg_data_updated$sign_pval <- -log10(IFNg_data_updated$pvalue) * sign(IFNg_data_updated$log2FoldChange)
IFNg_ranks_sign_pval <- setNames(IFNg_data_updated$sign_pval, IFNg_data_updated$hgnc_name)
IFNg_ranks_sign_pval <- sort(IFNg_ranks_sign_pval, decreasing = TRUE)
IFNg_ranks_sign_pval <- IFNg_ranks_sign_pval[!duplicated(names(IFNg_ranks_sign_pval))]

#some of them are infinite converting them to a high number
high_val <- max(IFNg_ranks_sign_pval[is.finite(IFNg_ranks_sign_pval)], na.rm = TRUE) + 1
IFNg_ranks_sign_pval[is.infinite(IFNg_ranks_sign_pval) & IFNg_ranks_sign_pval > 0] <- high_val

```

#prepare gsea input
##for healthy vs RA analysis

```{r}

#we will filter the ones that doesnt have hgnc names (undefined transcripts) and the ones that dont have lfc 

df_updated <- df %>%
  filter(!is.na(hgnc_name), !is.na(log2FoldChange))

unique_genes <- unique(df_updated$hgnc_name) 
cat("dataframe after filtration number of genes:", nrow(df_updated), 
    "\nNumber of unique hgnc symbols in the filtered dataframe:", length(unique_genes), "\n",
    "The difference between them:", nrow(df_updated)- length(unique_genes))

#now we will check if these genes with more than one entry is represented in our gene set
duplicated_genes <- df_updated$hgnc_name[duplicated(df_updated$hgnc_name)]
duplicated_genes <- unique(duplicated_genes)

coverage_check <- lapply(Fib_gene_sets, function(genes) {
  present <- genes %in% duplicated_genes
  list(
    present_genes = genes[present],
    missing_genes = genes[!present],
    n_present = sum(present),
    n_missing = sum(!present)
  )
})
sapply(coverage_check, function(x) c(present = x$n_present, missing = x$n_missing))
#they are not covered in gene_set so we dont care about them

#for GSEA we can use use both approaches -> signed p values or shrunken LFC (I have already shrunken them before doing deseq2 deg analysis)

#the lfc rank
ranks_shr <- setNames(df_updated$log2FoldChange, df_updated$hgnc_name)
ranks_shr <- sort(ranks_shr, decreasing = TRUE)
ranks_shr <- ranks_shr[!duplicated(names(ranks_shr))]

#signed pval rank
df_updated$sign_pval <- -log10(df_updated$pvalue) * sign(df_updated$log2FoldChange)
ranks_sign_pval <- setNames(df_updated$sign_pval, df_updated$hgnc_name)
ranks_sign_pval <- sort(ranks_sign_pval, decreasing = TRUE)
ranks_sign_pval <- ranks_sign_pval[!duplicated(names(ranks_sign_pval))]

#some of them are infinite converting them to a high number
high_val <- max(ranks_sign_pval[is.finite(ranks_sign_pval)], na.rm = TRUE) + 1
ranks_sign_pval[is.infinite(ranks_sign_pval) & ranks_sign_pval > 0] <- high_val

```

run gsea with both ranking systems

```{r}

fgsea_ifng_lfc <- fgsea(
  pathways = Fib_gene_sets,
  stats    = IFNg_ranks_shr,
)

fgsea_ifng_pval <- fgsea(
  pathways = Fib_gene_sets,
  stats    = IFNg_ranks_sign_pval,
)

plot_fgsea <- function(fgsea_result, name_of_the_stimulation,if_lfc) {
  if (if_lfc == TRUE){
    title = paste0("GSEA Pathways of ",name_of_the_stimulation," (log2foldchange)") 
  }else{
    title = paste0("GSEA Pathways of ",name_of_the_stimulation," (signed pvalue)") 
  }
  
  fgsea_result <- fgsea_result %>%
    mutate(
      padj_label = formatC(padj, format = "f", digits = 3),
      hjust_value = ifelse(NES > 0, 1.1, -0.1)  # adjust position depending on NES direction
    )
  
  # Create the plot
  p <- ggplot(fgsea_result, aes(x = reorder(pathway, NES), y = NES, fill = padj)) +
    geom_bar(stat = "identity", alpha = 0.9, width = 0.7) +
    geom_text(aes(label = padj_label, hjust = hjust_value), size = 4, color = "white") +
    scale_fill_gradient(low = "#4575b4", high = "#d73027", limits = c(0, max(fgsea_result$padj))) +
    coord_flip() +
    labs(
      title = title,
      x = "Pathway",
      y = "Normalized Enrichment Score (NES)",
      fill = "padj"
    ) +
    theme_publish(base_size = 14) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      legend.text = element_text(angle = 45)
    )
  
  return(p)
}

plot_fgsea_lfc <- plot_fgsea(fgsea_ifng_lfc,"IFNg",TRUE)
plot_fgsea_lfc

plot_fgsea_pval <- plot_fgsea(fgsea_ifng_pval,"IFNg",FALSE)
plot_fgsea_pval


ggsave(
  plot = plot_fgsea_pval,
  filename = "FGSEA_ifng_plot_pval.png",
  bg = "transparent",
  width = 9, height = 6
)

```

run gsea with both ranking systems
##for healthy vs RA analysis

```{r}

fgsea_ifng_lfc <- fgsea(
  pathways = Fib_gene_sets,
  stats    = ranks_shr,
)

fgsea_ifng_pval <- fgsea(
  pathways = Fib_gene_sets,
  stats    = ranks_sign_pval,
)

plot_fgsea <- function(fgsea_result, name_of_the_stimulation, if_lfc) {
  
  # Create label column for display
  fgsea_result$label <- ifelse(
    fgsea_result$padj < 0.001, 
    "< 0.01", 
    round(fgsea_result$padj, 2)
  )
  
  # Set title
  if (if_lfc) {
    title <- paste0("GSEA Pathways of ", name_of_the_stimulation, " (log2foldchange)") 
  } else {
    title <- paste0("GSEA Pathways of ", name_of_the_stimulation, " (signed pvalue)") 
  }
  
  # Plot
  p <- ggplot(fgsea_result, aes(x = reorder(pathway, -NES), y = NES, fill = padj)) +
    geom_bar(stat = "identity", alpha = 0.9, width = 0.7) +
    geom_text(
      aes(
        label = label,
        hjust = ifelse(NES >= 0, 1.1, -0.1)
      ),
      size = 5,
      color = "white"
    ) +
    scale_fill_gradient(
      low = "#4575b4", 
      high = "#d73027", 
      limits = c(0, max(fgsea_result$padj))
    ) +
    coord_flip() +
    labs(
      title = title,
      x = "Pathway",
      y = "Normalized Enrichment Score (NES)",
      fill = "padj"
    ) +
    theme_publish(base_size = 16) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      legend.text = element_text(angle = 45)
    )
  
  return(p)
}

plot_fgsea_lfc <- plot_fgsea(fgsea_ifng_lfc, "Healthy vs RA", TRUE)
plot_fgsea_lfc

plot_fgsea_pval <- plot_fgsea(fgsea_ifng_pval, "Healthy vs RA", FALSE)
plot_fgsea_pval


ggsave(
  plot = plot_fgsea_pval,
  filename = "FGSEA_ifng_plot_pval.png",
  bg = "transparent",
  width = 9, height = 6
)

```

# Extract gene list + padj from GSEA
##for healthy vs RA analysis

```{r}

#extract from log2foldchange plot

gene_list_autophagy_lipid <- fgsea_ifng_lfc %>%
  as_tibble() %>%
  filter(pathway %in% c("Autophagy", "Lipid Metabolism")) %>%
  dplyr::select(pathway, padj, pval, leadingEdge) %>%
  unnest(cols = leadingEdge) %>%
  rename(gene = leadingEdge)

# Merge with your original DE results
gene_list_with_lfc <- gene_list_autophagy_lipid %>%
  left_join(df %>% dplyr::select(hgnc_name, log2FoldChange, pvalue),
            by = c("gene" = "hgnc_name"))

gene_list_with_lfc

write.csv(gene_list_with_lfc, "gene_list_with_lfc.csv", row.names = FALSE)

```
#Loop for all cytokines

```{r}
run_gsea_for_cytokine <- function(cytokine_name, cytokine_data, Fib_gene_sets, output_dir = "fgsea_outputs_new") {
  
  message("Processing cytokine: ", cytokine_name)
  
  data <- cytokine_data[[cytokine_name]]
  
  # Convert to HGNC symbols
  data$hgnc_name <- checkGeneSymbols(data$genename)$Suggested.Symbol
  
  # Filter out invalid entries
  valid_genes <- !is.na(data$hgnc_name) & !is.na(data$log2FoldChange)
  data <- data[valid_genes, ]

  # Remove duplicates
  data <- data[!duplicated(data$hgnc_name), ]

  # Create ranking vectors
  lfc_ranks <- setNames(data$log2FoldChange, data$hgnc_name)
  lfc_ranks <- sort(lfc_ranks, decreasing = TRUE)

  data$sign_pval <- -log10(data$pvalue) * sign(data$log2FoldChange)
  sign_pval_ranks <- setNames(data$sign_pval, data$hgnc_name)
  sign_pval_ranks <- sort(sign_pval_ranks, decreasing = TRUE)

  # Replace Inf values with large finite numbers
  high_val <- max(sign_pval_ranks[is.finite(sign_pval_ranks)], na.rm = TRUE) + 1
  sign_pval_ranks[is.infinite(sign_pval_ranks) & sign_pval_ranks > 0] <- high_val
  sign_pval_ranks <- sign_pval_ranks[!is.na(sign_pval_ranks)]

  # Run fgsea
  fgsea_lfc <- fgsea(pathways = Fib_gene_sets, stats = lfc_ranks)
  fgsea_pval <- fgsea(pathways = Fib_gene_sets, stats = sign_pval_ranks)

  # Plotting function
  plot_fgsea <- function(fgsea_result, title) {
    fgsea_result$hjust <- ifelse(fgsea_result$NES >= 0, 1.1, -0.1)
    fgsea_result$label <- ifelse(fgsea_result$padj < 0.001, "< 0.01", round(fgsea_result$padj, 2))

    ggplot(fgsea_result, aes(x = reorder(pathway, -NES), y = NES, fill = padj)) +
      geom_bar(stat = "identity", alpha = 0.9, width = 0.7) +
      geom_text(aes(label = label, hjust = hjust), size = 5, color = "white") +
      scale_fill_gradient(low = "#4575b4", high = "#d73027",
                          limits = c(0, max(fgsea_result$padj, na.rm = TRUE))) +
      coord_flip() +
      labs(title = title, x = "Pathway", y = "Normalized Enrichment Score (NES)", fill = "padj") +
      theme_publish(base_size = 14) +
      theme(plot.title = element_text(size = 16, face = "bold"),
            legend.position = "bottom", legend.text = element_text(angle = 45))
  }
  
  # Create plots
  p_lfc <- plot_fgsea(fgsea_lfc, paste("GSEA -", cytokine_name, "(LFC)"))
  p_pval <- plot_fgsea(fgsea_pval, paste("GSEA -", cytokine_name, "(Signed p-value)"))

  # Save results
  dir.create(output_dir, showWarnings = FALSE)

  ggsave(filename = file.path(output_dir, paste0("FGSEA_", cytokine_name, "_lfc.png")),
         plot = p_lfc, width = 9, height = 6, bg = "transparent")
  
  ggsave(filename = file.path(output_dir, paste0("FGSEA_", cytokine_name, "_pval.png")),
         plot = p_pval, width = 9, height = 6, bg = "transparent")
  
  # Return results if needed later
  list(
    fgsea_lfc = fgsea_lfc,
    fgsea_pval = fgsea_pval,
    plot_lfc = p_lfc,
    plot_pval = p_pval,
    ranking_vector = lfc_ranks
  )
}
```
#Run for all cytokines

```{r}
cytokine_names <- names(cytokine_data)  # should include: "all", "IFNa", "IFNg", etc.

gsea_results_all <- lapply(cytokine_names, function(cytokine) {
  run_gsea_for_cytokine(cytokine, cytokine_data, Fib_gene_sets)
})

names(gsea_results_all) <- cytokine_names
#to look at a specific inflammation
gsea_results_all$IL6

```
#classic GSEA plot for each pathway

```{r}
#recreating lfc_ranks_IFNg

# Re-extract IFNg data
data_IFNg <- cytokine_data[["IFNg"]]

# Map gene names to HGNC
data_IFNg$hgnc_name <- checkGeneSymbols(data_IFNg$genename)$Suggested.Symbol

# Filter out invalid/NA entries
valid_genes <- !is.na(data_IFNg$hgnc_name) & !is.na(data_IFNg$log2FoldChange)
data_IFNg <- data_IFNg[valid_genes, ]

# Remove duplicates
data_IFNg <- data_IFNg[!duplicated(data_IFNg$hgnc_name), ]

# Create ranked vector
lfc_ranks_IFNg <- setNames(data_IFNg$log2FoldChange, data_IFNg$hgnc_name)
lfc_ranks_IFNg <- sort(lfc_ranks_IFNg, decreasing = TRUE)

##rerun your plotEnrichment() loop

top_pathways <- gsea_results_all$IFNg$fgsea_lfc[order(gsea_results_all$IFNg$fgsea_lfc$padj), ]

for (i in seq_len(min(5, nrow(top_pathways)))) {
  pathway_name <- top_pathways$pathway[i]
  
  enrichment_plot <- plotEnrichment(
    pathway = Fib_gene_sets[[pathway_name]],
    stats = lfc_ranks_IFNg
  ) +
    ggtitle(paste("Enrichment for:", pathway_name))
  
  print(enrichment_plot)
}

```
#analyse overlapping gene from cytokine_data to functions in gene_sets --> Venn diagram

```{r}
#for a specific cytokine:
IFNg_data <- cytokine_data$IFNg 
TNFa_data <- cytokine_data$TNFa
IL1b_data <- cytokine_data$IL1b

# --- Select gene set (just change this line)
pathway_name <- "Neutral Lipid"   # or "Proliferation",Autophagy,Stress,Lipid Metabolism
gene_set <- Fib_gene_sets[[pathway_name]]

# --- Filter each cytokine to the chosen gene set
filter_genes <- function(cytokine_data, gene_set) {
  tmp <- cytokine_data[cytokine_data$genename %in% gene_set & 
                         cytokine_data$log2FoldChange > 0 & 
                         cytokine_data$padj < 0.05, ]
  return(tmp$genename)
}

IFNg_genes <- filter_genes(IFNg_data, gene_set)
TNFa_genes <- filter_genes(TNFa_data, gene_set)
IL1b_genes <- filter_genes(IL1b_data, gene_set)

# --- Create Venn diagram
fit <- euler(list(
  IFNg = IFNg_genes,
  TNFa = TNFa_genes,
  IL1b = IL1b_genes
))

plot(fit,
     fills = list(fill = c("skyblue", "pink", "lightgreen"), alpha = 0.5),
     labels = list(font = 4),
     main = paste("Overlap of upregulated & significant IFNg, TNFa, IL1b genes in", pathway_name, "set"),
     quantities = TRUE)


# --- Find overlap sets
genes_IFNg_only      <- setdiff(IFNg_genes, union(TNFa_genes, IL1b_genes))
genes_TNFa_only      <- setdiff(TNFa_genes, union(IFNg_genes, IL1b_genes))
genes_IL1b_only      <- setdiff(IL1b_genes, union(IFNg_genes, TNFa_genes))

genes_IFNg_TNFa      <- intersect(IFNg_genes, TNFa_genes)
genes_IFNg_IL1b      <- intersect(IFNg_genes, IL1b_genes)
genes_TNFa_IL1b      <- intersect(TNFa_genes, IL1b_genes)
genes_all_three      <- Reduce(intersect, list(IFNg_genes, TNFa_genes, IL1b_genes))

# --- Combine into a wide table
max_len <- max(length(genes_IFNg_only), length(genes_TNFa_only), length(genes_IL1b_only),
               length(genes_IFNg_TNFa), length(genes_IFNg_IL1b), length(genes_TNFa_IL1b), 
               length(genes_all_three))

pad <- function(x) c(x, rep(NA, max_len - length(x)))

venn_table_3 <- data.frame(
  IFNg_only       = pad(genes_IFNg_only),
  TNFa_only       = pad(genes_TNFa_only),
  IL1b_only       = pad(genes_IL1b_only),
  IFNg_TNFa       = pad(genes_IFNg_TNFa),
  IFNg_IL1b       = pad(genes_IFNg_IL1b),
  TNFa_IL1b       = pad(genes_TNFa_IL1b),
  All_three       = pad(genes_all_three),
  stringsAsFactors = FALSE
)

# View table
head(venn_table_3)

# --- Export to CSV (name automatically adjusted by set)
outfile <- paste0("IFNg_TNFa_IL1b_Venn_", pathway_name, ".csv")
write.csv(venn_table_3, outfile, row.names = FALSE)


```
#Cytokine up reg gene VS all Fib gene in each pathways -> Venn

```{r}

# extract cytokine data like above

#----1.Select TNFa upregulated genes
TNFa_up <- TNFa_data %>%
     filter(log2FoldChange > 0, padj < 0.05) %>%
     pull(genename)

# ---- 2. Select fibroblast proliferation genes ----
Fib_prolif <- Fib_gene_sets$Proliferation

# ---- 3. Create Venn input ----
venn_list <- list(
  TNFa_upregulated = TNFa_up,
  Fibroblast_Proliferation = Fib_prolif
)

# ---- 4. Plot Venn diagram ----
venn.plot <- venn.diagram(
  x = list(
    TNFa_upregulated = TNFa_up,
    Fibroblast_Proliferation = Fib_prolif
  ),
  filename = NULL,
  fill = c("lightpink1", "#56B4E9"),
  alpha = 0.5,
  cex = 1.5,               # size of numbers inside circles
  cat.cex = 1.5,           # size of category labels
  scaled = TRUE,
  cat.fontfamily = "Helvetica", # font for category labels
  cat.fontface = "bold",    # style for category labels
  fontfamily = "Helvetica",     # font for numbers inside circles
  fontface = "plain",
  cat.pos = c(0, 180),    # angle position of the labels
  cat.dist = c(0.05, 0.05) # distance from the circles; increase if labels are long
# style for numbers inside circles
)

grid::grid.newpage()      
grid::grid.draw(venn.plot)

```
# Autophagy analysis from HADB.csv file -> volcano plot

```{r}

### --- Choose cytokine here ---
selected_cytokine <- "IL1b"   # <-- change to "IL1b" or TNFa IFNg

### --- Extract selected dataset ---
df <- cytokine_data[[selected_cytokine]]

### --- Make sure gene column names match ---
# autophagy_genes$GENE  should match df$genename
df$genename <- toupper(df$genename)
autophagy$GENE <- toupper(autophagy$GENE)

### --- Identify autophagy genes present in cytokine-stimulated dataset ---
autophagy_hits <- df[df$genename %in% autophagy$GENE, ]

### --- Optional: only significantly changed genes ---
sig_cutoff <- 0.05
autophagy_hits_sig <- autophagy_hits[autophagy_hits$padj < sig_cutoff, ]

### --- Enrichment test (Fisher exact test) ---
background_genes <- df$genename
auto_set <- autophagy$GENE

# Build contingency table
in_auto <- background_genes %in% auto_set
is_sig <- df$padj < sig_cutoff

tbl <- table(Autophagy = in_auto, Significant = is_sig)

fisher_result <- fisher.test(tbl)

### --- Print results ---
cat("Cytokine selected:", selected_cytokine, "\n\n")

cat("Number of autophagy genes in dataset =", nrow(autophagy_hits), "\n")
cat("Significant autophagy genes =", nrow(autophagy_hits_sig), "\n\n")

cat("Fisher enrichment p-value =", fisher_result$p.value, "\n")
cat("Odds ratio =", fisher_result$estimate, "\n\n")

### --- View genes ---
autophagy_hits_sig[order(autophagy_hits_sig$log2FoldChange, decreasing = TRUE), ]

#-------------------------
# volcano plot
library(ggrepel)

# Prepare Autophagy column
df$Autophagy <- ifelse(df$genename %in% auto_set, "Autophagy", "Other")

# Filter significant autophagy genes to label
label_genes <- df %>%
    filter(Autophagy == "Autophagy", padj < 0.05)

# Separate datasets for plotting order
df_other <- df %>% filter(Autophagy == "Other")
df_auto  <- df %>% filter(Autophagy == "Autophagy")

# -----------------------
# Fisher enrichment test (same as before)
background_genes <- df$genename
sig_cutoff <- 0.05
in_auto <- df$genename %in% auto_set
is_sig  <- df$padj < sig_cutoff
tbl <- table(Autophagy = in_auto, Significant = is_sig)
fisher_result <- fisher.test(tbl)
fisher_label <- paste0("Fisher enrichment p = ", signif(fisher_result$p.value,3),
                       ", OR = ", round(fisher_result$estimate,2))

# -----------------------
# Volcano plot
ggplot() +
  # Other genes (gray)
  geom_point(data = df_other,
             aes(x = log2FoldChange, y = -log10(padj), color = Autophagy, shape = Autophagy, size = Autophagy),
             alpha = 0.6, show.legend = TRUE) +
  # Autophagy genes (red, on top)
  geom_point(data = df_auto,
             aes(x = log2FoldChange, y = -log10(padj), color = Autophagy, shape = Autophagy, size = Autophagy),
             alpha = 0.9, stroke = 0.8, show.legend = TRUE) +
  # Scale manual for color, shape, size
  scale_color_manual(values = c("Autophagy" = "red", "Other" = "gray60")) +
  scale_shape_manual(values = c("Autophagy" = 16, "Other" = 16)) +
  scale_size_manual(values = c("Autophagy" = 3.5, "Other" = 2)) +
  # Lines
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  # Labels for significant autophagy genes
  geom_text_repel(
    data = label_genes,
    aes(x = log2FoldChange, y = -log10(padj), label = genename),
    size = 4,
    max.overlaps = 50,
    box.padding = 0.3,
    point.padding = 0.2
  ) +
  theme_bw(base_size = 16) +
  labs(
    title = paste0("Volcano plot — ", selected_cytokine),
    subtitle = fisher_label,
    x = "log2 Fold Change",
    y = "-log10(adj p-value)",
    color = "Gene Type",
    shape = "Gene Type",
    size = "Gene Type"
  )


```
# extract data from volcano -> Venn diagram

```{r}
### FUNCTION TO RUN ANALYSIS
### -------------------------------
analyze_cytokine <- function(cytokine_name, cytokine_data, autophagy, sig_cutoff = 0.05) {

  # Extract dataset
  df <- cytokine_data[[cytokine_name]]

  # Harmonize gene names
  df$genename <- toupper(df$genename)
  autophagy$GENE <- toupper(autophagy$GENE)

  # Tag autophagy genes
  df$Autophagy <- ifelse(df$genename %in% autophagy$GENE, "Autophagy", "Other")

  # Fisher test
  background_genes <- df$genename
  in_auto <- df$genename %in% autophagy$GENE
  is_sig <- df$padj < sig_cutoff

  tbl <- table(Autophagy = in_auto, Significant = is_sig)
  fisher_result <- fisher.test(tbl)

  # Genes enriched (autophagy + significant)
  enriched_genes <- df %>%
    filter(Autophagy == "Autophagy", padj < sig_cutoff) %>%
    pull(genename)

  list(
    df = df,
    enriched_genes = enriched_genes,
    fisher = fisher_result
  )
}

### -------------------------------
### RUN FOR BOTH CYTOKINES
### -------------------------------
res_IL1b <- analyze_cytokine("IL1b", cytokine_data, autophagy)
res_TNFa <- analyze_cytokine("TNFa", cytokine_data, autophagy)

# Extract ONLY genes enriched by Fisher analysis + significant
sig_IL1b  <- res_IL1b$enriched_genes
sig_TNFa  <- res_TNFa$enriched_genes

### CHECK
cat("IL1b enriched genes:", length(sig_IL1b), "\n")
cat("TNFa enriched genes:", length(sig_TNFa), "\n")

### -------------------------------
### FIX VENN — remove NAs
### -------------------------------
sig_IL1b <- sig_IL1b[!is.na(sig_IL1b)]
sig_TNFa <- sig_TNFa[!is.na(sig_TNFa)]

### -------------------------------
### DRAW VENN
### -------------------------------
venn.plot <- venn.diagram(
  x = list(
    IL1b = sig_IL1b,
    TNFa = sig_TNFa
  ),
  filename = NULL,
  fill = c("#ff9999", "#9999ff"),
  alpha = 0.5,
  cex = 2,
  cat.cex = 2,
  cat.pos = c(-20, 20)
)

grid::grid.newpage()
grid::grid.draw(venn.plot)

### -------------------------------
### GET GENES EXCLUSIVE TO IL1b
### -------------------------------
exclusive_IL1b <- setdiff(sig_IL1b, sig_TNFa)
exclusive_IL1b

write.csv(
  data.frame(Gene = exclusive_IL1b),
  "exclusive_IL1b_genes.csv",
  row.names = FALSE
)

cat("\nSaved: exclusive_IL1b_genes.csv\n")

#--------
### ============================================================
### Function to extract enriched autophagy genes per cytokine
### ============================================================

get_autophagy_hits <- function(df, autophagy, sig_cutoff = 0.05) {
  
  # harmonize case
  df$genename <- toupper(df$genename)
  autophagy$GENE <- toupper(autophagy$GENE)
  
  # autophagy present in dataset
  hits <- df[df$genename %in% autophagy$GENE, ]
  
  # significant autophagy genes
  hits_sig <- hits[hits$padj < sig_cutoff, ]
  
  return(hits_sig$genename)
}

### ============================================================
### --- RUN FOR IL1b ---
### ============================================================

df_IL1b <- cytokine_data[["IL1b"]]
sig_IL1b <- get_autophagy_hits(df_IL1b, autophagy)

### ============================================================
### --- RUN FOR TNFa ---
### ============================================================

df_TNFa <- cytokine_data[["TNFa"]]
sig_TNFa <- get_autophagy_hits(df_TNFa, autophagy)

### ============================================================
### Compute exclusive + intersect
### ============================================================

intersect_genes   <- intersect(sig_IL1b, sig_TNFa)
exclusive_IL1b    <- setdiff(sig_IL1b, sig_TNFa)
exclusive_TNFa    <- setdiff(sig_TNFa, sig_IL1b)

### ============================================================
### Save to CSV
### ============================================================

write.csv(data.frame(gene = exclusive_IL1b),
          file = "exclusive_IL1b.csv",
          row.names = FALSE)

write.csv(data.frame(gene = exclusive_TNFa),
          file = "exclusive_TNFa.csv",
          row.names = FALSE)

write.csv(data.frame(gene = intersect_genes),
          file = "intersect_IL1b_TNFa.csv",
          row.names = FALSE)

### Print summary
cat("IL1b significant autophagy genes:", length(sig_IL1b), "\n")
cat("TNFa significant autophagy genes:", length(sig_TNFa), "\n\n")

cat("Exclusive IL1b:", length(exclusive_IL1b), "genes\n")
cat("Exclusive TNFa:", length(exclusive_TNFa), "genes\n")
cat("Intersect:", length(intersect_genes), "genes\n")

#------add log2foldchange, adjp----
# IL1b exclusive genes table
IL1b_exclusive_table <- df_IL1b %>%
  filter(genename %in% exclusive_IL1b) %>%
   dplyr::select(genename, log2FoldChange, padj)

# TNFa exclusive genes table
TNFa_exclusive_table <- df_TNFa %>%
  filter(genename %in% exclusive_TNFa) %>%
  dplyr::select(genename, log2FoldChange, padj)

# Intersect genes table
intersect_table <- df_IL1b %>%
  filter(genename %in% intersect_genes) %>%
   dplyr::select(genename, log2FoldChange, padj)  # can also take from TNFa, should be identical

# Save to CSV
write.csv(IL1b_exclusive_table, "IL1b_exclusive_genes.csv", row.names = FALSE)
write.csv(TNFa_exclusive_table, "TNFa_exclusive_genes.csv", row.names = FALSE)
write.csv(intersect_table, "intersect_IL1b_TNFa_genes.csv", row.names = FALSE)

```
#-----visualise-----

```{r}
ggplot(IL1b_exclusive_table, aes(x = reorder(genename, log2FoldChange), y = log2FoldChange)) +
  geom_bar(stat = "identity", fill = "salmon") +
  coord_flip() +
  geom_text(aes(label = round(log2FoldChange, 2)), hjust = -0.1) +
  theme_bw(base_size = 14) +
  labs(x = "Gene", y = "log2 Fold Change",
       title = "IL1b-exclusive autophagy genes") 

#OR
ggplot(IL1b_exclusive_table, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(color = "salmon", size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted") +
  geom_text_repel(aes(label = genename), max.overlaps = 20) +
  theme_bw(base_size = 14) +
  labs(title = "IL1b-exclusive autophagy genes volcano",
       x = "log2 Fold Change", y = "-log10(adj p-value)")

# or
library(pheatmap)
# Extract IL1b data for exclusive genes
IL1b_data_for_heatmap <- df_IL1b %>%
  filter(genename %in% exclusive_IL1b) %>%
  dplyr::select(genename, log2FoldChange) %>%
  mutate(condition = "IL1b")

# Extract TNFa data for exclusive genes
TNFa_data_for_heatmap <- df_TNFa %>%
  filter(genename %in% exclusive_TNFa) %>%
  dplyr::select(genename, log2FoldChange) %>%
  mutate(condition = "TNFa")

# Extract IL1b data for intersect genes (use df_IL1b or df_TNFa as preferred)
intersect_data_for_heatmap <- df_IL1b %>%
  filter(genename %in% intersect_genes) %>%
  dplyr::select(genename, log2FoldChange) %>%
  mutate(condition = "Intersect")

# Combine all datasets
all_heatmap_data <- bind_rows(IL1b_data_for_heatmap,
                              TNFa_data_for_heatmap,
                              intersect_data_for_heatmap)

# Pivot to wide format (genes as rows, conditions as columns)
heatmap_wide <- all_heatmap_data %>%
  pivot_wider(names_from = condition,
              values_from = log2FoldChange)

# Prepare matrix for heatmap
heatmap_mat <- as.data.frame(heatmap_wide)
rownames(heatmap_mat) <- heatmap_mat$genename
heatmap_mat$genename <- NULL
heatmap_mat <- as.matrix(heatmap_mat)

# Replace NA values with 0 or other appropriate value
heatmap_mat[is.na(heatmap_mat)] <- 0

# Optional scaling per gene (row)
heatmap_scaled <- t(scale(t(heatmap_mat)))

# Plot heatmap
pheatmap(heatmap_scaled,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         show_rownames = TRUE,
         show_colnames = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         main = "log2 Fold Change Heatmap of IL1b, TNFa Exclusive and Intersect Genes")

#or
IL1b_exclusive_table <- IL1b_exclusive_table %>%
  mutate(Regulation = ifelse(log2FoldChange > 0, "Up", "Down"))

ggplot(IL1b_exclusive_table, aes(x = reorder(genename, log2FoldChange), y = log2FoldChange, color = Regulation)) +
  geom_segment(aes(x = genename, xend = genename, y = 0, yend = log2FoldChange), size = 0.8) +
  geom_point(size = 4) +
  coord_flip() +
  scale_color_manual(values = c("Up" = "salmon", "Down" = "lightblue")) +
  theme_bw(base_size = 14) +
  labs(x = "Gene", y = "log2 Fold Change",
       title = "IL1b-exclusive autophagy genes",
       color = "Regulation")
```
# run GSEA from Autophagy analysis (HADB.csv file)

```{r}

# ------------------------
# Function to run GSEA on any cytokine dataset
run_gsea <- function(cytokine_df, gene_set, cytokine_name) {

  # --- Filter valid genes (remove NA genename or log2FoldChange) ---
  valid_genes <- !is.na(cytokine_df$genename) & !is.na(cytokine_df$log2FoldChange)
  df <- cytokine_df[valid_genes, ]
  
  cat("Cytokine dataset:", cytokine_name, "\n",
      "Number of genes after filtration:", nrow(df), "\n",
      "Unique genes:", length(unique(df$genename)), "\n\n")
  
  # --- Remove duplicated genes (keep first occurrence) ---
  df <- df[!duplicated(df$genename), ]
  
  # --- Create ranking vectors ---
  # log2FoldChange ranking
  ranks_lfc <- setNames(df$log2FoldChange, df$genename)
  ranks_lfc <- sort(ranks_lfc, decreasing = TRUE)
  
  # signed p-value ranking
  df$sign_pval <- -log10(df$pvalue) * sign(df$log2FoldChange)
  ranks_sign_pval <- setNames(df$sign_pval, df$genename)
  ranks_sign_pval <- sort(ranks_sign_pval, decreasing = TRUE)
  
  # Handle infinite values
  high_val <- max(ranks_sign_pval[is.finite(ranks_sign_pval)], na.rm = TRUE) + 1
  ranks_sign_pval[is.infinite(ranks_sign_pval) & ranks_sign_pval > 0] <- high_val
  
  # --- Run fgsea ---
  fgsea_lfc <- fgsea(pathways = gene_set, stats = ranks_lfc)
  fgsea_pval <- fgsea(pathways = gene_set, stats = ranks_sign_pval)
  
  return(list(lfc = fgsea_lfc, sign_pval = fgsea_pval))
}

# ------------------------
# Function to plot multiple cytokines on the same y-axis
plot_gsea_single_axis_labels <- function(gsea_combined, title = "GSEA NES Plot") {
  
  gsea_combined <- gsea_combined %>%
    mutate(
      padj_label = formatC(padj, format = "f", digits = 3),
      pathway = factor(pathway, levels = rev(unique(pathway)))
    )
  
  # Cytokine colors
  cytokine_colors <- setNames(
    RColorBrewer::brewer.pal(n = length(unique(gsea_combined$Cytokine)), name = "Set2"),
    unique(gsea_combined$Cytokine)
  )
  
  p <- ggplot(gsea_combined, aes(y = pathway, x = NES, fill = Cytokine)) +
    geom_bar(stat = "identity", position = position_dodge(width = 0.7), alpha = 0.9, width = 0.6) +
    geom_text(aes(label = padj_label), position = position_dodge(width = 0.7),
              hjust = -0.1, size = 5, color = "black") +
    geom_text(aes(x = min(NES) - 0.2*max(NES), label = Cytokine),
              position = position_dodge(width = 0.7),
              hjust = 1, size = 5, fontface = "bold", color = "black") +
    scale_fill_manual(values = cytokine_colors) +
    labs(title = title, x = "Normalized Enrichment Score (NES)", y = "Pathway", fill = "Cytokine") +
    theme_bw(base_size = 14) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      axis.text.y = element_text(size = 12),
      axis.text.x = element_text(size = 12),
      panel.grid.major.y = element_blank(),
      panel.grid.minor.y = element_blank()
    )
  
  return(p)
}

# ------------------------
# Example usage
# Assuming gsea_all was generated with run_gsea_multiple()
# Step 1: select which cytokines you want to include
selected_cytokines <- cytokine_data[c("TNFa", "IL1b")]   # <-- define here

# Step 2: run GSEA for those selected cytokines
gsea_all <- run_gsea_multiple(selected_cytokines, list(Autophagy = toupper(autophagy_genes)))

plot_single_axis_labels <- plot_gsea_single_axis_labels(gsea_all, title = "Autophagy GSEA — TNFa & IL1b")
plot_single_axis_labels 

# Save plot

ggsave("GSEA_Autophagy_Multiple_Cytokines.png", plot_single_axis_labels , width = 10, height = 6, dpi = 300)

#-------
# with the filtered list of autophagy
# --- Example usage ---
# Filter autophagy genes from your autophagy list
autophagy_gene_set <- list(Autophagy = toupper(autophagy$Autophagy))

# Select cytokines to analyze
selected_cytokines <- cytokine_data[c("TNFa", "IL1b")]

# Run GSEA
gsea_all <- run_gsea_multiple(selected_cytokines, autophagy_gene_set)

# Plot NES for all cytokines on same y-axis
p <- plot_gsea_single_axis_labels(gsea_all$lfc, title = "Autophagy GSEA — TNFa & IL1b")
p

```
# heat map IL1 exclusive in basal cells (healthy, RA)

```{r}
 setwd("~/data/Gene expression_healthy RA OA")

    df <- read.csv("nrm_counts_copy_with_be_removed.csv")
    names(df)
    
# load IL1-exclusive gene 
    il1_genes <- read.csv("IL1b_exclusive_genes.csv",
                      header = TRUE, stringsAsFactors = FALSE)

# 1. assume df$X holds gene symbols
colnames(df)[1] <- "gene"

# 2. Keep only non‑TNF samples (drop columns ending with _TNF)
non_tnf_cols <- c(
  "gene",
  grep("_TNF$", colnames(df), invert = TRUE, value = TRUE)
)
df_non_tnf <- df[, non_tnf_cols]

# 3. Subset to IL1b‑exclusive genes
genes_use <- intersect(df_non_tnf$gene, il1_genes$genename)
# Rebuild numeric matrix cleanly
mat <- df_non_tnf %>%
  filter(gene %in% genes_use) %>%
  column_to_rownames("gene")

# Keep only numeric columns (defensive)
num_cols <- sapply(mat, is.numeric)
mat <- as.matrix(mat[, num_cols, drop = FALSE])

# Remove genes with no numeric data
mat <- mat[rowSums(is.na(mat)) < ncol(mat), , drop = FALSE]

# Log2 transform; replace remaining NA with 0 (or a small value) if needed
mat_log <- log2(mat + 1)

# Optionally drop rows that are all zero after log
mat_log <- mat_log[rowSums(mat_log) > 0, , drop = FALSE]

# Z‑score per gene
mat_scaled <- t(scale(t(mat_log)))


# 1) Remove rows with any NA/NaN/Inf
bad_row <- !is.finite(rowSums(mat_scaled))
mat_scaled_clean <- mat_scaled[!bad_row, , drop = FALSE]

# 2) Remove columns with any NA/NaN/Inf (should be rare)
bad_col <- !is.finite(colSums(mat_scaled_clean))
mat_scaled_clean <- mat_scaled_clean[, !bad_col, drop = FALSE]

# 3) Optional: if any row has zero variance, remove (scale() can create NaN)
var0 <- apply(mat_scaled_clean, 1, function(x) var(x, na.rm = TRUE) == 0)
mat_scaled_clean <- mat_scaled_clean[!var0, , drop = FALSE]

# 4) Now plot
pheatmap(
  mat_scaled_clean,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col[colnames(mat_scaled_clean), , drop = FALSE],
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 8,
  main = "IL1b-exclusive autophagy genes (non-treated samples)"
)

#-----------
## plot only RA Knee (healthy were all knee)

# Vector of samples to keep
samples_keep <- c("Healthy_81", "Healthy_82", "Healthy_83",
                  "RA_196", "RA_217", "RA_280", "RA_336", "RA_346", "RA_368")

# Subset and reorder columns
keep <- intersect(samples_keep, colnames(mat_scaled_clean))
mat_sub <- mat_scaled_clean[, keep, drop = FALSE]

# Annotation: Healthy vs RA, in this order
condition <- ifelse(grepl("^Healthy", keep), "Healthy", "RA")
annotation_col <- data.frame(
  Condition = factor(condition, levels = c("Healthy", "RA")),
  row.names = keep
)

png("IL1b_exclusive_autophagy_Healthy_RA_heatmap.png",
    width = 2000, height = 1600, res = 300)
pheatmap(
  mat_sub,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  annotation_col = annotation_col,
  show_rownames = TRUE,
  show_colnames = TRUE,
  fontsize_row = 8,
  fontsize_col = 8,
  main = "IL1b-exclusive autophagy genes (Healthy vs RA)"
)
dev.off()