GSEA analysis
authour: Mim
Date: 5/2/2026

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(clusterProfiler)
library(tibble)
library(forcats)
library(purrr)
library(ComplexUpset)
library(ggplot2)
library(org.Hs.eg.db)
library(enrichplot)
library(ReactomePA)
library(msigdbr)
library(patchwork)
library(pheatmap)
library(ggrepel)

```
Load data

```{r}


# Base directory
base_dir <- "/shares/rheumatologie.usz/mim/GSEA analysis"

# Load raw file
cytokine_data <- readRDS(
  file.path(base_dir, "ra_de_list_only_expressed_genes.rds")
)

names(cytokine_data)

#Load autophagy genes
autophagy <- readr::read_csv(
  file.path(base_dir, "HADB.csv")
)

# Base directory where the GMT files are
gmt_dir <- "/home/plapha/data/Cytokine stimulation gene Tsuchiya et al/autophagy genes/autophagy GSEA"

gmt_files <- c(
  "GOBP_NEGATIVE_REGULATION_OF_AUTOPHAGY.v2026.1.Hs.gmt",
  "GOBP_POSITIVE_REGULATION_OF_AUTOPHAGY.v2026.1.Hs.gmt",
  "GOBP_MITOPHAGY.v2026.1.Hs.gmt",
  "GOBP_REGULATION_OF_AUTOPHAGY.v2026.1.Hs.gmt"
)

# Named list: one TERM2GENE per pathway
term2gene_list <- setNames(
  lapply(gmt_files, function(f) {
    suppressWarnings(read.gmt(file.path(gmt_dir, f)))
  }),
  gsub("\\.v2026\\.1\\.Hs\\.gmt", "", gmt_files)
)

names(term2gene_list)

```
# Prepare TERM2GENE for GSEA
 
```{r}
term2gene <- autophagy %>%
  transmute(
    term = "Autophagy",
    gene = GENE
  ) %>%
  distinct()


```
# Helper functions
 
```{r}
replace_0 <- function(x, eps = 1e-10) {
  x[x == 0 | is.na(x)] <- eps
  x
}

run_gsea <- function(df) {

  stopifnot(all(c("genename", "log2FoldChange", "pvalue") %in% colnames(df)))

  set.seed(123)

  df_ranked <- df %>%
    mutate(
      pvalue = ifelse(pvalue == 0 | is.na(pvalue), 1e-10, pvalue),
      rank = sign(log2FoldChange) * -log10(pvalue)
    ) %>%
    group_by(genename) %>%
    slice_min(order_by = pvalue, n = 1, with_ties = FALSE) %>%
    ungroup()

  genes4gsea <- setNames(df_ranked$rank, df_ranked$genename) %>%
    sort(decreasing = TRUE)

  clusterProfiler::GSEA(
    genes4gsea,
    TERM2GENE = term2gene,
    pvalueCutoff = 1,
    maxGSSize = 1000,
    nPermSimple = 10000,
    seed = TRUE,
    scoreType = "std",
    eps = 0
  )
}


```

## Run GSEA for each cytokine

```{r}

gsea_results <- imap(cytokine_data, function(df, stim) {
  message("Running GSEA for ", stim)
  run_gsea(df)
})

names(gsea_results)


## ===============================
## Format summaries
## ===============================
gsea_summary <- imap_dfr(
  gsea_results,
  ~ as_tibble(.x@result) %>%
      mutate(
        cytokine = .y,
        p_display = case_when(
          p.adjust < 0.001 ~ "< 0.001",
          p.adjust < 0.01  ~ "< 0.01",
          p.adjust < 0.1   ~ "< 0.1",
          TRUE ~ as.character(round(p.adjust, 2))
        ),
        significant = p.adjust < 0.01
      )
)

gsea_summary

```

## Run GSEA for each cytokine --> for gmt file list

```{r}

cytokines_keep <- c("IFNg", "TNFa", "IL1b")

replace_0 <- function(x, eps = 1e-300) {
  x[x == 0] <- eps
  x
}

gsea_all <- imap(
  cytokine_data[cytokines_keep],
  function(df, stim) {
    imap(
      term2gene_list,
      function(t2g, pathway) {
        message("Running ", stim, " – ", pathway)
        run_gsea(df, t2g)
      }
    )
  }
)


```
# Plot GSEA
Helper plotting function

```{r}

plot_gsea_single <- function(gsea_res, title, colors = NULL) {

  stopifnot(inherits(gsea_res, "gseaResult"))

  gseaplot2(
    gsea_res,
    geneSetID = 1,           # only autophagy
    pvalue_table = TRUE,
    color = colors,
    title = title
  )
}

## batch for selected cytokines

cytokines_keep <- c("IL1b")   #"IFNg", "TNFa",

for (stim in cytokines_keep) {
  p <- plot_gsea_single(
    gsea_results[[stim]],
    title = paste0(stim, " ")
  )
  
  print(p)
  
  ggsave(
    filename = paste0("GSEA_", stim, ".pdf"),
    plot = p,
    width = 6,
    height = 5
  )
}

##### run below here if use gmt ####

plot_dir <- "GSEA_plots"
dir.create(plot_dir, showWarnings = FALSE)

plot_gsea_curve <- function(gsea_res, cytokine, pathway) {

  if (nrow(gsea_res@result) == 0) return(NULL)

  p <- gseaplot2(
    gsea_res,
    geneSetID = 1,
    title = paste(cytokine, pathway, sep = " – "),
    pvalue_table = TRUE
  )

  ggsave(
    filename = file.path(
      plot_dir,
      paste0(cytokine, "_", pathway, "_GSEA.pdf")
    ),
    plot = p,
    width = 8,
    height = 6
  )

  p
}


gsea_plots <- imap(
  gsea_all,
  function(pathway_list, cytokine) {
    imap(
      pathway_list,
      function(gsea_res, pathway) {
        plot_gsea_curve(gsea_res, cytokine, pathway)
      }
    )
  }
)

```
# Leading-edge extraction (core enrichment genes)

```{r}
extract_leading_edge <- function(gsea_res, df_deg, cytokine) {

  # get GSEA table
  res <- as_tibble(gsea_res@result)

  # only autophagy term (row 1)
  le_genes <- res$core_enrichment[1] %>%
    strsplit("/") %>%
    unlist()

  # add DEG info
  le_df <- df_deg %>%
    filter(genename %in% le_genes) %>%
    mutate(
      direction = case_when(
        log2FoldChange > 0 ~ "Up",
        log2FoldChange < 0 ~ "Down"
      ),
      cytokine = cytokine
    )

  le_df
}

####from gmt file ####

extract_leading_edge <- function(gsea_res, cytokine, pathway) {

  if (nrow(gsea_res@result) == 0) return(NULL)

  res <- gsea_res@result[1, ]

  tibble(
    genename = unlist(strsplit(res$core_enrichment, "/")),
    cytokine = cytokine,
    pathway = pathway,
    NES = res$NES,
    direction = ifelse(res$NES > 0, "Up", "Down")
  )
}


leading_edge_df <- imap_dfr(
  gsea_all,
  function(pathway_list, cytokine) {
    imap_dfr(
      pathway_list,
      function(gsea_res, pathway) {
        extract_leading_edge(gsea_res, cytokine, pathway)
      }
    )
  }
)

table(leading_edge_df$pathway, leading_edge_df$cytokine)


```
# Run leading-edge extraction

```{r}
cytokines_keep <- c("IL1b")  #"IFNg", "TNFa", 

leading_edge_df <- bind_rows(
  lapply(cytokines_keep, function(stim) {
    extract_leading_edge(
      gsea_res = gsea_results[[stim]],
      df_deg   = cytokine_data[[stim]],
      cytokine = stim
    )
  })
)

leading_edge_df


#-----plot directly from leading edge----

# bar plot -------------------

plot_leading_edge_bar <- function(df, top_n = 20) {
  
  df_plot <- df %>%
    mutate(rank_score = sign(log2FoldChange) * -log10(pvalue)) %>%
    arrange(desc(abs(rank_score))) %>%
    slice_head(n = top_n) %>%
    mutate(
      genename = factor(genename, levels = rev(genename)),
      direction = ifelse(log2FoldChange > 0, "Up", "Down")
    )
  
  ggplot(df_plot, aes(x = genename, y = rank_score, fill = direction)) +
    geom_col() +
    coord_flip() +
    labs(
      title = "Leading-edge genes (Autophagy)",
      x = NULL,
      y = "Rank score"
    ) +
    theme_bw(base_size = 14)
}

p_bar <- plot_leading_edge_bar(leading_edge_df, top_n = 25)
print(p_bar)

ggsave("LeadingEdge_barplot.pdf", p_bar, width = 6, height = 5)



## Volcano --------------------

leading_edge_il1b <- leading_edge_df %>%
  dplyr::filter(cytokine == "IL1b")


plot_leading_edge_only <- function(df_deg, leading_edge,
                                  lfc_thresh = 0.5,
                                  p_thresh = 0.05) {
  
  df_plot <- df_deg %>%
    mutate(
      pvalue = replace_0(pvalue),
      neglog10p = -log10(pvalue)
    ) %>%
    filter(genename %in% leading_edge$genename) %>%
    mutate(
      pass_thresh = abs(log2FoldChange) > lfc_thresh & pvalue < p_thresh
    )
  
  ggplot(df_plot, aes(x = log2FoldChange, y = neglog10p)) +
    geom_point(aes(color = pass_thresh), size = 2) +
    
    # label only meaningful genes
    geom_text_repel(
      data = subset(df_plot, pass_thresh),
      aes(label = genename),
      size = 3,
      max.overlaps = 20
    ) +
    
    geom_hline(yintercept = -log10(p_thresh), linetype = "dashed") +
    geom_vline(xintercept = c(-lfc_thresh, lfc_thresh), linetype = "dashed") +
    
    scale_color_manual(values = c("grey70", "red")) +
    
    labs(
      title = "Leading-edge genes (filtered)",
      x = "log2 Fold Change",
      y = "-log10(pvalue)",
      color = "Significant"
    ) +
    theme_bw(base_size = 14)
}

p_le <- plot_leading_edge_only(
  df_deg = cytokine_data[["IL1b"]],
  leading_edge = leading_edge_il1b,
  lfc_thresh = 0.5,
  p_thresh = 0.001
)

print(p_le)

ggsave("LeadingEdge_only_IL1b.pdf", p_le, width = 6, height = 5)

# plot bar from volcano----------


extract_leading_edge_filtered <- function(df_deg, leading_edge,
                                          lfc_thresh = 0.5,
                                          padj_thresh = 0.05) {
  
  df_deg %>%
    filter(genename %in% leading_edge$genename) %>%
    mutate(padj = ifelse(is.na(padj), 1, padj)) %>%
    filter(
      abs(log2FoldChange) > lfc_thresh,
      padj < padj_thresh
    ) %>%
    arrange(padj)
}

le_filtered <- extract_leading_edge_filtered(
  df_deg = cytokine_data[["IL1b"]],
  leading_edge = leading_edge_il1b,
  lfc_thresh = 0.5,
  padj_thresh = 0.05
)

le_filtered



plot_le_bar <- function(df, top_n = 20) {
  
  df_plot <- df %>%
    slice_head(n = top_n) %>%
    mutate(
      genename = factor(genename, levels = rev(genename)),
      direction = ifelse(log2FoldChange > 0, "Up", "Down"),
      neglog10padj = -log10(padj)
    )
  
  ggplot(df_plot, aes(x = genename, y = log2FoldChange, fill = neglog10padj)) +
    geom_col() +
    coord_flip() +
    
    scale_fill_gradient(
      low = "lightblue",
      high = "red"
    ) +
    
    labs(
      title = "Leading-edge genes (Autophagy, IL1b)",
      x = NULL,
      y = "log2 Fold Change",
      fill = "-log10(padj)"
    ) +
    
    theme_bw(base_size = 14)
}

p_bar <- plot_le_bar(le_filtered, top_n = 20)

print(p_bar)

ggsave("LeadingEdge_bar_IL1b.pdf", p_bar, width = 6, height = 5)



```
# Prepare data for UpSet plot

```{r}
upset_input <- leading_edge_df %>%
  mutate(set = paste(cytokine, direction, sep = "_")) %>%
  select(genename, set) %>%
  distinct()

#Convert to list format (UpSetR style)
upset_list <- split(upset_input$genename, upset_input$set)


```
# Make the UpSet plot

```{r}

leading_edge_df %>%
  mutate(set = paste(cytokine, direction, sep = "_")) %>%
  distinct(genename, set) %>%
  pivot_wider(
    names_from = set,
    values_from = genename,
    values_fn = length,
    values_fill = 0
  ) %>%
  ComplexUpset::upset(
    intersect = colnames(.)[-1],
    name = "Autophagy leading-edge genes"
  )
# sanity check

leading_edge_df %>%
  count(cytokine, direction)


upset_df <- leading_edge_df %>%
  filter(cytokine %in% c("IFNg", "TNFa", "IL1b")) %>%
  mutate(set = paste(cytokine, direction, sep = "_")) %>%
  distinct(genename, set) %>%
  mutate(value = TRUE) %>%
  pivot_wider(
    names_from = set,
    values_from = value,
    values_fill = FALSE
  )

colnames(upset_df)

# plot the UpSet

p <- ComplexUpset::upset(
  upset_df,
  intersect = c("IFNg_Up", "TNFa_Up", "IL1b_Up"),
  name = "Autophagy leading-edge genes",
  width_ratio = 0.2
)

# Save the plot
ggsave(
  filename = "UpSet_Autophagy_IL1b_IFNg_TNFa.pdf",
  plot = p,           # the ggplot object returned by upset()
  width = 6,
  height = 5
)

```
# Upset plot from gmt file

```{r}

upset_dir <- "UpSet_plots"
dir.create(upset_dir, showWarnings = FALSE)

make_upset_df <- function(df, pw_name, direction = "Up") {
  df %>%
    # keep only this pathway and direction
    filter(pathway == pw_name, direction == direction) %>%
    # now distinct gene × cytokine within this pathway only
    distinct(genename, cytokine) %>%
    # assign presence = 1
    mutate(value = 1) %>%
    # pivot only columns for cytokines
    pivot_wider(
      names_from = cytokine,
      values_from = value,
      values_fill = 0
    )
}


lapply(unique(leading_edge_df$pathway), function(pw) {
  df_up <- make_upset_df(leading_edge_df, pw, "Up")
  colSums(df_up[cytokines_keep])  # sum genes per cytokine
})


# the plots

for(pw in unique(leading_edge_df$pathway)) {

  df_upset <- make_upset_df(leading_edge_df, pw, "Up")

  p <- ComplexUpset::upset(
    df_upset,
    intersect = cytokines_keep,
    name = paste("Leading-edge genes –", pw)
  )

  ggsave(
    filename = file.path(upset_dir, paste0(pw, "_UpSet.pdf")),
    plot = p,
    width = 6,
    height = 5
  )
}


```
# Extract genes exclusive to IL1b

```{r}
# Filter for IL1b Up only
il1b_up_exclusive <- leading_edge_df %>%
  filter(cytokine == "IL1b", direction == "Up") %>%
  filter(!genename %in% leading_edge_df$genename[leading_edge_df$cytokine %in% c("IFNg","TNFa")]) %>%
  select(genename, log2FoldChange, pvalue)

il1b_up_exclusive

### for gmt 

get_IL1b_exclusive <- function(df_leading, pw_name, cytokine_col = "IL1b", other_cytokines = c("IFNg", "TNFa")) {
  df_upset <- make_upset_df(df_leading, pw_name, direction = "Up")
  
  df_upset %>%
    filter(!!sym(cytokine_col) == 1) %>%
    filter(rowSums(across(all_of(other_cytokines))) == 0) %>%
    pull(genename)
}

get_IL1b_expr <- function(df_leading, pw_name) {
  genes <- get_IL1b_exclusive(df_leading, pw_name)
  
  cytokine_data$IL1b %>%
    filter(genename %in% genes) %>%
    mutate(
      p_neglog10 = -log10(pvalue)
    )
}

```
# plot for genes exclusive to IL1b

```{r}

# Order genes by log2FoldChange for clarity
plot_df <- il1b_up_exclusive %>%
  arrange(log2FoldChange) %>%
  mutate(genename = factor(genename, levels = genename))

plot_df <- plot_df %>%
  mutate(p_neglog10 = -log10(pvalue))

p <-  ggplot(plot_df, aes(x = genename, y = log2FoldChange, color = p_neglog10)) +
  geom_point(size = 4) +
  scale_color_viridis_c(option = "plasma", direction = -1) +  # reverse: higher = darker = more significance
  theme_bw(base_size = 14) +
  labs(
    title = "IL1b Up-Regulated Exclusive Autophagy Genes",
    x = "Gene",
    y = "log2 Fold Change",
    color = "-log10(p-value)"
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p)

# Save as PDF
ggsave(
  filename = "IL1b_Up_Exclusive_genes.pdf",
  plot = p,
  width = 8,
  height = 5
)


##### for gmt ####

plot_IL1b_exclusive <- function(df_expr, pw_name, save_dir = "IL1b_exclusive_plots") {
  dir.create(save_dir, showWarnings = FALSE)
  
  p <- ggplot(df_expr, aes(x = genename, y = log2FoldChange, color = p_neglog10)) +
    geom_point(size = 3) +
    scale_color_viridis_c(option = "viridis") +
    geom_hline(yintercept = 0, linetype = "dashed") +
    theme_bw(base_size = 14) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(
      title = paste("IL1b Up-regulated exclusive genes –", pw_name),
      x = "Gene",
      y = "log2 Fold Change",
      color = "-log10(p-value)"
    )
  
  ggsave(
    filename = file.path(save_dir, paste0("IL1b_", pw_name, "_exclusive.pdf")),
    plot = p,
    width = 7,
    height = 5
  )
  
  p
}

plots_IL1b <- map(
  unique(leading_edge_df$pathway),
  function(pw) {
    df_expr <- get_IL1b_expr(leading_edge_df, pw)
    if(nrow(df_expr) > 0) {
      plot_IL1b_exclusive(df_expr, pw)
    } else {
      message("No IL1b-exclusive genes in ", pw)
      NULL
    }
  }
)

names(plots_IL1b) <- unique(leading_edge_df$pathway)


```
# pathway analysis from KEGG

```{r}

#Select cytokine dataset

deg_df <- cytokine_data$IL1b

head(deg_df)

## 3. Convert gene symbol -> Entrez ID
gene_df <- bitr(
  deg_df$genename,
  fromType = "SYMBOL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

## merge with DEG table
gene_df <- gene_df %>%
  left_join(deg_df, by = c("SYMBOL" = "genename"))

gene_df <- gene_df %>%
  group_by(ENTREZID) %>%
  slice_max(order_by = abs(log2FoldChange), n = 1) %>%
  ungroup()

## ===============================
## 4. Create ranked gene list
## ===============================

gene_list <- gene_df$log2FoldChange
names(gene_list) <- gene_df$ENTREZID

gene_list <- sort(gene_list, decreasing = TRUE)

## ===============================
## 5. Run KEGG GSEA
## ===============================

      # gsea_res <- gseKEGG(
      #   geneList = gene_list,
      #   organism = "hsa",
      #   pvalueCutoff = 0.05,
      #   use_internal_data = TRUE
      # )
      # 
      # ## check results
      # head(gsea_res@result)

### 5. Run Reactome GSEA

gsea_res <- gsePathway(
  geneList = gene_list,
  organism = "human",
  pvalueCutoff = 0.05
)

head(gsea_res@result)

head(gsea_res@result[, c("Description","NES","p.adjust")], 20)

## ===============================
## 6. Visualization
## ===============================

## Dotplot (best overview)
dotplot(gsea_res, showCategory = 20) +
  ggtitle("Pathway Enrichment")

## Ridgeplot
ridgeplot(gsea_res, showCategory = 20)


#To specifically check autophagy / mitophagy
gsea_res@result %>%
  dplyr::filter(grepl("autophagy|mitophagy|lysosome|mitochond", Description, ignore.case = TRUE))



## ===============================
## 7. Plot specific pathway
## ===============================

# choose pathway ID
gseaplot2(
  gsea_res,
  geneSetID = gsea_res@result$ID[1],
  title = gsea_res@result$Description[1]
)

## ===============================
## 8. Save results
## ===============================

write.csv(
  gsea_res@result,
  file = file.path(base_dir, "KEGG_GSEA_results.csv"),
  row.names = FALSE
)

```
# Run GSEA on database MSigDB.


```{r}
# Recreate gene list using gene symbols
deg_df <- cytokine_data$IL1b

gene_list <- deg_df$log2FoldChange
names(gene_list) <- deg_df$genename

gene_list <- sort(gene_list, decreasing = TRUE)


gene_list <- gene_list[!duplicated(names(gene_list))]

# get gene set from database
msig <- msigdbr(species = "Homo sapiens")

mito_sets <- msig %>%
  filter(gs_name %in% c(
    "REACTOME_AUTOPHAGY",
    "REACTOME_MITOPHAGY",
    "REACTOME_MACROAUTOPHAGY",
    "REACTOME_PINK_PARKIN_MEDIATED_MITOPHAGY",
    "REACTOME_LYSOSOME_VESICLE_BIOGENESIS",
    "REACTOME_RESPIRATORY_ELECTRON_TRANSPORT"
  )) %>%
  split(x = .$gene_symbol, f = .$gs_name)

# Convert to TERM2GENE format
term2gene <- msig %>%
  filter(gs_name %in% c(
    "REACTOME_AUTOPHAGY",
    "REACTOME_MITOPHAGY",
    "REACTOME_MACROAUTOPHAGY",
    "REACTOME_PINK_PARKIN_MEDIATED_MITOPHAGY",
    "REACTOME_LYSOSOME_VESICLE_BIOGENESIS",
    "REACTOME_RESPIRATORY_ELECTRON_TRANSPORT"
  )) %>%
  dplyr::select(gs_name, gene_symbol)

# run GSEA
gsea_res <- GSEA(
  geneList = gene_list,
  TERM2GENE = term2gene,
  pvalueCutoff = 1
)

gsea_res@result

# Plot

gseaplot2(gsea_res, geneSetID = "REACTOME_AUTOPHAGY") 
gseaplot2(gsea_res, geneSetID = "REACTOME_MITOPHAGY")
gseaplot2(gsea_res, geneSetID = "REACTOME_RESPIRATORY_ELECTRON_TRANSPORT")

# check if pathway is enriched
gsea_res@result[, c("ID","NES","pvalue","p.adjust")]

````
# nicer Plot

```{R}

plot_and_save_gsea <- function(gsea_res, pathway, filename){

  p <- gseaplot2(
    gsea_res,
    geneSetID = pathway,
    subplots = c(1,2,3),
    rel_heights = c(1.5,0.5,1),
    title = gsub("REACTOME_", "", pathway),
    pvalue_table = TRUE   # automatically adds NES, p, padj
  )

  ggsave(paste0(filename, ".png"), p, width = 8, height = 6, dpi = 300)
  ggsave(paste0(filename, ".pdf"), p, width = 8, height = 6)

  return(p)
}


## generate plots


plot_and_save_gsea(gsea_res,"REACTOME_AUTOPHAGY","GSEA_AUTOPHAGY")

plot_and_save_gsea(gsea_res,"REACTOME_MITOPHAGY","GSEA_MITOPHAGY")

plot_and_save_gsea(gsea_res,"REACTOME_RESPIRATORY_ELECTRON_TRANSPORT","GSEA_ETC")


````
# Leading edge 

```{R}
# Extract leading edge string
le_string <- gsea_res@result %>%
  filter(ID == "REACTOME_MITOPHAGY") %>%
  pull(core_enrichment)

# Convert to vector
leading_edge_genes <- unlist(strsplit(le_string, "/"))

leading_edge_genes


## Get fold changes for those genes
le_df <- data.frame(
  gene = leading_edge_genes,
  log2FC = gene_list[leading_edge_genes]
) %>%
  na.omit() %>%
  arrange(desc(log2FC))

le_df$padj <- cytokine_data$IL1b$padj[
  match(le_df$gene, cytokine_data$IL1b$genename)
]

# plot


p <- ggplot(le_df, aes(x = reorder(gene, log2FC), y = log2FC)) +
  geom_col(aes(fill = log2FC)) +
  geom_point(aes(size = -log10(padj)), color = "black") +
  coord_flip() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(
    title = "Mitophagy Leading Edge Genes",
    y = "log2 Fold Change",
    size = "-log10(padj)"
  ) +
  theme_minimal()

ggsave("Mitophagy_LeadingEdge.png", plot = p, width = 7, height = 5, dpi = 300)

```
# Overlay leading edge genes on GSEA curve

```{R}
le_string <- gsea_res@result$core_enrichment[
  gsea_res@result$ID == "REACTOME_MITOPHAGY"
]

leading_edge_genes <- unlist(strsplit(le_string, "/"))

gene_rank <- data.frame(
  gene = names(gene_list),
  rank = seq_along(gene_list),
  value = gene_list
)

gene_rank$leading_edge <- gene_rank$gene %in% leading_edge_genes

ggplot(gene_rank, aes(x = rank, y = value)) +
  geom_line(color = "grey70") +
  geom_point(
    data = subset(gene_rank, leading_edge),
    aes(x = rank, y = value),
    color = "red",
    size = 2
  ) +
  labs(
    title = "Leading Edge Genes on Ranked List",
    x = "Gene Rank",
    y = "log2 Fold Change"
  ) +
  theme_minimal()

ggsave("Mitophagy_LeadingEdge_overlay.pdf", width = 7, height = 5)


```
# Gene–pathway network plot

```{R}
library(enrichplot)

p_net <- cnetplot(
  gsea_res,
  showCategory = "REACTOME_MITOPHAGY",
  foldChange = gene_list
)

p_net

ggsave("Mitophagy_network.pdf", plot = p_net, width = 8, height = 6)