```{r}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(tidyr)
library(dplyr)
library(uwot)
library(ggpubr)
```

# path to files

```{r}

file_plate <- paste0(
  "/scratch/plapha/250410 metabolism_RBPJ/output/",
  "MyExpt_Cells.csv"
)

sce <- loadData(file_plate)

```
## Data Transformation

Prepare the data for machine learning.

```{r}
# fix plate id
sce$Plate <- unique(sce$Patient)[1]
 
# remove missing values in cells
mat <- assay(sce, name = "features")
miss_cells <- apply(mat, 2, function(x) sum(is.na(x)))
cell_ids <- which(miss_cells > 0)
sce <- sce[, -cell_ids]
 
# define reference level
sce$Treatment <- as.factor(sce$Treatment)
sce$Treatment <- relevel(sce$Treatment, ref = "cont")
 
# batch correct and normalize features
sce <- normalizeExclude(sce)
 
plotCellsPerImage(sce)
sce <- removeOutliers(sce, min = 20, max = 100)

```

## Unsupervised Analysis

Use `scater` for exploratory data analysis.

PCA:

```{r}
sce <- runPCA(sce, exprs_values = "tfmfeatures")
plotReducedDim(sce, dimred = "PCA", colour_by = "Patient")
plotReducedDim(sce, dimred = "PCA", colour_by = "Treatment")
plotPCACor(sce, filter_by = 1, top = 20)
plotPCACor(sce, filter_by = 2, top = 20)

```
UMAP:

```{r}
sce <- runUMAP(sce, exprs_values = "tfmfeatures")
plotReducedDim(sce, dimred = "UMAP", colour_by = "Patient")
plotReducedDim(sce, dimred = "UMAP", colour_by = "Treatment")

```
flter only Intensity data

```{r}
#EXTRACT NORMALISED FEATURES
mat <- assay(sce, "features")

#Get only 'Intensity' features
intensity_features <- grep("^Intensity_", rownames(mat), value = TRUE)
mat_intensity <- mat[intensity_features, ]

#transpose for PCA
mat_pca <- t(mat_intensity)

#run PCA
pca <- prcomp(mat_pca, scale. = TRUE)

# Calculate % variance for each PC
percentVar <- pca$sdev^2 / sum(pca$sdev^2) * 100
pc1_label <- paste0("PC1 (", round(percentVar[1], 1), "%)")
pc2_label <- paste0("PC2 (", round(percentVar[2], 1), "%)")

#plot PCA
pca_df <- as.data.frame(pca$x)
pca_df$Treatment <- sce$Treatment  # add metadata if needed

ggplot(pca_df, aes(x = PC1, y = PC2, color = Treatment)) +
  geom_point(alpha = 0.3, size = 1.5) +
  theme_minimal() +
  labs(
    title = "PCA of Intensity Features",
    x = pc1_label,
    y = pc2_label
  ) +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )

```
UMAp from Intensity data

```{r}


mat_umap <- t(mat_intensity)

set.seed(42)  # for reproducibility
umap_result <- umap(mat_umap, n_neighbors = 15, min_dist = 0.1, metric = "euclidean")

#plot UMAP
umap_df <- as.data.frame(umap_result)
colnames(umap_df) <- c("UMAP1", "UMAP2")
umap_df$Treatment <- sce$Treatment
umap_df$Patient <- sce$Patient

ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Treatment)) +
  geom_point(alpha = 0.3, size = 1.5) +
  facet_wrap(~ Patient) +
  theme_minimal() +
  labs(
    title = "UMAP of Intensity Features",
    x = "UMAP1",
    y = "UMAP2"
  ) +
  theme(
    legend.title = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )

```
## Fit Single Model from all features


Train machine learning model on one perturbation.

```{r}
result <- fitModel(sce,
                   target = "Treatment", 
                   interest_level = "RBPJ", reference_level = "cont", 
                   group = "Patient", strata = c("Disease"), 
                   n_folds = 20, n_threads = 8)


##test diff if panel can distinguish between disease

sce_RBPJ <- sce[, sce$Treatment == "RBPJ"]
 
result <- fitModel(sce_RBPJ,
                   target = "Disease", 
                   interest_level = "RA", reference_level = "Healthy", 
                   group = "Patient", strata = "Disease",
                   n_folds = 20, n_threads = 8)

```

Plot ROC curves.

```{r}
plotROC(result)

```
plot specific feature

```{r}
# 1. Extract metadata and intensity values
meta_data <- as.data.frame(colData(sce))
intensity_values <- mat_intensity["Intensity_MeanIntensity_CorrEdU", ]

# 2. Convert to a numeric vector (in case it's stored differently)
intensity_values <- as.numeric(intensity_values)

# 3. Create a data frame for plotting
plot_data <- data.frame(Intensity = intensity_values)

# 4. Plot using ggplot2 (set cut-off at 0.001)

ggplot(data.frame(Intensity = intensity_values), aes(x = Intensity)) +
  geom_histogram(fill = "skyblue", color = "black", bins = 30) +
  geom_vline(xintercept = 0.001, color = "red", linetype = "dashed", linewidth = 1) +  # Add vertical line
  ggtitle("Histogram of Intensity_MeanIntensity_CorrEdU") +
  xlab("Intensity Value") +
  ylab("Frequency") +
  theme_minimal()

#5. plot percentage
# Combine into a data frame
plot_data <- data.frame(
  Patient = meta_data$Patient,
  Treatment = meta_data$Treatment,
  Intensity = intensity_values
)

# Calculate % cells > 0.001 per Patient-Treatment group
summary_data <- plot_data %>%
  group_by(Patient, Treatment) %>%
  summarise(
    Percent_Above = 100 * mean(Intensity > 0.001),
    .groups = "drop"
  )

# plot
ggplot(summary_data, aes(x = Treatment, y = Percent_Above, color = Patient, group = Patient)) +
    geom_point(size = 3) +
    geom_line() +
    labs(
        title = "% Cells with Intensity > 0.001",
        y = "% Cells Above Threshold",
        x = "Treatment"
    ) +
    theme_minimal()

#test stat
## paired t-test
# Convert to wide format (one row per Patient, columns for each Treatment)
wide_data <- summary_data %>%
  pivot_wider(
    names_from = Treatment, 
    values_from = Percent_Above
  )

# Check structure (replace "TreatmentA" and "TreatmentB" with actual names)
head(wide_data)

t_test_result <- t.test(
     wide_data$cont,  # Replace with actual column names
     wide_data$RBPJ,
     paired = TRUE          # Now works!
 )
 t_test_result
 
 
```
Explore Lipid feature

```{r}

lipid_intensity <- as.numeric(mat_intensity["Intensity_MeanIntensity_CorrLipid", ])

plot_data <- data.frame(
  Patient = meta_data$Patient,
  Treatment = meta_data$Treatment,
  Lipid_Intensity = lipid_intensity
) %>%
  filter(!is.na(Lipid_Intensity))


ggplot(plot_data, aes(x = Lipid_Intensity, fill = Patient)) +
    geom_histogram(color = "black", bins = 30, alpha = 0.7, position = "identity") +
    geom_vline(xintercept = 0.095, color = "red", linetype = "dashed", linewidth = 1) +
    facet_wrap(~ Treatment, scales = "free") +  # Facet by Treatment
    labs(
        title = "Histogram of Lipid Intensity (Threshold = 0.001)",
        x = "Lipid Intensity Value",
        y = "Frequency"
    ) +
    theme_minimal() +
    theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

# 1. Calculate % above threshold for each Patient-Treatment group
percent_data <- plot_data %>%
  group_by(Patient, Treatment) %>%
  summarise(
    Percent_Above = 100 * mean(Lipid_Intensity > 0.095),
    .groups = "drop"
  )

# 2. Plot lines connecting Treatments for each Patient
ggplot(percent_data, aes(x = Treatment, y = Percent_Above, 
                         group = Patient, color = Patient)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  labs(
    title = "% Cells with Lipid Intensity > 0.095",
    y = "% Cells Above Threshold",
    x = "Treatment"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  # Optional: Add mean difference p-value
  stat_compare_means(
    aes(group = Patient), 
    method = "t.test", 
    paired = TRUE,
    label = "p.signif",
    label.y = max(percent_data$Percent_Above) * 1.1
  )

 
```
Explore MitoSOX feature

```{r}
sox_intensity <- as.numeric(mat_intensity["Intensity_MeanIntensity_CorrMitosox", ])

plot_data <- data.frame(
  Patient = meta_data$Patient,
  Treatment = meta_data$Treatment,
  Sox_Intensity = sox_intensity
) %>%
  filter(!is.na(Sox_Intensity))


ggplot(plot_data, aes(x = Sox_Intensity, fill = Patient)) +
    geom_histogram(color = "black", bins = 30, alpha = 0.7, position = "identity") +
    geom_vline(xintercept = 0.08, color = "red", linetype = "dashed", linewidth = 1) +
    facet_wrap(~ Treatment, scales = "free") +  # Facet by Treatment
    labs(
        title = "Histogram of MitoSOX Intensity",
        x = "MitoSOX Intensity Value",
        y = "Frequency"
    ) +
    theme_minimal() +
    theme(
        legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1)
    )

# 1. Calculate % above threshold for each Patient-Treatment group
percent_data <- plot_data %>%
  group_by(Patient, Treatment) %>%
  summarise(
    Percent_Above = 100 * mean(Sox_Intensity > 0.08),
    .groups = "drop"
  )