---
title: "Metabolism"
author: "Phatthamon"
date: now
format: html
editor: source
embed-resources: true
---

## Load Data

```{r}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(stringr)
library(dplyr)
library(batchelor)
library(BiocParallel)
library(tidyr)
library(MultiAssayExperiment)
library(lme4)
library(lmerTest)
library(ggrepel)
library(ggpubr)
library(viridis)
library(rstatix)
library(RColorBrewer)
library(readr)
n_threads <- 8

```

Path to files.

```{r}
  # path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# metabolism panel
files_met <- c(
  paste0(base, "250709 cytokines_metabolism/MyExpt_Cells-plate1.csv"), 
  paste0(base, "250709 cytokines_metabolism/MyExpt_Cells-plate2.csv"), 
  paste0(base, "250717 cytokines_metabolism lot2/MyExpt_Cells_plate1.csv"),
  paste0(base, "250717 cytokines_metabolism lot2/MyExpt_Cells_plate2.csv")
)
sce <- Reduce(cbind, lapply(files_met, loadData))

```
Path to files.-> develop panel exp.

```{r}
# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

#stress panel develope plate
files_met <- c(paste0(base, "250228 lipid panel_hand/MyExpt_Cells.csv"))

sce <- Reduce(cbind, lapply(files_met, loadData))


```
## Data Transformation

Prepare the data for machine learning.

```{r}
data_transformation <- function(sce) {

  # per plate
  plates <- unique(sce[["Plate"]])

  plate_list <- lapply(plates, function(plate_id) {

    # normalize per plate
    sce_plate <- sce[, sce[["Plate"]] == plate_id]

    # remove cells with missing features
    sce_plate <- removeNAs(sce_plate)
    
    # remove outlier images
    sce_plate <- removeOutliers(sce_plate, min = 5, max = 100)
    
    # remove outlier cells
    stats <- perCellQCMetrics(sce_plate, assay.type = "features")
    sce_plate <- sce_plate[, !isOutlier(stats$sum)]
    
    # remove features with zero spread
    sce_plate <- removeLowVariance(sce_plate, robust = TRUE)
    
    # remove zero-inflated features
    sce_plate <- removeZeroInflation(sce_plate)
    
    # transform non-negatives features and scale all features
    sce_plate <- transformLogScale(sce_plate, robust = TRUE)
  
    # fix factors
    sce_plate$Treatment <- as.factor(sce_plate$Treatment)
    sce_plate$Treatment <- relevel(sce_plate$Treatment, ref = "control")
    
    sce_plate
  
  })

  # combine plates
  feature_name_list <- lapply(plate_list,
                              function(sce_plate) rownames(sce_plate))
  include_features <- Reduce(intersect, feature_name_list)
  plate_list <- lapply(plate_list,
                       function(sce_plate) sce_plate[include_features, ])
  sce <- do.call(cbind, plate_list)
  sce$Plate <- as.factor(sce$Plate)
  sce
  
}
sce <- data_transformation(sce)

lapply(unique(sce[["Plate"]]), function(plate_id) {
  
  plotCellsPerImage(sce[, sce[["Plate"]] == plate_id])
  
})

```

Pseudo-bulk over images.

```{r}
sce_aggr <- aggregateAcrossCells(
  sce,
  id = colData(sce)[, c("ImageNumber", "Well", "Patient", "Plate")], 
  use.assay.type = "tfmfeatures", 
  statistics = "mean"
  )

```
#compare by Disease -> select feature that drive differences?code from CHAT

````{r}
sce_sub <- sce_aggr[, sce_aggr$Treatment == "control"]

# if include every feature
# Extract ALL features from tfmfeatures assay
all_features_mat <- assay(sce_sub, "tfmfeatures")

# Transpose so rows = cells, columns = features
all_features_data <- as.data.frame(t(all_features_mat))

# Metadata
meta <- as.data.frame(colData(sce_sub)[, c("Patient", "Disease", "Treatment", "Plate")])

# Combine metadata and features
df <- cbind(meta, all_features_data)

# ANOVA / Kruskal-Wallis (nonparametric)
# to select single feature differ among diseases.

## for Intensity feature
results <- df %>%
  pivot_longer(cols = starts_with("Intensity_"), names_to = "Feature", values_to = "Value") %>%
  group_by(Feature) %>%
  kruskal_test(Value ~ Disease) %>%
  adjust_pvalue(method = "bonferroni") %>%
  arrange(p.adj)

# then Post-hoc pairwise test
pairwise <- df %>%
  pivot_longer(cols = starts_with("Intensity_"), names_to = "Feature", values_to = "Value") %>%
  group_by(Feature) %>%
  pairwise_wilcox_test(Value ~ Disease, p.adjust.method = "bonferroni")

## for all feature
# results <- df %>%
#   pivot_longer(
#     cols = -c(Patient, Disease, Treatment, Plate),  # pivot all other columns
#     names_to = "Feature",
#     values_to = "Value"
#   ) %>%
#   group_by(Feature) %>%
#   kruskal_test(Value ~ Disease) %>%
#   adjust_pvalue(method = "bonferroni") %>%
#   arrange(p.adj)
# 
# # --- Pairwise post-hoc test ---
# pairwise <- df %>%
#   pivot_longer(
#     cols = -c(Patient, Disease, Treatment, Plate),
#     names_to = "Feature",
#     values_to = "Value"
#   ) %>%
#   group_by(Feature) %>%
#   pairwise_wilcox_test(Value ~ Disease, p.adjust.method = "bonferroni")

top_features <- results %>% filter(p.adj < 0.05) %>% arrange(p.adj)
head(top_features)

top10_feature_names <- top_features$Feature[1:15]

# Filter df for top 10 features
df_top10 <- df %>%
  select(Patient, Disease, all_of(top10_feature_names)) %>%
  pivot_longer(cols = all_of(top10_feature_names),
               names_to = "Feature",
               values_to = "Value")

# Summarize per patient (mean per patient)
df_plot <- df_top10 %>%
  group_by(Patient, Disease, Feature) %>%
  summarise(MeanValue = mean(Value), .groups = "drop")


# Plot with bars for group mean and dots for patients
df_summary <- df_plot %>%
  group_by(Disease, Feature) %>%
  summarise(GroupMean = mean(MeanValue), .groups = "drop")

pairwise_top <- pairwise %>%
  filter(Feature %in% top10_feature_names) %>%
  mutate(
    p_label = case_when(
      p.adj <= 0.001 ~ "***",
      p.adj <= 0.01  ~ "**",
      p.adj <= 0.05  ~ "*",
      TRUE ~ "ns"
    )
  )

# --- 2. Loop over top features to plot individually ---
for (f in top10_feature_names) {
  
  df_f <- df_plot %>% filter(Feature == f)
  df_sum_f <- df_summary %>% filter(Feature == f)
  stats_f <- pairwise_top %>% filter(Feature == f)
  
  # Skip if no results
  if (nrow(stats_f) == 0) {
    message(paste("⚠️ Skipping", f, "- no pairwise test results found."))
    next
  }
  
  # --- Add y positions for each comparison dynamically ---
  y_max <- max(df_f$MeanValue, na.rm = TRUE)
  stats_f <- stats_f %>%
    arrange(p.adj) %>%
    mutate(
      y.position = seq(y_max * 1.05, y_max * 1.05 + (n() - 1) * (y_max * 0.1), length.out = n())
    )
  
  # --- Plot ---
  p <- ggplot() +
    geom_bar(
      data = df_sum_f,
      aes(x = Disease, y = GroupMean, fill = Disease),
      stat = "identity", alpha = 0.4, width = 0.6, color = "black"
    ) +
    geom_jitter(
      data = df_f,
      aes(x = Disease, y = MeanValue, color = Disease),
      width = 0.15, size = 3, alpha = 0.9
    ) +
    stat_pvalue_manual(
      stats_f,
      label = "p_label",
      tip.length = 0.01,
      hide.ns = TRUE,
      size = 8,
      bracket.size = 0.8,
      step.increase = 0.08   # stagger brackets automatically
    ) +
    scale_fill_brewer(palette = "Set2") +
    scale_color_brewer(palette = "Set2") +
    theme_bw(base_size = 20) +
    labs(
      title = paste0("Feature: ", f),
      y = "Mean Intensity per Patient",
      x = "Disease"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
      axis.title = element_text(face = "bold", size = 16),
      axis.text = element_text(size = 16),
      legend.position = "none"
    )
  # --- Save each plot ---
  file_base <- gsub("[:/\\s]", "_", f)  # clean feature name for filename
  ggsave(
    filename = paste0("Plot_", file_base, ".tiff"),
    plot = p,
    width = 6,
    height = 6,
    units = "in",
    dpi = 600,
    compression = "lzw"
  )
  print(p)
}


````
# Feature importance analysis _ code base on carpenter-singh-lab

````{r}
# --- 1️⃣ Load required packages ---
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(randomForest)
library(car)       # for VIF
library(caret)     # optional for data preprocessing

# --- 2️⃣ User-adjustable parameters ---
assay_name <- "tfmfeatures"        # assay name
target_var <- "Treatment"          # "Treatment" or "Disease"
comparison_levels <- c("control", "IL1")   # groups to compare "control", "IL1", "Healthy", "OA", "RA
feature_keyword <- "Intensity"             # e.g. "Intensity" or "" for all or "Corr"
topN <- 10                         # number of features to plot
cor_threshold <- 0.7               # correlation clustering depth
vif_limit <- 10                    # VIF limit
use_vif <- TRUE                    # ✅ toggle ON/OFF for VIF filtering
set.seed(123)

# --- 3️⃣ Extract features + metadata ---
features <- as.data.frame(t(assay(sce_aggr, assay_name)))
meta <- as.data.frame(colData(sce_aggr)[, c("Patient", "Disease", "Treatment", "Plate")])
df <- cbind(meta, features)

# --- 4️⃣ Filter for comparison groups ---
df <- df %>%
    filter(.data[[target_var]] %in% comparison_levels) %>%
    mutate(across(all_of(target_var), ~ factor(.x, levels = comparison_levels)))

# --- 5️⃣ Select features (optional keyword filtering) ---
if (feature_keyword != "") {
    selected_features <- grep(feature_keyword, colnames(df), value = TRUE)
    if (length(selected_features) == 0) stop(paste0("❌ No features found containing keyword: ", feature_keyword))
    X <- df %>% select(all_of(selected_features))
} else {
    X <- df %>% select(where(is.numeric))
}
y <- df[[target_var]]

# --- 6️⃣ Compute Pearson correlation + clustering ---
cor_mat <- cor(X, use = "pairwise.complete.obs", method = "pearson")
dist_mat <- as.dist(1 - abs(cor_mat))
hc <- hclust(dist_mat, method = "average")

# --- 7️⃣ Form clusters based on correlation distance threshold ---
clusters <- cutree(hc, h = (1 - cor_threshold))
cluster_df <- data.frame(Feature = names(clusters), Cluster = clusters)

# --- 8️⃣ Pick representative feature per cluster ---
representative_features <- cluster_df %>%
    group_by(Cluster) %>%
    summarise(Representative = Feature[ceiling(length(Feature) / 2)], .groups = "drop")

X_reduced <- X[, representative_features$Representative, drop = FALSE]

# --- 9️⃣ (Optional) Check VIF for multicollinearity ---
if (use_vif && ncol(X_reduced) <= 1000) {
    message("🔍 Checking VIF for multicollinearity...")
    
    dummy_y <- rnorm(nrow(X_reduced))  # dummy numeric response
    df_vif <- data.frame(dummy_y, X_reduced)
    
    vif_values <- tryCatch({
        model <- lm(dummy_y ~ ., data = df_vif)
        vif(model)
    }, error = function(e) {
        message("⚠️ Skipping VIF filtering due to computation error: ", e$message)
        NULL
    })
    
    if (!is.null(vif_values)) {
        keep <- names(vif_values[vif_values < vif_limit])
        if (length(keep) > 0) {
            X_reduced <- X_reduced[, keep, drop = FALSE]
            message("✅ Retained ", length(keep), " features after VIF filtering.")
        } else {
            message("⚠️ No features passed VIF filtering; skipping this step.")
        }
    }
} else if (!use_vif) {
    message("ℹ️ VIF filtering skipped (use_vif = FALSE).")
} else {
    message("⚠️ Skipping VIF filtering due to large feature count (", ncol(X_reduced), ").")
}

# --- 🔟 Random Forest on reduced features ---
rf_model <- randomForest(
    x = X_reduced,
    y = y,
    importance = TRUE,
    ntree = 1000
)

importance_df <- as.data.frame(importance(rf_model, type = 1))
importance_df$Feature <- rownames(importance_df)
importance_df <- importance_df %>% arrange(desc(MeanDecreaseAccuracy))

# --- 11️⃣ Build title + filename ---
comparison_str <- paste(comparison_levels, collapse = "_vs_")
feature_str <- if (feature_keyword != "") paste0("_", feature_keyword) else ""

plot_title <- paste0(
    "Top ", topN, " Representative Features for ", target_var,
    " (", comparison_str, ")", feature_str
)

file_name <- paste0(
    "TopFeatures_", target_var, "_", comparison_str, feature_str, ".pdf"
)
file_name <- gsub("\\s+", "_", file_name)


# --- 12️⃣ Plot top features ---
top_features_plot <- ggplot(importance_df[1:topN, ], aes(
    x = reorder(Feature, MeanDecreaseAccuracy),
    y = MeanDecreaseAccuracy
)) +
    geom_bar(stat = "identity", fill = "steelblue") +
    coord_flip() +
    theme_bw(base_size = 18) +
    labs(
        title = plot_title,
        x = "Representative Feature",
        y = "Importance (Mean Decrease Accuracy)"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)))

# --- 13️⃣ Save publication-quality plot ---
ggsave(
  filename = file_name,
  plot = top_features_plot,
  width = 8,
  height = 6,
  units = "in"
)

cat("✅ Saved high-resolution plot as:", file_name, "\n")



````

# Feature importance ----> new code
new median feature method

````{r}
library(mrfDepth)
library(SingleCellExperiment)
library(dplyr)
library(ggplot2)
library(randomForest)
library(car)       # for VIF
library(caret)
library(distops)

# --- 2️⃣ User-adjustable parameters ---
assay_name <- "tfmfeatures"        # assay name
target_var <- "Treatment"          # "Treatment" or "Disease"
comparison_levels <- c("control", "IL1")   # groups to compare "cont", "MG","Healthy", "OA", "RA", "control", "IL1"
feature_keyword <- "Corr"              # e.g. "Intensity" or "" for all
topN <- 10                         # number of features to plot
vif_limit <- 10                    # VIF limit
use_vif <- TRUE                    # ✅ toggle ON/OFF for VIF filtering
set.seed(123)
treatment_filter <- NULL          # "control"

# --- 3️⃣ Extract features + metadata ---
features <- as.data.frame(t(assay(sce_aggr, assay_name)))     #sce_aggr
meta <- as.data.frame(colData(sce_aggr)[, c("Patient", "Disease", "Treatment", "Plate")])
df <- cbind(meta, features)

if (!is.null(treatment_filter)) {
  message("Filtering to Treatment == '", treatment_filter, "'")
  df <- df %>% filter(Treatment == treatment_filter)
} else {
  message("Using all treatments (no Treatment filter)")
}


# --- 4️⃣ Filter for comparison groups ---
df <- df %>%
    filter(.data[[target_var]] %in% comparison_levels) %>%
    mutate(across(all_of(target_var), ~ factor(.x, levels = comparison_levels)))

# Check how many samples remain after filtering
cat("Samples after filtering:", nrow(df), "\n")
cat("Treatment distribution:\n")
print(table(df$Treatment))
cat("Disease distribution:\n")
print(table(df[[target_var]]))


# --- 5️⃣ Select features (optional keyword filtering) ---
if (feature_keyword != "") {
    selected_features <- grep(feature_keyword, colnames(df), value = TRUE)
    if (length(selected_features) == 0) stop(paste0("❌ No features found containing keyword: ", feature_keyword))
    X <- df %>% select(all_of(selected_features))
} else {
    X <- df %>% select(where(is.numeric))
}
y <- df[[target_var]]

# ============================================================================
# === Search for best cor_threshold that maximizes retained features under VIF < 10 ===
# ============================================================================

# Define candidate thresholds (higher = less aggressive clustering → more features initially)
threshold_candidates <- seq(0.50, 0.95, by = 0.025)  # or seq(0.6, 0.9, by = 0.05)

# Storage for results
results <- data.frame(
    cor_threshold = numeric(),
    n_clusters    = integer(),
    n_medoids     = integer(),
    n_after_vif   = integer(),
    max_vif       = numeric(),
    passed_vif    = logical(),
    stringsAsFactors = FALSE
)

best_threshold   <- NA
best_n_features  <- -Inf
best_X_reduced   <- NULL
best_importance  <- NULL

for (th in threshold_candidates) {
    message("\nTesting cor_threshold = ", sprintf("%.3f", th), " ...")
    
    # --- Clustering ---
    cor_mat <- cor(X, use = "pairwise.complete.obs", method = "pearson")
    dist_mat <- as.dist(1 - abs(cor_mat))
    hc <- hclust(dist_mat, method = "average")
    clusters <- cutree(hc, h = (1 - th))
    
    # --- Medoids ---
    rep_ids <- find_medoids(dist_mat, as.factor(clusters))
    X_temp <- X[, rep_ids, drop = FALSE]
    
    n_medoids <- ncol(X_temp)
    
    # --- VIF filtering ---
    vif_ok <- FALSE
    max_vif_val <- Inf
    n_after <- 0
    
    if (ncol(X_temp) <= 1000) {
        dummy_y <- rnorm(nrow(X_temp))
        df_vif <- data.frame(dummy_y, X_temp)
        
        vif_values <- tryCatch({
            model <- lm(dummy_y ~ ., data = df_vif)
            vif(model)
        }, error = function(e) {
            message("  VIF error: ", e$message)
            NULL
        })
        
        if (!is.null(vif_values)) {
            max_vif_val <- max(vif_values, na.rm = TRUE)
            keep <- names(vif_values[vif_values < vif_limit])
            if (length(keep) > 0) {
                X_after_vif <- X_temp[, keep, drop = FALSE]
                n_after <- ncol(X_after_vif)
                
                # Re-check max VIF after removal (should be < 10, but confirm)
                if (n_after > 1) {
                    df_vif2 <- data.frame(dummy_y, X_after_vif)
                    vif_after <- tryCatch(vif(lm(dummy_y ~ ., df_vif2)), error = function(e) Inf)
                    max_vif_val <- max(vif_after, na.rm = TRUE)
                } else {
                    max_vif_val <- 1
                }
                
                if (max_vif_val < 10) {
                    vif_ok <- TRUE
                }
            }
        }
    }
    cat(sprintf(
        "  th=%.3f | clusters=%3d | medoids=%3d | after_vif=%3d | max_vif=%.1f | passed=%s\n",
        th, length(unique(clusters)), ncol(X_temp), n_after, max_vif_val, vif_ok
    ))
    
    # Also print how many were removed by VIF
    removed_by_vif <- ncol(X_temp) - n_after
    cat("  Removed by VIF:", removed_by_vif, "\n")
    
    # Record result
    results <- rbind(results, data.frame(
        cor_threshold = th,
        n_clusters    = length(unique(clusters)),
        n_medoids     = n_medoids,
        n_after_vif   = n_after,
        max_vif       = round(max_vif_val, 2),
        passed_vif    = vif_ok
    ))
    
    # Update best if better
    if (vif_ok && n_after > best_n_features) {
        best_n_features <- n_after
        best_threshold  <- th
        best_X_reduced  <- X_after_vif
        message("  → New best: ", n_after, " features kept (max VIF = ", round(max_vif_val, 2), ")")
    }
}

# Print summary table
print("Summary of all tested thresholds:")
print(results %>% arrange(desc(n_after_vif)))

cat("\nBest cor_threshold:", best_threshold, 
    "→", best_n_features, "features retained after VIF < 10\n")

# ============================================================================
# === Proceed with Random Forest and Plotting on Best Reduced Set ============
# ============================================================================

if (!is.null(best_X_reduced) && ncol(best_X_reduced) > 0) {
  X_reduced <- best_X_reduced
  
  # --- Random Forest on optimal reduced features ---
  rf_model <- randomForest(
    x = X_reduced,
    y = y,
    importance = TRUE,
    ntree = 1000,
    importance.type = 1  # MeanDecreaseAccuracy
  )
  
  importance_df <- as.data.frame(importance(rf_model, type = 1))
  importance_df$Feature <- rownames(importance_df)
  importance_df <- importance_df %>%
    arrange(desc(MeanDecreaseAccuracy)) %>%
    mutate(Rank = row_number())
  
  # --- Build title + filename using best_threshold ---
  comparison_str <- paste(comparison_levels, collapse = "_vs_")
  feature_str    <- if (feature_keyword != "") paste0("_", feature_keyword) else ""
  threshold_str  <- paste0("_opt_cor", gsub("\\.", "", sprintf("%.3f", best_threshold)))  # e.g. _opt_cor0625
  
  plot_title <- paste0(
    "Top ", topN, " Features (", target_var, ": ", comparison_str, ")", feature_str
  )
  
  file_name <- paste0(
    "TopFeatures_",
    target_var, "_",
    comparison_str,
    feature_str,
    threshold_str,
    ".pdf"
  )
  file_name <- gsub("\\s+", "_", file_name)  # clean any spaces
  
  # --- Plot ---
  top_features_plot <- ggplot(importance_df[1:topN, ], aes(
    x = reorder(Feature, MeanDecreaseAccuracy),
    y = MeanDecreaseAccuracy
  )) +
    geom_bar(stat = "identity", fill = "steelblue") +   #steelblue seagreen2 maroon
    coord_flip() +
    theme_bw(base_size = 18) +
    labs(
      title = plot_title,
      subtitle = paste("After correlation clustering (threshold", best_threshold, ") + VIF <", vif_limit, "filtering"),
      x = "Feature",
      y = "Importance (Mean Decrease in Accuracy)"
    ) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.05)))
  
  # --- Save ---
  ggsave(
    filename = file_name,
    plot = top_features_plot,
    width = 10,
    height = 7,
    dpi = 300
  )
  
  cat("✅ Analysis complete. Plot saved as:", file_name, "\n")
  cat("Top 5 features:\n")
  print(head(importance_df, 10))
  
} else {
  stop("No configuration found with VIF < 10 and >0 features.")
}

````

# plot top feature from Feature Importance analysis

````{r}


# --- 1️⃣ Prepare Data ---
top_features <- importance_df$Feature[1:topN]

# Extract features and metadata
features <- as.data.frame(t(assay(sce, assay_name)))
meta <- as.data.frame(colData(sce)[, c("Patient", "Disease", "Treatment", "Plate")])
df <- cbind(meta, features)

# Filter for control and IL1 treatments
df <- df %>% filter(Treatment %in% c("control", "IL1"))

# Convert to long format
df_long <- df %>%
  select(Patient, Disease, Treatment, all_of(top_features)) %>%
  pivot_longer(
    cols = all_of(top_features),
    names_to = "Feature",
    values_to = "Value"
  )

# Summarize to patient level
df_patient_mean <- df_long %>%
  group_by(Patient, Disease, Treatment, Feature) %>%
  summarise(MeanValue = mean(Value, na.rm = TRUE), .groups = "drop")

# Group mean for bars
df_group_mean <- df_patient_mean %>%
  group_by(Disease, Treatment, Feature) %>%
  summarise(GroupMean = mean(MeanValue, na.rm = TRUE), .groups = "drop")

# --- 2️⃣ Create folder for saving ---
dir.create("TopFeatures_TreatmentFacet_IL1", showWarnings = FALSE)

# --- 3️⃣ Loop through each feature and save individual plot ---
for (f in top_features) {
  
  df_f_patient <- df_patient_mean %>% filter(Feature == f)
  df_f_group <- df_group_mean %>% filter(Feature == f)
  
  p_f <- ggplot() +
    geom_bar(
      data = df_f_group,
      aes(x = Disease, y = GroupMean, fill = Disease),
      stat = "identity",
      alpha = 0.4,
      width = 0.6,
      color = "black"
    ) +
    geom_jitter(
      data = df_f_patient,
      aes(x = Disease, y = MeanValue, color = Disease),
      width = 0.15,
      size = 3,
      alpha = 0.9
    ) +
    facet_wrap(~ Treatment, ncol = 2, scales = "fixed") +
    theme_bw(base_size = 18) +
    labs(
      title = paste0(f, " across Diseases (faceted by Treatment)"),
      x = "Disease",
      y = "Mean Feature Value per Patient"
    ) +
    scale_fill_brewer(palette = "Set2") +
    scale_color_brewer(palette = "Set2") +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      axis.title = element_text(face = "bold", size = 16),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      strip.text = element_text(size = 14, face = "bold"),
      legend.position = "none"
    )
  
  # --- 4️⃣ Save TIFF ---
  ggsave(
    filename = paste0("TopFeatures_TreatmentFacet_IL1/", f, "_Disease_TreatmentFacet.tiff"),
    plot = p_f,
    width = 7,
    height = 5,
    dpi = 600,
    units = "in"
  )
  
  message("✅ Saved: ", f)
}

cat("\n✅ All individual TIFF plots saved in 'TopFeatures_TreatmentFacet_IL1' folder.\n")


````

#correlation of metabolism for disease

````{r}
# df from previous step

df_plot <- df %>%
  select(Patient, Disease, Treatment,
         Intensity_MeanIntensity_CorrLipid,
         Intensity_MeanIntensity_CorrMitosox)

custom_colors <- c(
  "Healthy" = "#1b9e77",
  "OA" = "#d95f02",
  "RA" = "#7570b3"
)

# Plot correlation (Spearman) faceted by Disease
p_corr <- ggpubr::ggscatter(
  df_plot,
  x = "Intensity_MeanIntensity_CorrLipid",
  y = "Intensity_MeanIntensity_CorrMitosox",
  color = "Disease",                 # color by Disease
  palette = custom_colors,
  add = "reg.line",
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.method = "spearman",
  add.params = list(color = "gray30", fill = "lightgray"),
  cor.coef.size = 6,   # Font size of correlation text
  cor.coeff.args = list(
    label.x.npc = "center",
    label.y.npc = "top",
    color = "black",
    fontface = "bold"
  ),
  xlab = "Mean Intensity (LipidTOX)",
  ylab = "Mean Intensity (MitoSOX)",
  title = "Correlation between LipidTOX and MitoSOX intensities per Disease"
) +
  facet_wrap(~Disease, scales = "free") +  # facet by Disease
  theme_bw(base_size = 16) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    legend.position = "right",
    strip.text = element_text(face = "bold", size = 14)
  )

# Display plot
p_corr

````

#correlation of metabolism for Treatment

````{r}
# df from previous step

library(ggpubr)
library(ggplot2)
library(dplyr)

# --- 1️⃣ Filter data for desired treatments ---
comparison_levels <- c("control", "IL1")

df_plot <- df %>%
    filter(Treatment %in% comparison_levels) %>%
    select(Patient, Disease, Treatment,
           Intensity_MeanIntensity_CorrEdU,
           Intensity_MaxIntensity_CorrMitosox)

# --- 2️⃣ Custom colors ---
custom_colors <- c(
    "Healthy" = "#1b9e77",
    "OA" = "#d95f02",
    "RA" = "#7570b3"
)

# --- 3️⃣ Plot ---
p_corr <- ggpubr::ggscatter(
    df_plot,
    x = "Intensity_MeanIntensity_CorrEdU",
    y = "Intensity_MaxIntensity_CorrMitosox",
    color = "Disease",
    palette = custom_colors,
    add = "reg.line",
    conf.int = TRUE,
    cor.coef = TRUE,
    cor.method = "spearman",
    add.params = list(color = "gray30", fill = "lightgray"),
    cor.coef.size = 5,
    cor.coeff.args = list(
    label.x.npc = "center",   # horizontally centered
    label.y.npc = "center",   # vertically centered (safe value)
    label.sep = "\n",         # ✅ separate rho and p-value into two lines
    color = "black",
    fontface = "bold",
    size = 6
  ),
    xlab = "Mean Intensity (EdU)",
    ylab = "Max Intensity (MitoSOX)",
    title = "Correlation between EdU and MitoSOX Intensities"
) +
    facet_wrap(Treatment ~ Disease, scales = "free") +
    theme_bw(base_size = 18) +
    theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
        axis.title = element_text(face = "bold", size = 16),
        axis.text = element_text(size = 14),
        legend.title = element_text(size = 14, face = "bold"),
        legend.text = element_text(size = 13),
        strip.text = element_text(face = "bold", size = 15),
        plot.margin = unit(c(10, 20, 20, 20), "pt"),     # ✅ added "pt"
        panel.spacing = unit(1.2, "lines")
    ) +
    coord_cartesian(clip = "off")  # ✅ ensures correlation text isn't cut

# --- 4️⃣ Save high-res image ---
ggsave(
    filename = "Corr_EdU_MitoSOX_byTreatment.png",
    plot = p_corr,
    width = 11,
    height = 6,
    dpi = 600,
    bg = "white",
    limitsize = FALSE
)

## correlation from Importance feature analysis

library(reshape2)

top_features <- importance_df %>%
  slice_max(order_by = MeanDecreaseAccuracy, n = 16) %>%
  pull(Feature)

# --- 2️⃣ Optionally filter only CorrMitosox / CorrLipid if needed ---
top_features <- grep("CorrMitosox|CorrLipid", top_features, value = TRUE, invert = FALSE) #|CorrEdU
# set invert = TRUE if you want to exclude them instead

# --- 2️⃣ Subset the data ---
df_corr <- X_reduced[, top_features, drop = FALSE]

# --- 3️⃣ Compute Pearson correlation ---
cor_mat <- cor(df_corr, use = "pairwise.complete.obs", method = "pearson")

# --- 4️⃣ Melt for ggplot ---
# Mask upper triangle
cor_mat[upper.tri(cor_mat)] <- NA
corr_melt <- melt(cor_mat, na.rm = TRUE)
colnames(corr_melt) <- c("Feature1", "Feature2", "Correlation")

# --- 5️⃣ Plot heatmap ---
heatmap_plot <- ggplot(corr_melt, aes(x = Feature1, y = Feature2, fill = Correlation)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0,
                       limits = c(-1,1), name = "Pearson r") +
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 4) +
  theme_minimal(base_size = 16) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title = element_blank())

# --- 6️⃣ Save high-resolution plot ---

ggsave("Correlation_Heatmap_CorrFeatures.png", heatmap_plot, width = 10, height = 8, dpi = 300)

heatmap_plot

````

# single cell plot / per patient for feature - violin

````{r}
  "Correlation_Correlation_CorrLipid_CorrMitosox",
  "Correlation_K_CorrMitosox_CorrLipid"
 "Intensity_MeanIntensity_CorrBlue", 
 "Intensity_IntegratedIntensity_CorrEdU",
    "Intensity_IntegratedIntensityEdge_CorrEdU", 
    "Intensity_IntegratedIntensityEdge_CorrLipid"
     "Intensity_MeanIntensity_CorrEdU", 
   "Intensity_MeanIntensity_CorrLipid", 
   "Intensity_MeanIntensity_CorrMitosox" 
     "RadialDistribution_RadialCV_CorrBlue_4of4",
    "Correlation_Costes_CorrBlue_CorrEdU",
    "Intensity_StdIntensityEdge_CorrLipid"
## ===============================
## 1. Features and data prep
## ===============================
my_features  <- c("Intensity_IntegratedIntensity_CorrLipid",
                  "Intensity_IntegratedIntensity_CorrMitosox",
                  "Intensity_IntegratedIntensity_CorrEdU")


all_features <- my_features

feature_data <- as.data.frame(t(assay(sce_aggr, "tfmfeatures")))
feature_data <- feature_data[, all_features, drop = FALSE]

metadata <- as.data.frame(colData(sce_aggr)[,
  c("Patient", "Treatment", "Disease", "Joint", "Plate")
])

df <- cbind(metadata, feature_data)

df_long <- df %>%
  pivot_longer(
    cols      = all_of(all_features),
    names_to  = "Feature",
    values_to = "Intensity"
  )

## ===============================
## 2. Global switches
## ===============================
compare_mode      <- "disease"   # "treatment" or "disease" (lowercase)
diseases_to_plot  <- c("Healthy","OA", "RA")   #, "OA", "RA"
treatments_to_plot<- c("control")   #, "IL1"
plot_per_patient  <- TRUE

## ===============================
## 3. Data preparation
## ===============================
df_plot_ready <- df_long

x_var     <- NULL
facet_var <- NULL
levels_x  <- NULL

if (compare_mode == "treatment") {
  # compare control vs IL1 within diseases
  df_plot_ready <- df_plot_ready %>%
    filter(Disease %in% diseases_to_plot,
           Treatment %in% treatments_to_plot)

  x_var     <- "Treatment"
  facet_var <- "Disease"
  levels_x  <- treatments_to_plot

} else if (compare_mode == "disease") {
  # compare Healthy vs OA vs RA within treatments
  df_plot_ready <- df_plot_ready %>%
    filter(Treatment %in% treatments_to_plot,
           Disease %in% diseases_to_plot)

  x_var     <- "Disease"
  facet_var <- "Treatment"
  levels_x  <- diseases_to_plot
} else {
  stop("compare_mode must be 'treatment' or 'disease'")
}

# Aggregate per patient if needed
if (plot_per_patient) {
  df_plot_ready <- df_plot_ready %>%
    group_by(Patient, .data[[x_var]], .data[[facet_var]], Feature) %>%
    summarise(
      Intensity = mean(Intensity, na.rm = TRUE),
      .groups   = "drop"
    )
}

# Factor order for x-axis
df_plot_ready[[x_var]] <- factor(df_plot_ready[[x_var]], levels = levels_x)

## ===============================
## 4. Colors
## ===============================
    set2 <- brewer.pal(8, "Set2")
    set3 <- brewer.pal(12, "Set3")

# plot_colors <- c(
#   "control" = set2[6],    # your requested Set2 color 6
#   "IL1"     = set2[5]     # for PIC  "PIC"     = set3[5]
# )

# for disease compare
plot_colors <- c(
  "Healthy" = set3[3],
  "OA"      = set3[4],
  "RA"      = set3[5]
)

## ===============================
## 5. Loop over features and plot
## ===============================
outdir <- "Flexible_Plots"
dir.create(outdir, showWarnings = FALSE)

for (f in unique(df_plot_ready$Feature)) {

  df_plot <- df_plot_ready %>% filter(Feature == f)

  ## ---------- statistics ----------
  if (compare_mode == "treatment") {
    stat_result <- df_plot %>%
      group_by(Disease) %>%
      pairwise_wilcox_test(
        Intensity ~ Treatment,
        p.adjust.method = "none"
      ) %>%
      ungroup()
    x_for_add <- "Treatment"
  } else {  # compare_mode == "disease"
    stat_result <- df_plot %>%
      group_by(Treatment) %>%
      pairwise_wilcox_test(
        Intensity ~ Disease,
        p.adjust.method = "none"
      ) %>%
      ungroup()
    x_for_add <- "Disease"
  }

  stat.test <- NULL
  if (nrow(stat_result) > 0) {
    stat.test <- stat_result %>%
      add_xy_position(
        fun  = "max",
        x    = x_for_add,
        step.increase = 0.15
      ) %>%
      mutate(
        p.signif = case_when(
          p <= 0.001 ~ "***",
          p <= 0.01  ~ "**",
          p <= 0.05  ~ "*",
          TRUE       ~ paste0("p = ", signif(p, 2))
        )
      )
  }

  ## ---------- plot ----------
 p <- ggplot(df_plot, aes_string(x = x_var, y = "Intensity", fill = x_var)) +

  # violin + boxplot first
  geom_violin(
    trim = FALSE, alpha = 0.65,
    width = 0.9, color = "black", size = 0.9
  ) +
  geom_boxplot(
    width = 0.12, outlier.shape = NA,
    fill = "white", alpha = 0.4,
    color = "black", size = 0.5
  ) +

  # now add patient points on top
  {
    if (plot_per_patient) {
      geom_point(
        aes_string(color = NULL),  # ignore fill/color mapping
        size = 3,
        position = position_jitter(width = 0.18),
        color = "black",
        alpha = 0.6
      )
    } else {
      geom_jitter(
        aes_string(color = NULL),
        size = 1,
        width = 0.25,
        color = "black",
        alpha = 0.6
      )
    }
  } +

  # fill for violins only
  scale_fill_manual(values = plot_colors) +

  labs(
    title = paste("Feature:", f),
    y     = "Z-scored log intensity",
    x     = x_var
  ) +
  theme_bw(base_size = 18) +
  theme(
    plot.title  = element_text(size = 20, face = "bold", hjust = 0.5),
    axis.title  = element_text(size = 18, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
    panel.grid  = element_blank(),
    strip.text  = element_text(size = 14, face = "bold")
  ) +

  # stat test if exists
  {
    if (!is.null(stat.test) && nrow(stat.test) > 0) {
      stat_pvalue_manual(
        stat.test,
        label        = "p.signif",
        size         = 5,
        tip.length   = 0.01,
        bracket.size = 0.6
      )
    }
  } +

  facet_wrap(as.formula(paste("~", facet_var)))

print(p)
ggsave(
  file.path(outdir, paste0(gsub("[^A-Za-z0-9_]+", "_", f), ".pdf")),
  p, width = 8, height = 6, dpi = 300
)
}



````

# single cell plot / per patient for feature -> violin (develope panel)

````{r}

# correlation_features <- c(
#     "Correlation_RWC_CorrTMRM_CorrCytoID",
#     "Correlation_Correlation_CorrCytoID_CorrTMRM",
#     "RadialDistribution_RadialCV_CorrCytoID_4of4"
# )
intensity_features <- c(
    "Intensity_MeanIntensity_CorrLipid",
    "Intensity_MeanIntensity_CorrMitosox",
    "Intensity_MeanIntensity_CorrEdU"
)

all_features <- c(intensity_features)

feature_data <- as.data.frame(t(assay(sce_aggr, "tfmfeatures")))
feature_data <- feature_data[, all_features, drop = FALSE]

metadata <- as.data.frame(colData(sce_aggr)[, 
                   c("Patient", "Treatment", "Disease", "Joint", "Plate")])

df <- cbind(metadata, feature_data)

df_long <- df %>%
    pivot_longer(cols = all_of(all_features),
                 names_to = "Feature",
                 values_to = "Intensity")

### -------------------------
### 3. Global switches
### -------------------------

plot_per_patient <- FALSE

# 👇 EDIT HERE WHICH TREATMENTS TO COMPARE
treatments_to_plot <- c("cont", "MG")   # or c("control", "IL1") R+C

df_plot_ready <- df_long %>%
    filter(Disease %in% c("healthy")) %>%   #, "RA"
    filter(Treatment %in% treatments_to_plot)

if (plot_per_patient) {
    df_plot_ready <- df_plot_ready %>%
        group_by(Patient, Disease, Treatment, Feature) %>%
        summarise(Intensity = mean(Intensity, na.rm = TRUE), .groups = "drop")
}

### ------------------------------------------------------------
### 4. FIXED + SAFE Treatment factor ordering (CRITICAL PART)
### ------------------------------------------------------------

df_plot_ready$Treatment <- factor(df_plot_ready$Treatment,
                                  levels = treatments_to_plot)

### ------------------------------------------------------------
### 5. FLEXIBLE COLOR SYSTEM — ALWAYS CORRECT
### ------------------------------------------------------------

pal_set3 <- brewer.pal(12, "Set3")
pal_set2 <- brewer.pal(8, "Set2")

treatments <- levels(df_plot_ready$Treatment)

# Default colors for ALL treatments → Set2
treatment_colors <- setNames(pal_set3[seq_along(treatments)], treatments)

# Override rules (ONLY if present)
if ("R+C" %in% treatments) treatment_colors["R+C"] <- pal_set2[3]
if ("MG"  %in% treatments) treatment_colors["MG"]  <- pal_set2[2]

### Now colors are correct ALWAYS.

### -------------------------
### 6. LOOP OVER FEATURES
### -------------------------

outdir <- "Development panel"
dir.create(outdir, showWarnings = FALSE)

for (f in unique(df_plot_ready$Feature)) {
    
    df_plot <- df_plot_ready %>% filter(Feature == f)
    
    ### Statistics per disease
    stat.test <- df_plot %>%
        group_by(Disease) %>%
        pairwise_wilcox_test(Intensity ~ Treatment,
                             ref.group = treatments_to_plot[1]) %>%
        mutate(
            p.signif = case_when(
                p <= 0.001 ~ "***",
                p <= 0.01  ~ "**",
                p <= 0.05  ~ "*",
                TRUE ~ paste0("p=", signif(p, 2))
            ),
            y.position = max(df_plot$Intensity, na.rm = TRUE) +
                (1:n()) * (0.07 * diff(range(df_plot$Intensity, na.rm = TRUE)))
        )
    
    ### -----------------------------
    ### 7. PLOT
    ### -----------------------------
    
    p <- ggplot(df_plot, aes(x = Treatment, y = Intensity, fill = Treatment)) +
        
        ## raw points behind violin
        {
            if (plot_per_patient) {
                geom_point(aes(color = Treatment),
                           size = 3, alpha = 0.45,
                           position = position_jitter(width = 0.18))
            } else {
                geom_jitter(aes(color = Treatment),
                            size = 1, alpha = 0.35, width = 0.25)
            }
        } +
        
        ## Violin (emphasized)
        geom_violin(trim = FALSE, alpha = 0.65,
                    width = 0.9, color = "black", size = 0.9) +
        
        ## thin boxplot on top
        geom_boxplot(width = 0.12, outlier.shape = NA,
                     fill = "white", alpha = 0.4,
                     color = "black", size = 0.5) +
        
        scale_fill_manual(values = treatment_colors) +
        scale_color_manual(values = treatment_colors) +
        
        labs(
            title = paste("Feature:", f),
            y = "Z-scored log intensity",
            x = "Treatment"
        ) +
        theme_bw(base_size = 18) +
        theme(
            plot.title = element_text(size = 20, face = "bold", hjust = 0.5),
            axis.title = element_text(size = 18, face = "bold"),
            axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
            panel.grid = element_blank()
        ) +
        
        stat_pvalue_manual(stat.test, label = "p.signif",
                           size = 6, tip.length = 0.01) +
        
        facet_wrap(~ Disease)
    
    print(p)
    
    ### Save publication-quality PNG
    fname <- gsub("[^A-Za-z0-9_]+", "_", f)
    ggsave(file.path(outdir, paste0(fname, ".png")),
           p, width = 6, height = 6, dpi = 300)
}


````
#for develpoment panel -> EdU

````{r}
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # or use the assay name you used earlier
metadata <- as.data.frame(colData(sce)[, c("Plate", "Well", "Patient", "Treatment", "Disease", "Field")])

df <- cbind(metadata, intensity_data)

# Subset EdU data
df_edu <- df %>%
  filter(Treatment %in% c("cont", "MG")) %>%
  select(Treatment, Intensity_MeanIntensity_CorrEdU)

# Plot histogram
ggplot(df, aes(x = Intensity_MeanIntensity_CorrEdU)) +
    geom_histogram(
        bins = 50,
        fill = "steelblue",
        color = "black",
        alpha = 0.6
    ) +
    geom_vline(
        xintercept = 4, 
        color = "red", 
        linetype = "dashed", 
        linewidth = 1
    ) +
    theme_bw(base_size = 14) +
    labs(
        title = "Histogram of Intensity_MeanIntensity_CorrEdU",
        x = "Intensity_MeanIntensity_CorrEdU (a.u.)",
        y = "Cell count"
    )

group_level <- "Well"     # options: "Well" or "Patient"

cutoff <- 4

edu_summary <- df %>%
  filter(Disease == "healthy",
         Treatment %in% c("cont", "MG")) %>%
  mutate(Positive = Intensity_MeanIntensity_CorrEdU > cutoff) %>%
  group_by(.data[[group_level]], Treatment) %>%
  summarise(
    Total = n(),
    Positive = sum(Positive),
    Percent_Positive = 100 * Positive / Total,
    .groups = "drop"
  )

summary(edu_summary$Percent_Positive)


palette_set2 <- brewer.pal(8, "Set2")
treatment_colors <- c("cont" = palette_set2[1], "MG" = palette_set2[2])

p <- ggplot(
  edu_summary,
  aes(x = Treatment, y = Percent_Positive, fill = Treatment)
) +

  # Mean bar
  stat_summary(
    fun = mean,
    geom = "col",
    width = 0.6,
    color = "black",
    alpha = 0.85
  ) +

  # Individual wells or patients
  geom_jitter(
    width = 0.15,
    size = 3,
    alpha = 0.9
  ) +

  # Statistical comparison vs control
  stat_compare_means(
    method = "wilcox.test",
    comparisons = list(c("cont", "MG")),
    label = "p.signif",
    hide.ns = FALSE
  ) +

  scale_fill_manual(values = treatment_colors) +

  labs(
    title = paste0(
      "EdU-positive Cells (Healthy only, per ", group_level,
      ", cutoff = ", cutoff, ")"
    ),
    x = "Treatment",
    y = "% EdU-positive Cells"
  ) +

  theme_bw(base_size = 16) +
  theme(
    panel.grid.major = element_blank(),   # ✅ remove grid
    panel.grid.minor = element_blank(),   # ✅ remove grid
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "none"
  )

print(p)

outname <- paste0(
  "EdU_percent_positive_healthy_per_", tolower(group_level)
)

# PNG
ggsave(
  filename = paste0(outname, ".png"),
  plot = p,
  width = 6,
  height = 5,
  dpi = 300
)

# PDF (vector, journal-ready)
ggsave(
  filename = paste0(outname, ".pdf"),
  plot = p,
  width = 6,
  height = 5,
  device = cairo_pdf
)

````

# correlation plot for lipid

````{r}
library(RColorBrewer)

intensity_features <- c(
    "Intensity_MeanIntensity_CorrLipid",
    "Intensity_MeanIntensity_CorrMitosox",
    "Intensity_MeanIntensity_CorrEdU"
)

all_features <- c(intensity_features)

feature_data <- as.data.frame(t(assay(sce_aggr, "tfmfeatures")))
feature_data <- feature_data[, all_features, drop = FALSE]

metadata <- as.data.frame(colData(sce_aggr)[, 
                   c("Patient", "Treatment", "Disease", "Joint", "Plate")])

df <- cbind(metadata, feature_data)

df_corr <- df %>%
  #filter(Disease == "Healthy") %>%
  filter(Treatment %in% c("control", "IL1")) %>%  # <--- keep only these two groups
  select(Treatment,Disease, Intensity_MeanIntensity_CorrLipid, Intensity_MeanIntensity_CorrMitosox)

set2_colors <- RColorBrewer::brewer.pal(8, "Set2")

custom_colors <- setNames(
  set2_colors[c(6, 5)],
  c("control", "IL1")
)


# --- Scatter plot with correlation test ---
p_corr <- ggpubr::ggscatter(
  df_corr,
  x = "Intensity_MeanIntensity_CorrLipid",
  y = "Intensity_MeanIntensity_CorrMitosox",
  color = "Treatment",
  palette = custom_colors,
  add = "reg.line",
  conf.int = TRUE,
  cor.coef = TRUE,
  cor.method = "spearman",
  add.params = list(color = "gray30", fill = "lightgray"),
  cor.coef.size = 6,   # Increase font size of correlation text
  cor.coeff.args = list(
    label.x.npc = "center",  # Horizontal position ("left", "center", or "right")
    label.y.npc = "top",  # Vertical position ("bottom", "center", or "top")
    color = "black",         # Optional: set color of correlation label
    fontface = "bold"        # Optional: make it bold
  ),
  xlab = "Mean Intensity (LipidTOX)",
  ylab = "Mean Intensity (Mitosox)",
  title = "Correlation between LipidTOX and Mitosox intensities"
) +
  facet_wrap(Disease ~ Treatment) +
  theme_bw(base_size = 16) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 16),
    legend.position = "right"
  )

print(p_corr)

ggsave(
  filename = "Correlation_Lipid_Mitosox_IL1.pdf",
  plot = p_corr,
  width = 10,
  height = 8,
  units = "in"
)


###### split EdU poppulation
# --- PARAMETERS ---
cutoff <- 3   # adjust based on your histogram
palette_set2 <- brewer.pal(8, "Set2")
treatment_colors <- c("cont" = palette_set2[1], "MG" = palette_set2[2])

# --- 1. Add EdU status ---
df_corr <- df %>%
  filter(Treatment %in% c("cont", "MG")) %>%
  mutate(
    EdU_status = ifelse(Intensity_MeanIntensity_CorrEdU > cutoff, "EdU+", "EdU-")
  )

# --- 2. Loop through each EdU status and plot correlation ---
for (edu_group in c("EdU+", "EdU-")) {
  
  df_sub <- df_corr %>%
    filter(EdU_status == edu_group)
  
  # --- 3. Plot correlation between Mitosox and Lipid ---
  p_corr <- ggscatter(
    df_sub,
    x = "Intensity_MeanIntensity_CorrLipid",
    y = "Intensity_MeanIntensity_CorrMitosox",
    color = "Treatment",
    palette = treatment_colors,
    add = "reg.line",
    conf.int = TRUE,
    cor.coef = TRUE,
    cor.method = "spearman",
    facet.by = "Treatment",      # show cont vs MG separately
    title = paste0("Correlation (", edu_group, " cells)"),
    xlab = "Mean Intensity (Lipid)",
    ylab = "Mean Intensity (Mitosox)"
  ) +
    theme_bw(base_size = 16) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      axis.title = element_text(size = 15, face = "bold"),
      axis.text = element_text(size = 13),
      legend.position = "right"
    )
  
  print(p_corr)
}


````
#explore each markers in details (only Intensity data)

````{r}

feature_names <- rownames(sce)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU", 
                                    "Intensity_IntegratedIntensity_CorrMitosox",
                                    "Intensity_IntegratedIntensity_CorrLipid",
                                    "Intensity_IntegratedIntensity_CorrEdU" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Joint", "Plate")])
df <- cbind(metadata, intensity_data)  # Replace with correct column names

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Calculate mean per patient per condition
df_summary <- df_long %>%
  group_by(Patient, Treatment, Disease, Plate, Feature) %>%
  summarise(MeanIntensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


features <- unique(df_summary$Feature)

# Loop over each feature
for (f in features) {
  df_plot <- df_summary %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = MeanIntensity, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    geom_text_repel(
    aes(label = Patient, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE,
    max.overlaps = 30
    ) +
    facet_wrap(~ Disease, scales = "fixed") +
    coord_cartesian(ylim = c(-2, 2)) +
    theme_bw(base_size = 16) +
    labs(
      title = paste("Z-scored log-transformed feature intensities per Treatment_", f),
      y = "Z-scored log intensity", #Mean Intensity (a.u.)
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
   print(p) 
}

# --- Mean & SD summary ---
df_stats <- df_summary %>%
  group_by(Disease, Feature, Treatment) %>%
  summarise(
    Mean = mean(MeanIntensity, na.rm = TRUE),
    SD = sd(MeanIntensity, na.rm = TRUE),
    N = n(),
    .groups = "drop"
  )

# --- Pairwise Wilcoxon test (vs. control) ---
# Set this flag: TRUE = use Bonferroni-adjusted p-values, FALSE = use raw p-values
use_adjusted <- FALSE

stat_results <- df_summary %>%
  group_by(Disease, Feature) %>%
  pairwise_wilcox_test(
    MeanIntensity ~ Treatment,
    p.adjust.method = "bonferroni",
    ref.group = "control"
  ) %>%
  ungroup() %>%
  mutate(
    # Choose which p-value to use for significance
    p_for_sig = if (use_adjusted) p.adj else p,
    Significance = case_when(
      p_for_sig <= 0.001 ~ "***",
      p_for_sig <= 0.01  ~ "**",
      p_for_sig <= 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  filter(group1 == "control")

# --- Correct merge: match group2 to Treatment ---
final_table <- stat_results %>%
  left_join(df_stats, 
            by = c("Disease", "Feature", "group2" = "Treatment")) %>%
  select(
    Disease, Feature, group1, group2,
    Mean, SD, N, p, p.adj, Significance
  ) %>%
  arrange(Disease, Feature, group2)

# --- Export ---
write.csv(final_table, "FeatureStats_vsControl.csv", row.names = FALSE)

# Preview result
head(final_table)

```
## arrange result into table

````{r}
library(flextable)
library(officer)
library(webshot2)

library(dplyr)
library(tidyr)
library(flextable)
library(officer)

# Set this flag: TRUE = show p-adj, FALSE = show only raw p
show_adj <- FALSE

table_wide <- final_table %>%
  filter(grepl("MeanIntensity", Feature)) %>%
  select(Disease, Feature, group2, Mean, SD, p, p.adj, Significance) %>%
  mutate(
    Summary = if (show_adj) {
      paste0(
        "mean = ", round(Mean, 3), ", SD = ", round(SD, 3), "\n",
        "p = ", ifelse(is.na(p), "-", round(p, 3)),
        ", p-adj = ", ifelse(is.na(p.adj), "-", round(p.adj, 3)),
        ifelse(!is.na(Significance), paste0(", ", Significance), "")
      )
    } else {
      paste0(
        "mean = ", round(Mean, 3), ", SD = ", round(SD, 3), "\n",
        "p = ", ifelse(is.na(p), "-", round(p, 3)),
        ifelse(!is.na(Significance), paste0(", ", Significance), "")
      )
    }
  ) %>%
  select(Disease, Feature, group2, Summary) %>%
  pivot_wider(
    names_from = group2,
    values_from = Summary
  ) %>%
  arrange(factor(Disease, levels = c("Healthy", "OA", "RA")), Feature)

# --- 2. Create the flextable ---
ft <- flextable(table_wide) %>%
  set_header_labels(
    Feature = "Features",
    IFN = "IFN",
    IL1 = "IL1",
    LPS = "LPS",
    PIC = "PIC",
    TNF = "TNF"
  ) %>%
  add_header_row(
    values = c(" ", "Treatment (mean, SD, pvalue, significance)"),
    colwidths = c(1, ncol(table_wide) - 1)
  ) %>%
  merge_v(j = c("Disease", "Feature")) %>%
  bold(i = ~!is.na(Disease), j = "Disease") %>%
  theme_vanilla() %>%
  font(fontname = "Arial", part = "all") %>%   # ← ADD THIS
  autofit() %>%
  align(align = "center", part = "header") %>%
  align(j = c("Disease","Feature"), align = "left", part = "body") %>%
  align(j = c("IFN","IL1","LPS","PIC","TNF"), align = "center", part = "body") %>%
  line_spacing(space = 1.2, part = "all") %>%
  set_table_properties(width = 1, layout = "autofit")

# --- 3. Export to Word ---
doc <- read_docx()
doc <- body_add_flextable(doc, value = ft)
print(doc, target = "MeanIntensity_table_merged.docx")


# --- 4. Save as pdf ---
library(rmarkdown)

rmarkdown::pandoc_convert(
  input = "MeanIntensity_table_merged.docx",
  output = "MeanIntensity_table_merged.pdf"
)

````
## plot differences from control

````{r}

# Combine both plates and compute difference from control per patient
# (Assuming control treatment is named "Control" — change if needed)
intensity_data <- as.data.frame(t(assay(sce_aggr, "tfmfeatures")))
intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU" )]

metadata <- as.data.frame(colData(sce_aggr)[, c("Patient", "Treatment", "Disease", "Joint", "Plate")])
df <- cbind(metadata, intensity_data)  # Replace with correct column names

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Calculate mean per patient per condition
df_summary <- df_long %>%
  group_by(Patient, Treatment, Disease, Plate, Feature) %>%
  summarise(MeanIntensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


control_label <- "control"
#excluded_patient <- "TRA27"

# if no exclude applied
df_summary_filtered <- df_summary

# df_summary_filtered <- df_summary %>%
#   filter(Patient != excluded_patient)


df_diff <- df_summary_filtered %>%
  group_by(Patient, Disease, Feature) %>%
  mutate(
    ControlValue = mean(MeanIntensity[Treatment == control_label], na.rm = TRUE),
    DiffFromControl = MeanIntensity - ControlValue
  ) %>%
  filter(Treatment != control_label) %>%
  ungroup()

# --- Import significance info (from previous step) ---
# If you already have it in memory:
# final_table <- read_csv("FeatureStats_vsControl.csv") # optional

sig_table <- final_table %>%
  filter(Significance %in% c("*", "**", "***")) %>%
  select(Disease, Feature, group2, Significance) %>%
  rename(Treatment = group2)

# --- Loop through features ---
features <- unique(df_diff$Feature)

out_dir <- "feature_plots"
dir.create(out_dir, showWarnings = FALSE)

for (f in features) {

  df_plot <- df_diff %>% filter(Feature == f)
  sig_info <- sig_table %>% filter(Feature == f)

  y_pos <- df_plot %>%
    group_by(Disease, Treatment) %>%
    summarise(max_y = max(DiffFromControl, na.rm = TRUE), .groups = "drop") %>%
    left_join(sig_info, by = c("Disease", "Treatment")) %>%
    mutate(y_pos = max_y + 0.2) %>%
    filter(!is.na(Significance))

  p <- ggplot(df_plot, aes(x = Treatment, y = DiffFromControl, fill = Treatment)) +
    
    stat_summary(fun = mean, geom = "bar",
                 color = "black", alpha = 0.4, width = 0.6) +
    
    stat_summary(fun.data = mean_se, geom = "errorbar",
                 width = 0.3, color = "black") +
    
    geom_point(
      aes(color = Treatment),
      position = position_jitter(width = 0.1),
      size = 3.5, stroke = 1, alpha = 0.9
    ) +
    
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    
    facet_wrap(~ Disease, scales = "fixed") +
    
    coord_cartesian(ylim = c(-1.5, 2)) +
    
    geom_text(
      data = y_pos,
      aes(x = Treatment, y = y_pos, label = Significance),
      color = "black", size = 8, fontface = "bold", vjust = 0
    ) +
    
    theme_bw(base_size = 20) +
    labs(
      title = paste("Relative feature intensity compared to control -", f),
      y = "Difference from control (z-scored intensity)",
      x = "Treatment"
    ) +
    
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right",
      
      # 🔹 REMOVE GRID
      panel.grid = element_blank()
    )

  # -------- Save PNG --------
  ggsave(
    filename = file.path(out_dir, paste0("Feature_", f, ".png")),
    plot = p,
    width = 10, height = 6, dpi = 300
  )

  # -------- Save PDF --------
  ggsave(
    filename = file.path(out_dir, paste0("Feature_", f, ".pdf")),
    plot = p,
    width = 10, height = 6
  )
}

````
#compare by Disease

````{r}
# 1. Extract features + metadata
intensity_data <- as.data.frame(t(assay(sce, "features")))

# Select only the features you want
selected_features <- c(
                                    "Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU", 
                                    "Intensity_IntegratedIntensity_CorrMitosox",
                                    "Intensity_IntegratedIntensity_CorrLipid",
                                    "Intensity_IntegratedIntensity_CorrEdU"
  )
intensity_data <- intensity_data[, selected_features]

metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Plate")])

# Combine metadata and intensity data
df <- cbind(metadata, intensity_data)

# Filter only control samples
df_control <- df %>% filter(Treatment == "control")

# Loop through each selected feature
for (feature in selected_features) {
  
  # Summarize mean per patient x disease x plate
  df_summary <- df_control %>%
    group_by(Patient, Disease, Plate) %>%
    summarise(MeanIntensity = mean(.data[[feature]], na.rm = TRUE), .groups = "drop")
  
  # Plot
  p <- ggplot(df_summary, aes(x = Disease, y = MeanIntensity, fill = Disease)) +
    stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Disease), 
               position = position_jitter(width = 0.1),
               size = 3.5) +
    #facet_wrap(~ Plate) + 
    theme_bw(base_size = 14) +
    labs(
      title = paste("Mean", feature, "per Disease (Control Samples)"),
      y = "Mean Intensity (a.u.)",
      x = "Disease"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  # Print plot
  print(p)

}

```
#LipidTOX

````{r}
# Plot histogram for lipidtox
# to select cut off
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # or use the assay name you used earlier
metadata <- as.data.frame(colData(sce)[, c("Plate", "Well", "Patient", "Treatment", "Disease")])

# Combine
df <- cbind(metadata, intensity_data)

# 3. Compute 75th percentile per Plate
percentiles <- df %>%
  group_by(Plate) %>%
  summarise(P75 = quantile(Intensity_IntegratedIntensity_CorrLipid, 0.75, na.rm = TRUE)) 

print(percentiles)

# 4. Plot histogram with vertical line per plate
ggplot(df, aes(x = Intensity_IntegratedIntensity_CorrLipid)) +
  geom_histogram(fill = "#2C7BB6", alpha = 0.5, bins = 40, color = "black") +
  geom_vline(
    data = percentiles,
    aes(xintercept = P75),
    color = "red", linetype = "dashed", size = 1
  ) +
  facet_wrap(~ Plate, scales = "free_y") +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Distribution of Intensity (CorrLipid) per Plate",
    x = "Intensity (CorrLipid)",
    y = "Cell Count",
    caption = "Red dashed line = 75th percentile"
  )

#plot positive cell >75th percentile

# 3. Assign positivity per cell
df <- df %>%
  left_join(percentiles, by = "Plate") %>%
  mutate(Positive = Intensity_IntegratedIntensity_CorrLipid > P75)

# 4. Summarize percent positive per Patient x Treatment x Plate
df_percent <- df %>%
  group_by(Patient, Treatment, Plate, Disease) %>%
  summarise(PercentPositive = mean(Positive) * 100, .groups = "drop")

# 5. Plot percent positive cells, combine both plates
ggplot(df_percent, aes(x = Treatment, y = PercentPositive, fill = Treatment)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
  geom_point(aes(shape = Plate, color = Treatment),
             position = position_jitter(width = 0.15),
             size = 3.5, alpha = 0.9) +
  geom_text_repel(
    aes(label = Patient, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
  ) +
  facet_wrap(~ Disease, scales = "fixed") +
  theme_bw(base_size = 14) +
  labs(
    title = "Percent LipidTOX Positive Cells per Treatment (Cutoff = 75th percentile per Plate)",
    y = "Percent Positive Cells (%)",
    x = "Treatment"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = "gray60"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

# plot correlation with between (+) and (-) lipid 
df_filtered <- df %>%
  filter(Treatment %in% c("control", "IL1"),
         Disease %in% c("Healthy", "OA", "RA")) %>%
  mutate(LipidStatus = ifelse(Positive, "Positive", "Negative"))

# Summarize per patient
df_summary <- df_filtered %>%
  group_by(Patient, Treatment, Disease, LipidStatus) %>%
  summarise(MeanMito = mean(Intensity_MeanIntensity_CorrMitosox, na.rm = TRUE), 
            .groups = "drop")


ggplot(df_summary, aes(x = LipidStatus, y = MeanMito, color = LipidStatus)) +
  geom_boxplot(alpha = 0.3, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.2), size = 3) +
  facet_grid(Disease ~ Treatment, scales = "free_y") +
  scale_color_manual(values = c("Positive" = "#fc8d59", "Negative" = "#91bfdb")) +
  stat_compare_means(
    aes(group = LipidStatus),
    method = "wilcox.test",
    label = "p.signif",
    size = 6,
    label.y = NULL,           # use label.y.npc instead
    label.y.npc = 0.5          # 0 = bottom, 0.5 = middle, 1 = top
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "MitoSOX Intensity by Lipid Positivity",
    x = "Lipid Status",
    y = "Mean MitoSOX Intensity per Patient"
  ) +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )


```
#EdU -> for cytokine treatment

```{r}
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # or use the assay name you used earlier
metadata <- as.data.frame(colData(sce)[, c("Plate", "Well", "Patient", "Treatment", "Disease")])

df <- cbind(metadata, intensity_data)


#plot histogram
ggplot(df, aes(x = Intensity_MeanIntensity_CorrEdU)) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "black",
    alpha = 0.6
  ) +
  geom_vline(
    xintercept = 1.5, 
    color = "red", 
    linetype = "dashed", 
    linewidth = 1
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "Histogram of Intensity_MeanIntensity_CorrEdU",
    x = "Intensity_MeanIntensity_CorrEdU (a.u.)",
    y = "Cell count"
  )

cutoff <- 1.5

### ------------------------------------------------
### 2. Compute % positive per patient
### ------------------------------------------------

df_patient <- df %>% 
  mutate(Pos = Intensity_MeanIntensity_CorrEdU > cutoff) %>%
  group_by(Patient, Treatment, Disease) %>%
  summarise(
    Total = n(),
    Positive = sum(Pos),
    Percent_Positive = 100 * Positive / Total,
    .groups = "drop"
  )

### ------------------------------------------------
### 3. Compute mean ± SD (bars)
### ------------------------------------------------

df_avg <- df_patient %>%
  group_by(Treatment) %>%
  summarise(
    Mean_Percent = mean(Percent_Positive),
    SD_Percent   = sd(Percent_Positive),
    .groups = "drop"
  )

### ------------------------------------------------
### 4. Plot
### ------------------------------------------------

p <- ggplot() +
  
  # Bars (mean per disease)
  geom_col(
    data = df_avg,
    aes(x = Treatment, y = Mean_Percent, fill = Treatment),
    color = "black",
    width = 0.6,
    alpha = 0.85
  ) +
  
  # Dots (individual patients)
  geom_jitter(
    data = df_patient,
    aes(x = Treatment, y = Percent_Positive),
    width = 0.15,
    size = 3,
    alpha = 0.9
  ) +
  
  # Statistical test vs control (stars only if significant)
  stat_compare_means(
    data = df_patient,
    aes(x = Treatment, y = Percent_Positive),
    method = "wilcox.test",
    ref.group = "control",
    label = "p.signif",
    hide.ns = TRUE,
    label.y.npc = "top"
  ) +
  
  # Facet by disease
  facet_wrap(~ Disease, nrow = 1) +
  
  scale_fill_brewer(palette = "Set2") +
  
  theme_bw(base_size = 18) +
  labs(
    title = paste0("CorrEdU+ Cells (cutoff > ", cutoff, ")"),
    x = "Treatment",
    y = "Percent Positive Cells (%)"
  ) +
  theme(
    panel.grid = element_blank(),
    legend.position = "none",
    strip.text = element_text(face = "bold", size = 16)
  )

print(p)


### ------------------------------------------------
### 5. Save
### ------------------------------------------------

ggsave(
  filename = "CorrEdU_percent_positive_by_disease.pdf",
  plot = p,
  width = 7,
  height = 5,
  units = "in",
  device = cairo_pdf
)

```
## Fit Single Model

```{r}
result <- fitModel(sce,
                   target = "Disease", 
                   interest_level = "RA", reference_level = "healthy", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 20, n_threads = 8)

plotROC(result)


```

For Multiple model

```{r}

interest <- setdiff(unique(sce$Treatment), c("control"))
result_list <- lapply(interest, function(level) {
  fitModel(sce, 
           target = "Treatment", 
           interest_level = level, reference_level = "control", 
           group = "Patient", strata = c("Disease"), 
           n_folds = 10, n_threads = 8)
})

plotAUC(result_list)


```
Only selected features

````{r}

# overview of CellProfiler feature categories
#   include: Texture, Intensity, AreaShape, 
#   maybe include: RadialDistribution,  Neighbors, Location
#   exclude because many corner cases: Granularity, Correlation

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

interest <- setdiff(unique(sce_select$Treatment), c("control"))
#sce_select <- sce_select[str_detect(names(sce_select), "CorrCyto"), ]

result_list <- lapply(interest, function(level) {
  fitModel(sce_select,
                   target = "Treatment", 
                   interest_level = level, reference_level = "control", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 10, n_threads = 8)
  })

plotAUC(result_list)


```

filter only Intensity data

````{r}
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

feature_names <- rownames(sce_select)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce_select, "features")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU", 
                                    "Intensity_IntegratedIntensity_CorrMitosox",
                                    "Intensity_IntegratedIntensity_CorrLipid",
                                    "Intensity_IntegratedIntensity_CorrEdU" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce_select)[, c("Patient", "Treatment", "Disease")])  # Replace with correct column names

# Combine metadata with intensity data
df <- cbind(metadata, intensity_data)

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Aggregate intensity values by Patient and Treatment
df_aggregated <- df_long %>%
  group_by(Patient, Treatment, Feature, Disease) %>%
  summarize(Intensity = mean(Intensity, na.rm = TRUE)) %>%  # Use mean or median
  ungroup()

````

#Handle with EdU plot

````{r}
# Plot histogram for Intensity_IntegratedIntensity_CorrEdU
# to select cut off

ggplot(
    df_long[df_long$Feature == "Intensity_IntegratedIntensity_CorrEdU", ],
    aes(x = Intensity)
) +
    geom_histogram(bins = 100, fill = "steelblue", color = "black") +
    scale_x_continuous(
        limits = c(0, quantile(df_long$Intensity[df_long$Feature == "Intensity_IntegratedIntensity_CorrEdU"], 0.99)),
        expand = expansion(mult = c(0, 0.05))  # small padding on the right
    ) +
    labs(
        title = "Histogram of Intensity_IntegratedIntensity_CorrEdU",
        x = "Integrated Intensity",
        y = "Cell Count"
    ) +
    theme_minimal()

# Define the cutoff value
cutoff <- 9

df_corrEdU <- df_long %>%
  filter(Feature == "Intensity_IntegratedIntensity_CorrEdU", !is.na(Intensity))

# Create the histogram with adjusted axes
ggplot(df_corrEdU, aes(x = Intensity)) +
  geom_histogram(
    binwidth = 0.2,   # reasonable based on your spread; adjust if needed
    fill = "blue",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = cutoff,  # define this if you haven’t already
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Histogram of Intensity_IntegratedIntensity_CorrEdU",
    x = "Intensity",
    y = "Count"
  ) +
  facet_wrap(~ Disease + Treatment) +
  scale_x_continuous(
    limits = c(0, 20),  # shows main data bulk; remove or increase to show full range
    breaks = seq(0, 20, 2)
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )


#Calculate % of cells above cutoff

df_summary <- df_corrEdU %>%
  group_by(Disease, Treatment) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

# Print the result
print(df_summary)

#Plot — compare Treatments, facet by Disease

ggplot(df_summary, aes(x = Treatment, y = percent_above, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    y = "% Cells Above Cutoff",
    x = "Treatment"
  ) +
  facet_wrap(~ Disease) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

#facet by both Disease and Treatment
##Calculate per-patient percent above cutoff
df_per_patient <- df_corrEdU %>%
  group_by(Disease, Treatment, Patient) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

#Summarize for bars (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(
    mean_percent = mean(percent_above, na.rm = TRUE),
    .groups = "drop"
  )

#Plot with bar + dot (each patient) and facet grid
ggplot() +
  # Bar: average percentage per disease-treatment group
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_percent, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # Dots: individual patient values
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = percent_above),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +  # one panel per Treatment
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    x = "Disease",
    y = "% Cells Above Cutoff"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )


````

# Create the box plot

````{r}
# Create the plot with Healthy patients in blue
ggplot(df_aggregated, aes(x = Treatment, y = Intensity, fill = Treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +  # Filled boxes with transparency
  geom_jitter(aes(color = Disease), width = 0.2, alpha = 0.5, size = 2) +  # Map 'Disease' to color
  scale_color_manual(values = c("Healthy" = "blue", "RA" = "red")) +  # Set colors for 'Disease'
  facet_wrap(~Feature, scales = "free_y") +  # Facet by feature
  labs(
    title = "Comparison of Intensity Features Across Treatments",
    subtitle = "Each dot represents one patient (RA is in red)",
    x = "Treatment",
    y = "Intensity",
    fill = "Treatment",  # Legend for fill (boxplot)
    color = "Disease"    # Legend for color (dots)
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 14),  # Increase general text size
    axis.text = element_text(size = 12),  # Increase axis text size
    axis.title = element_text(size = 14),  # Increase axis title size
    plot.title = element_text(size = 16, face = "bold"),  # Increase plot title size and make it bold
    plot.subtitle = element_text(size = 14),  # Increase plot subtitle size
    strip.text = element_text(size = 14)  # Increase facet label size
  )

````
Plot patient box side-by-side

````{r}
# Create the plot with Healthy and RA patients side-by-side
ggplot(df_long, aes(x = Treatment, y = Intensity, fill = Treatment)) +
  geom_boxplot(aes(color = Disease), outlier.shape = NA, alpha = 0.8, position = position_dodge(width = 0.8)) +  # Boxplots with side-by-side grouping
  geom_jitter(aes(color = Disease), alpha = 0.5, size = 2, position = position_dodge(width = 0.8)) +  # Jittered points with side-by-side grouping (no width argument)
  scale_color_manual(values = c("Healthy" = "blue", "RA" = "red")) +  # Set colors for 'Disease'
  scale_fill_manual(values = c("Treatment1" = "lightgray", "Treatment2" = "darkgray")) +  # Set fill colors for treatments
  facet_wrap(~Feature, scales = "free_y") +  # Facet by feature
  labs(
    title = "Comparison of Intensity Features Across Treatments",
    subtitle = "Each dot represents one patient (Healthy in blue, RA in red)",
    x = "Treatment",
    y = "Intensity",
    fill = "Treatment",  # Legend for fill (boxplot)
    color = "Disease"    # Legend for color (dots)
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 14),  # Increase general text size
    axis.text = element_text(size = 12),  # Increase axis text size
    axis.title = element_text(size = 14),  # Increase axis title size
    plot.title = element_text(size = 16, face = "bold"),  # Increase plot title size and make it bold
    plot.subtitle = element_text(size = 14),  # Increase plot subtitle size
    strip.text = element_text(size = 14)  # Increase facet label size
  )

````
# Handle Lipidtox data

````{r}
# Plot histogram for Intensity_IntegratedIntensity_CorrLipid
# to select cut off

ggplot(
    df_long[df_long$Feature == "Intensity_IntegratedIntensity_CorrLipid", ],
    aes(x = Intensity)
) +
    geom_histogram(bins = 100, fill = "steelblue", color = "black") +
    scale_x_continuous(
        limits = c(0, quantile(df_long$Intensity[df_long$Feature == "Intensity_IntegratedIntensity_CorrLipid"], 0.99)),
        expand = expansion(mult = c(0, 0.05))  # small padding on the right
    ) +
    labs(
        title = "Histogram of Intensity_IntegratedIntensity_CorrLipid",
        x = "Integrated Intensity",
        y = "Cell Count"
    ) +
    theme_minimal()

# Calculate the 75th percentile (Q3)
q75 <- quantile(
  df_long$Intensity[df_long$Feature == "Intensity_IntegratedIntensity_CorrLipid"],
  probs = 0.75,
  na.rm = TRUE
)

q75

# Define the cutoff value => according from 75%
cutoff <- 969


df_corrLipid <- df_long %>%
  filter(Feature == "Intensity_IntegratedIntensity_CorrLipid", !is.na(Intensity))

# Create the histogram with adjusted axes
ggplot(df_corrLipid, aes(x = Intensity)) +
  geom_histogram(
    binwidth = 20,   # reasonable based on your spread; adjust if needed
    fill = "blue",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = cutoff,  # define this if you haven’t already
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Histogram of Intensity_IntegratedIntensity_CorrLipid",
    x = "Intensity",
    y = "Count"
  ) +
  facet_wrap(~ Disease + Treatment) +
  scale_x_continuous(
    limits = c(0, 3000),  # shows main data bulk; remove or increase to show full range
    breaks = seq(0, 20, 2)
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )


#Calculate % of cells above cutoff

df_summary <- df_corrLipid %>%
  group_by(Disease, Treatment) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

# Print the result
print(df_summary)

#Plot — compare Treatments, facet by Disease

ggplot(df_summary, aes(x = Treatment, y = percent_above, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    y = "% Cells Above Cutoff",
    x = "Treatment"
  ) +
  facet_wrap(~ Disease) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

#facet by both Disease and Treatment
##Calculate per-patient percent above cutoff
df_per_patient <- df_corrLipid %>%
  group_by(Disease, Treatment, Patient) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

#Summarize for bars (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(
    mean_percent = mean(percent_above, na.rm = TRUE),
    .groups = "drop"
  )

#Plot with bar + dot (each patient) and facet grid
ggplot() +
  # Bar: average percentage per disease-treatment group
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_percent, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # Dots: individual patient values
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = percent_above),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +  # one panel per Treatment
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    x = "Disease",
    y = "% Cells Above Cutoff"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

````
# Handle MitoSOX data

````{r}
# Plot histogram for Intensity_MeanIntensity_CorrMitosox
#or Intensity_IntegratedIntensity_CorrMitosox
# to select cut off

df_filtered <- df_long %>%
  filter(Feature == "Intensity_MeanIntensity_CorrMitosox")

#Summarize to get mean intensity per patient
df_per_patient <- df_filtered %>%
  group_by(Patient, Disease, Treatment) %>%
  summarise(mean_intensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


#Create summary data for bar plot (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(mean_intensity = mean(mean_intensity, na.rm = TRUE), .groups = "drop")

# 4. Plot
ggplot() +
  # Bars: mean across patients
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_intensity, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # One dot per patient
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = mean_intensity),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +
  labs(
    title = "Mean CorrMitosox Intensity per Patient",  #change to Mean if use
    x = "Disease",
    y = "Mean Intensity per Patient"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

````
#Filter one function at a time

```{r}
#Mitosox <- mitochondrial ROS
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrMitosox"),  # keep only those rows
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Mitochondrial ROS pathway (MitoSOX)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))



```

Lipid TOX 

```{r}
#LipidTOX <- Neural lipid
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrLipid"),  # keep only those rows
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Neutral lipid pathway (LipidTOX)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

```

EdU

```{r}
#EdU <- proliferation
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrEdU"),  # keep only those rows
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Neutral lipid pathway (LipidTOX)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

````
# Pre-validation analysis
Predict Disease

```{r}
# define new target
target = "Disease"

# set Healthy as reference level
sce_aggr$Disease <- as.factor(sce_aggr$Disease)
sce_aggr$Disease <- relevel(sce_aggr$Disease, ref = "Healthy") #Healthy

# filter to control cells only and plate
sce_aggr <- sce_aggr[, sce_aggr$Treatment == "control"]

# comparison
reference_level <- "Healthy"
interest_levels <- setdiff(levels(sce_aggr$Disease), reference_level)

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrLipid", "CorrBlue", "CorrMitosox", "CorrEdU")


# only run for "all" feature
sce_oa <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "OA", reference_level = "Healthy",
  types = NULL, channels = channels,
  n_threads = n_threads
  )

#n_tests <- length(types)+length(channels)
meta_vars = c("Patient", "Disease", "Plate")
plotLOO(sce_oa, meta_vars = meta_vars, target = target)
plotROC(sce_oa, meta_vars = meta_vars, target = target)
plotAUC(sce_oa, meta_vars = meta_vars, target = target)  #, batch = "Plate"
volcanoPlot(sce_oa, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target)

# compare RA vs Healthy
sce_ra <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "RA", reference_level = "Healthy",
  types = NULL, channels = channels,
  n_threads = n_threads
  )
plotLOO(sce_ra, meta_vars = meta_vars, target = target, )
plotROC(sce_ra, meta_vars = meta_vars, target = target)
plotAUC(sce_ra, meta_vars = meta_vars, target = target)
volcanoPlot(sce_ra, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5, 
            meta_vars = meta_vars, target = target)

# compare OA vs RA
sce_oara <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "RA", reference_level = "OA",
  types = NULL, channels = channels,
  n_threads = n_threads
  )
plotLOO(sce_oara, meta_vars = meta_vars, target = target, )

# combine OA and RA
plotAUC(list(sce_oa, sce_ra), meta_vars = meta_vars, target = target)
volcanoPlot(list(sce_oa, sce_ra), 
            p_cutoff = 0.1/(2*n_tests), 
            fc_cutoff = 0.5, 
            meta_vars = meta_vars, target = target)


```

# To save plots

```{r}
# Create folder to keep files tidy
dir.create("Publication_Plots", showWarnings = FALSE)

meta_vars <- c("Patient", "Disease", "Plate")
target <- "Disease"

# OA vs Healthy
p_loo_oa <- plotLOO(sce_oa, meta_vars = meta_vars, target = target)
p_roc_oa <- plotROC(sce_oa, meta_vars = meta_vars, target = target)
p_auc_oa <- plotAUC(sce_oa, meta_vars = meta_vars, target = target)

# RA vs Healthy
p_loo_ra <- plotLOO(sce_ra, meta_vars = meta_vars, target = target)
p_roc_ra <- plotROC(sce_ra, meta_vars = meta_vars, target = target)
p_auc_ra <- plotAUC(sce_ra, meta_vars = meta_vars, target = target)

# OA vs RA
p_loo_oara <- plotLOO(sce_oara, meta_vars = meta_vars, target = target)

theme_pub <- theme_bw(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid = element_blank()
  )

scale_y_01 <- scale_y_continuous(
  limits = c(0, 1),
  breaks = seq(0, 1, 0.2),
  expand = c(0, 0)
)

p_auc_oa <- p_auc_oa + theme_pub + scale_y_01
p_auc_ra <- p_auc_ra + theme_pub + scale_y_01

p_roc_oa <- p_roc_oa + theme_pub + scale_y_01
p_roc_ra <- p_roc_ra + theme_pub + scale_y_01

p_loo_oa <- p_loo_oa + theme_pub + scale_y_01
p_loo_ra <- p_loo_ra + theme_pub + scale_y_01

p_loo_oara <- p_loo_oara + theme_pub + scale_y_01



# Optional high-res PNGs for slides
ggsave("Publication_Plots/LOO_OA_vs_Healthy.png", p_loo_oa, width = 8, height = 6)
ggsave("Publication_Plots/ROC_OA_vs_Healthy.png", p_roc_oa, width = 8, height = 6)
ggsave("Publication_Plots/AUC_OA_vs_Healthy.png", p_auc_oa, width = 8, height = 6)

ggsave("Publication_Plots/LOO_OA_vs_Healthy.pdf", p_loo_oa, width = 8, height = 6)

#OA vs RA

ggsave("Publication_Plots/LOO_OA_vs_RA.png", p_loo_oara, width = 8, height = 6)
ggsave("Publication_Plots/ROC_OA_vs_RA.png", p_roc_oara, width = 8, height = 6)
ggsave("Publication_Plots/AUC_OA_vs_RA.png", p_auc_oara, width = 8, height = 6)

ggsave("Publication_Plots/LOO_OA_vs_RA.pdf", p_loo_oara, width = 8, height = 6)

# --- RA plots ---
ggsave("Publication_Plots/LOO_RA_vs_Healthy.png", p_loo_ra, width = 8, height = 6)
ggsave("Publication_Plots/ROC_RA_vs_Healthy.png", p_roc_ra, width = 8, height = 6)
ggsave("Publication_Plots/AUC_RA_vs_Healthy.png", p_auc_ra, width = 8, height = 6)

ggsave("Publication_Plots/LOO_RA_vs_Healthy.pdf", p_loo_ra, width = 8, height = 6)

p_auc_combined <- plotAUC(
  list(sce_oa, sce_ra),
  meta_vars = meta_vars,
  target = target
) +
  theme_bw(base_size = 20) +    # ✅ increase font size
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 22),
    axis.title = element_text(face = "bold", size = 18),
    axis.text = element_text(size = 14),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14)
  )

# --- Save publication-quality figure ---
dir.create("Publication_Plots", showWarnings = FALSE)

ggsave(
  filename = "Publication_Plots/AUC_Combined_OA_RA_vs_Healthy.tiff",
  plot = p_auc_combined,
  width = 8,
  height = 6,
  dpi = 600,   # ✅ high resolution
  units = "in"
)

```

# run LOO, plot volcano/function - save pdf

```{r}
# define new target
target = "Disease"

# set Healthy as reference level
sce_aggr$Disease <- as.factor(sce_aggr$Disease)
sce_aggr$Disease <- relevel(sce_aggr$Disease, ref = "Healthy") #Healthy

# filter to control cells only and plate
sce_aggr <- sce_aggr[, sce_aggr$Treatment == "control"]

# comparison
reference_level <- "Healthy"
interest_levels <- setdiff(levels(sce_aggr$Disease), reference_level)

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrLipid", "CorrBlue", "CorrMitosox", "CorrEdU")


# only run for "all" feature
sce_oa <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "OA", reference_level = "Healthy",
  types = NULL, channels = channels,
  n_threads = n_threads
  )

# compare RA vs Healthy
sce_ra <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "RA", reference_level = "Healthy",
  types = NULL, channels = channels,
  n_threads = n_threads
  )

# compare OA vs RA
sce_oara <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "RA", reference_level = "OA",
  types = NULL, channels = channels,
  n_threads = n_threads
  )

# Your custom theme (you already have this)
theme_pub <- theme_bw(base_size = 18) +
  theme(
    plot.title   = element_text(face = "bold", hjust = 0.5),
    axis.title   = element_text(face = "bold"),
    panel.grid   = element_blank(),
    # Optional extras for clean pub style
    axis.line    = element_line(color = "black"),
    strip.text   = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )

# Fix axis for volcano
scale_x_custom <- scale_x_continuous(
  limits = c(0, 2.5),
  breaks = seq(0, 2.5, by = 0.5),
  expand = c(0, 0),
  name   = "log2 Fold Change"   # adjust label if needed
)

scale_y_custom <- scale_y_continuous(
  limits = c(0, 8),
  breaks = seq(0, 8, by = 1),
  expand = c(0, 0),
  name   = "-log10(adjusted p-value)"
)

meta_vars <- c("Patient", "Disease", "Plate")
n_tests <- length(types)+length(channels)

# ────────────────────────────────────────────────────────────────
# Plot volcano + theme
# ────────────────────────────────────────────────────────────────

reference_level <- "Healthy"

p <- volcanoPlot(
  sce_oa,
  p_cutoff = 0.05/n_tests,
  fc_cutoff = 0.5,
  meta_vars = meta_vars,
  target = target
) + theme_pub + scale_x_custom + scale_y_custom + ggtitle("Volcano_OA_vs_Healthy")

ggsave("Volcano_OA_vs_healthy.pdf", p, width = 6, height = 5)

#--------------

reference_level <- "Healthy"

p2 <- volcanoPlot(
  sce_ra,
  p_cutoff = 0.05/n_tests,
  fc_cutoff = 0.5,
  meta_vars = meta_vars,
  target = target
) + theme_pub + scale_x_custom + scale_y_custom  + ggtitle("Volcano_RA_vs_Healthy")

ggsave("Volcano_RA_vs_healthy.pdf", p2, width = 6, height = 5)

#------------

reference_level <- "OA"

p3 <- volcanoPlot(
  sce_oara,
  p_cutoff = 0.05/n_tests,
  fc_cutoff = 0.5,
  meta_vars = meta_vars,
  target = target
) + theme_pub + scale_x_custom + scale_y_custom  + ggtitle("Volcano_OA_vs_RA")

ggsave("Volcano_RA_vs_OA.pdf", p3, width = 6, height = 5)

````
# Pre-validation analysis
Predict Treatment

```{r}
target = "Treatment"

# set Healthy as reference level
sce_aggr$Treatment <- as.factor(sce_aggr$Treatment)
sce_aggr$Treatment <- relevel(sce_aggr$Treatment, ref = "control") 


# filter to RA cells only 
sce_aggr <- sce_aggr[, sce_aggr$Disease == "RA"]

# comparison
reference_level <- "control"   
interest_levels <- setdiff(levels(sce_aggr$Treatment), reference_level)


types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", "Granularity")

# channels
channels <- c("CorrLipid", "CorrBlue", "CorrMitosox", "CorrEdU")

sce_aggr$Treatment <- factor(sce_aggr$Treatment)
sce_aggr$Treatment <- relevel(sce_aggr$Treatment, ref = "control")

res_IL1 <- predictLOO(
  sce_aggr,
  interest_level = "IL1",
  reference_level = "control",
  types = NULL,
  channels = channels,
  n_threads = n_threads
)

# downstream plotting
n_tests <- length(types)+length(channels)
meta_vars = c("Patient", "Disease", "Plate", "Treatment")
plotLOO(res_IL1, meta_vars = meta_vars, target = target)
plotROC(res_IL1, meta_vars = meta_vars, target = target)
plotAUC(res_IL1, meta_vars = meta_vars, target = target)  #, batch = "Plate"
volcanoPlot(res_IL1, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target)

#------generate plot---------------
sce_il1 <- res_IL1

meta_vars <- c("Patient", "Disease", "Plate", "Treatment")
target <- "Treatment"

p_loo_il1 <- plotLOO(sce_il1, meta_vars = meta_vars, target = target)
p_roc_il1 <- plotROC(sce_il1, meta_vars = meta_vars, target = target)
p_auc_il1 <- plotAUC(sce_il1, meta_vars = meta_vars, target = target)

theme_pub <- theme_bw(base_size = 18) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(face = "bold"),
    panel.grid = element_blank()
  )

scale_y_01 <- scale_y_continuous(
  limits = c(0, 1),
  breaks = seq(0, 1, 0.2),
  expand = c(0, 0)
)

p_loo_il1 <- p_loo_il1 + theme_pub 
p_roc_il1 <- p_roc_il1 + theme_pub 
p_auc_il1 <- p_auc_il1 + theme_pub 

dir.create("Publication_Plots", showWarnings = FALSE)

# PDF (vector)
ggsave("Publication_Plots/LOO_IL1_vs_Control.pdf", p_loo_il1, width = 8, height = 6)

# PNG (slides)
ggsave("Publication_Plots/LOO_IL1_vs_Control.png", p_loo_il1, width = 8, height = 6)
ggsave("Publication_Plots/ROC_IL1_vs_Control.png", p_roc_il1, width = 8, height = 6)
ggsave("Publication_Plots/AUC_IL1_vs_Control.png", p_auc_il1, width = 8, height = 6)


#-----Volcano plot--------------

# Your custom theme (you already have this)
theme_pub <- theme_bw(base_size = 18) +
  theme(
    plot.title   = element_text(face = "bold", hjust = 0.5),
    axis.title   = element_text(face = "bold"),
    panel.grid   = element_blank(),
    # Optional extras for clean pub style
    axis.line    = element_line(color = "black"),
    strip.text   = element_text(face = "bold"),
    legend.title = element_text(face = "bold")
  )

# Fix axis for volcano
scale_x_custom <- scale_x_continuous(
  limits = c(0, 2.5),
  breaks = seq(0, 2.5, by = 0.5),
  expand = c(0, 0),
  name   = "log2 Fold Change"   # adjust label if needed
)

scale_y_custom <- scale_y_continuous(
  limits = c(0, 8),
  breaks = seq(0, 8, by = 1),
  expand = c(0, 0),
  name   = "-log10(adjusted p-value)"
)


# ────────────────────────────────────────────────────────────────
# Plot volcano + theme
# ────────────────────────────────────────────────────────────────


p <- volcanoPlot(res_IL1, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target) + 
            theme_pub + scale_x_custom + scale_y_custom + ggtitle("Volcano_control_vs_IL1")

ggsave(
  filename = "Volcano_control_vs_IL1.pdf",
  plot = p,
  width = 6,
  height = 5,
  device = cairo_pdf
)

ggsave(
  filename = "Volcano_control_vs_IL1.png",
  plot = p,
  width = 6,
  height = 5,
  dpi = 300
)


```
## Unsupervised Analysis PCA

````{r}
sce <- runPCA(sce, exprs_values = "tfmfeatures")
plotReducedDim(sce, dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce, dimred = "PCA", colour_by = "Plate")

sce_select <- sce[str_detect(names(sce), "Intensity"), ]
sce_select <- runPCA(sce_select, exprs_values = "tfmfeatures", ncomponents = 10)
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Plate")

```

What is driving PC1?

```{r}
plotPCACor(sce_select, filter_by = 1, top = 20)
```

