---
title: "Metabolism"
author: "Phatthamon"
date: now
format: html
editor: source
embed-resources: true
---

## Load Data

```{r}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(stringr)
library(dplyr)
library(batchelor)
library(BiocParallel)
library(tidyr)
library(MultiAssayExperiment)
library(lme4)
library(lmerTest)
library(ggrepel)
n_threads <- 16

```

Path to files.

```{r}
  # path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# metabolism panel
files_met <- c(
  paste0(base, "250709 cytokines_metabolism/MyExpt_Cells-plate1.csv"), 
  paste0(base, "250709 cytokines_metabolism/MyExpt_Cells-plate2.csv"), 
  paste0(base, "250717 cytokines_metabolism lot2/MyExpt_Cells_plate1.csv"),
  paste0(base, "250717 cytokines_metabolism lot2/MyExpt_Cells_plate2.csv")
)
sce <- Reduce(cbind, lapply(files_met, loadData))

```
Path to files.

```{r}
# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

#stress panel develope plate
files_met <- c(paste0(base, "250228 lipid panel_hand/MyExpt_Cells.csv"))

sce <- Reduce(cbind, lapply(files_met, loadData))


```
## Data Transformation

Prepare the data for machine learning.

```{r}
data_transformation <- function(sce) {

  # per plate
  plates <- unique(sce[["Plate"]])

  plate_list <- lapply(plates, function(plate_id) {

    # normalize per plate
    sce_plate <- sce[, sce[["Plate"]] == plate_id]

    # remove cells with missing features
    sce_plate <- removeNAs(sce_plate)
    
    # remove outlier images
    sce_plate <- removeOutliers(sce_plate, min = 30, max = 300)
    
    # remove outlier cells
    stats <- perCellQCMetrics(sce_plate, assay.type = "features")
    sce_plate <- sce_plate[, !isOutlier(stats$sum)]
    
    # remove features with zero spread
    sce_plate <- removeLowVariance(sce_plate, robust = TRUE)
    
    # transform non-negatives features and scale all features
    sce_plate <- transformLogScale(sce_plate, robust = TRUE)
  
    # fix factors, code from ChatGPT
sce_plate$Treatment <- factor(sce_plate$Treatment, levels = unique(sce$Treatment))
sce_plate$Treatment <- relevel(sce_plate$Treatment, ref = "control")
    
    sce_plate
  
  })

  # combine plates
  feature_name_list <- lapply(plate_list,
                              function(sce_plate) rownames(sce_plate))
  include_features <- Reduce(intersect, feature_name_list)
  plate_list <- lapply(plate_list,
                       function(sce_plate) sce_plate[include_features, ])
  sce <- do.call(cbind, plate_list)
  sce$Plate <- as.factor(sce$Plate)
  sce
  
}
sce <- data_transformation(sce)

lapply(unique(sce[["Plate"]]), function(plate_id) {
  
  plotCellsPerImage(sce[, sce[["Plate"]] == plate_id])
  
})

```

Pseudo-bulk over images.

```{r}
sce_aggr <- aggregateAcrossCells(
  sce,
  id = colData(sce)[, c("ImageNumber", "Well", "Patient", "Plate")], 
  use.assay.type = "tfmfeatures", 
  statistics = "mean"
  )

```
#explore each markers in details (only Intensity data)

````{r}

feature_names <- rownames(sce)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU", 
                                    "Intensity_IntegratedIntensity_CorrMitosox",
                                    "Intensity_IntegratedIntensity_CorrLipid",
                                    "Intensity_IntegratedIntensity_CorrEdU" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Joint", "Plate")])
df <- cbind(metadata, intensity_data)  # Replace with correct column names

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Calculate mean per patient per condition
df_summary <- df_long %>%
  group_by(Patient, Treatment, Disease, Plate, Feature) %>%
  summarise(MeanIntensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


features <- unique(df_summary$Feature)

# Loop over each feature
for (f in features) {
  df_plot <- df_summary %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = MeanIntensity, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    facet_wrap(~ Disease, scales = "fixed") +
    theme_bw(base_size = 14) +
    labs(
      title = paste("Z-scored log-transformed feature intensities per Treatment_", f),
      y = "Z-scored log intensity", #Mean Intensity (a.u.)
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
   print(p) 
}

```
## plot differences from control

````{r}

# Combine both plates and compute difference from control per patient
# (Assuming control treatment is named "Control" — change if needed)
control_label <- "control"

df_diff <- df_summary %>%
  group_by(Patient, Disease, Feature) %>%
  mutate(
    ControlValue = mean(MeanIntensity[Treatment == control_label], na.rm = TRUE),
    DiffFromControl = MeanIntensity - ControlValue
  ) %>%
  filter(Treatment != control_label) %>%
  ungroup()



# Loop through each feature
features <- unique(df_diff$Feature)

for (f in features) {
  df_plot <- df_diff %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = DiffFromControl, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    geom_text_repel(
    aes(label = Patient, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
  ) +
    facet_wrap(~ Disease, scales = "fixed") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    theme_bw(base_size = 14) +
    labs(
      title = paste("Relative feature intensity compared to control -", f),
      y = "Difference from control (z-scored intensity)",
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  print(p)

}


````
#compare by Disease

````{r}
# 1. Extract features + metadata
intensity_data <- as.data.frame(t(assay(sce, "features")))

# Select only the features you want
selected_features <- c(
                                    "Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU", 
                                    "Intensity_IntegratedIntensity_CorrMitosox",
                                    "Intensity_IntegratedIntensity_CorrLipid",
                                    "Intensity_IntegratedIntensity_CorrEdU"
  )
intensity_data <- intensity_data[, selected_features]

metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Plate")])

# Combine metadata and intensity data
df <- cbind(metadata, intensity_data)

# Filter only control samples
df_control <- df %>% filter(Treatment == "control")

# Loop through each selected feature
for (feature in selected_features) {
  
  # Summarize mean per patient x disease x plate
  df_summary <- df_control %>%
    group_by(Patient, Disease, Plate) %>%
    summarise(MeanIntensity = mean(.data[[feature]], na.rm = TRUE), .groups = "drop")
  
  # Plot
  p <- ggplot(df_summary, aes(x = Disease, y = MeanIntensity, fill = Disease)) +
    stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Disease), 
               position = position_jitter(width = 0.1),
               size = 3.5) +
    #facet_wrap(~ Plate) + 
    theme_bw(base_size = 14) +
    labs(
      title = paste("Mean", feature, "per Disease (Control Samples)"),
      y = "Mean Intensity (a.u.)",
      x = "Disease"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  # Print plot
  print(p)

}

```
#LipidTOX

````{r}
# Plot histogram for lipidtox
# to select cut off
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # or use the assay name you used earlier
metadata <- as.data.frame(colData(sce)[, c("Plate", "Well", "Patient", "Treatment", "Disease")])

# Combine
df <- cbind(metadata, intensity_data)

# 3. Compute 75th percentile per Plate
percentiles <- df %>%
  group_by(Plate) %>%
  summarise(P75 = quantile(Intensity_IntegratedIntensity_CorrLipid, 0.75, na.rm = TRUE)) 

print(percentiles)

# 4. Plot histogram with vertical line per plate
ggplot(df, aes(x = Intensity_IntegratedIntensity_CorrLipid)) +
  geom_histogram(fill = "#2C7BB6", alpha = 0.5, bins = 40, color = "black") +
  geom_vline(
    data = percentiles,
    aes(xintercept = P75),
    color = "red", linetype = "dashed", size = 1
  ) +
  facet_wrap(~ Plate, scales = "free_y") +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Distribution of Intensity (CorrLipid) per Plate",
    x = "Intensity (CorrLipid)",
    y = "Cell Count",
    caption = "Red dashed line = 75th percentile"
  )

#plot positive cell >75th percentile

# 3. Assign positivity per cell
df <- df %>%
  left_join(percentiles, by = "Plate") %>%
  mutate(Positive = Intensity_IntegratedIntensity_CorrLipid > P75)

# 4. Summarize percent positive per Patient x Treatment x Plate
df_percent <- df %>%
  group_by(Patient, Treatment, Plate, Disease) %>%
  summarise(PercentPositive = mean(Positive) * 100, .groups = "drop")

# 5. Plot percent positive cells, combine both plates
ggplot(df_percent, aes(x = Treatment, y = PercentPositive, fill = Treatment)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
  geom_point(aes(shape = Plate, color = Treatment),
             position = position_jitter(width = 0.15),
             size = 3.5, alpha = 0.9) +
  geom_text_repel(
    aes(label = Patient, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
  ) +
  facet_wrap(~ Disease, scales = "fixed") +
  theme_bw(base_size = 14) +
  labs(
    title = "Percent LipidTOX Positive Cells per Treatment (Cutoff = 75th percentile per Plate)",
    y = "Percent Positive Cells (%)",
    x = "Treatment"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = "gray60"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```
#EdU

```{r}
#plot histogram
ggplot(df, aes(x = Intensity_MeanIntensity_CorrEdU)) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "black",
    alpha = 0.6
  ) +
  geom_vline(
    xintercept = 0.1, 
    color = "red", 
    linetype = "dashed", 
    linewidth = 1
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "Histogram of CorrEdU Integrated Intensity",
    x = "Intensity_IntegratedIntensity_CorrEdU (a.u.)",
    y = "Cell count"
  )

#compute percentage
cutoff <- 0.1

df_summary <- df %>%
  mutate(Pos = Intensity_MeanIntensity_CorrEdU > cutoff) %>%  # mark positive cells
  group_by(Plate, Disease, Treatment) %>%                      # adjust grouping as you need
  summarise(
    Total = n(),
    Positive = sum(Pos),
    Percent_Positive = (Positive / Total) * 100,
    .groups = "drop"
  )

print(df_summary)

#plot bar
ggplot(df_summary, aes(x = Disease, y = Percent_Positive, fill = Disease)) +
  geom_bar(stat = "identity", color = "black", width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", Percent_Positive)),
            vjust = -0.5, size = 4) +
  facet_wrap(~ Treatment) +  
  theme_bw(base_size = 14) +
  labs(
    title = "Percentage of CorrEdU+ Cells (x > 0.1)",
    x = "Disease",
    y = "Percent Positive Cells (%)"
  )


```
## Fit Single Model

```{r}
result <- fitModel(sce,
                   target = "Disease", 
                   interest_level = "RA", reference_level = "healthy", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 20, n_threads = 8)

plotROC(result)


```

For Multiple model

```{r}

interest <- setdiff(unique(sce$Treatment), c("control"))
result_list <- lapply(interest, function(level) {
  fitModel(sce, 
           target = "Treatment", 
           interest_level = level, reference_level = "control", 
           group = "Patient", strata = c("Disease"), 
           n_folds = 10, n_threads = 8)
})

plotAUC(result_list)


```
Only selected features

````{r}

# overview of CellProfiler feature categories
#   include: Texture, Intensity, AreaShape, 
#   maybe include: RadialDistribution,  Neighbors, Location
#   exclude because many corner cases: Granularity, Correlation

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

interest <- setdiff(unique(sce_select$Treatment), c("control"))
#sce_select <- sce_select[str_detect(names(sce_select), "CorrCyto"), ]

result_list <- lapply(interest, function(level) {
  fitModel(sce_select,
                   target = "Treatment", 
                   interest_level = level, reference_level = "control", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 10, n_threads = 8)
  })

plotAUC(result_list)


```

filter only Intensity data

````{r}
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

feature_names <- rownames(sce_select)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce_select, "features")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrMitosox", 
                                    "Intensity_MeanIntensity_CorrLipid", 
                                    "Intensity_MeanIntensity_CorrEdU", 
                                    "Intensity_IntegratedIntensity_CorrMitosox",
                                    "Intensity_IntegratedIntensity_CorrLipid",
                                    "Intensity_IntegratedIntensity_CorrEdU" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce_select)[, c("Patient", "Treatment", "Disease")])  # Replace with correct column names

# Combine metadata with intensity data
df <- cbind(metadata, intensity_data)

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Aggregate intensity values by Patient and Treatment
df_aggregated <- df_long %>%
  group_by(Patient, Treatment, Feature, Disease) %>%
  summarize(Intensity = mean(Intensity, na.rm = TRUE)) %>%  # Use mean or median
  ungroup()

````

#Handle with EdU plot

````{r}
# Plot histogram for Intensity_IntegratedIntensity_CorrEdU
# to select cut off

ggplot(
    df_long[df_long$Feature == "Intensity_IntegratedIntensity_CorrEdU", ],
    aes(x = Intensity)
) +
    geom_histogram(bins = 100, fill = "steelblue", color = "black") +
    scale_x_continuous(
        limits = c(0, quantile(df_long$Intensity[df_long$Feature == "Intensity_IntegratedIntensity_CorrEdU"], 0.99)),
        expand = expansion(mult = c(0, 0.05))  # small padding on the right
    ) +
    labs(
        title = "Histogram of Intensity_IntegratedIntensity_CorrEdU",
        x = "Integrated Intensity",
        y = "Cell Count"
    ) +
    theme_minimal()

# Define the cutoff value
cutoff <- 9

df_corrEdU <- df_long %>%
  filter(Feature == "Intensity_IntegratedIntensity_CorrEdU", !is.na(Intensity))

# Create the histogram with adjusted axes
ggplot(df_corrEdU, aes(x = Intensity)) +
  geom_histogram(
    binwidth = 0.2,   # reasonable based on your spread; adjust if needed
    fill = "blue",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = cutoff,  # define this if you haven’t already
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Histogram of Intensity_IntegratedIntensity_CorrEdU",
    x = "Intensity",
    y = "Count"
  ) +
  facet_wrap(~ Disease + Treatment) +
  scale_x_continuous(
    limits = c(0, 20),  # shows main data bulk; remove or increase to show full range
    breaks = seq(0, 20, 2)
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )


#Calculate % of cells above cutoff

df_summary <- df_corrEdU %>%
  group_by(Disease, Treatment) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

# Print the result
print(df_summary)

#Plot — compare Treatments, facet by Disease

ggplot(df_summary, aes(x = Treatment, y = percent_above, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    y = "% Cells Above Cutoff",
    x = "Treatment"
  ) +
  facet_wrap(~ Disease) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

#facet by both Disease and Treatment
##Calculate per-patient percent above cutoff
df_per_patient <- df_corrEdU %>%
  group_by(Disease, Treatment, Patient) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

#Summarize for bars (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(
    mean_percent = mean(percent_above, na.rm = TRUE),
    .groups = "drop"
  )

#Plot with bar + dot (each patient) and facet grid
ggplot() +
  # Bar: average percentage per disease-treatment group
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_percent, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # Dots: individual patient values
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = percent_above),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +  # one panel per Treatment
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    x = "Disease",
    y = "% Cells Above Cutoff"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )


````

# Create the box plot

````{r}
# Create the plot with Healthy patients in blue
ggplot(df_aggregated, aes(x = Treatment, y = Intensity, fill = Treatment)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +  # Filled boxes with transparency
  geom_jitter(aes(color = Disease), width = 0.2, alpha = 0.5, size = 2) +  # Map 'Disease' to color
  scale_color_manual(values = c("Healthy" = "blue", "RA" = "red")) +  # Set colors for 'Disease'
  facet_wrap(~Feature, scales = "free_y") +  # Facet by feature
  labs(
    title = "Comparison of Intensity Features Across Treatments",
    subtitle = "Each dot represents one patient (RA is in red)",
    x = "Treatment",
    y = "Intensity",
    fill = "Treatment",  # Legend for fill (boxplot)
    color = "Disease"    # Legend for color (dots)
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 14),  # Increase general text size
    axis.text = element_text(size = 12),  # Increase axis text size
    axis.title = element_text(size = 14),  # Increase axis title size
    plot.title = element_text(size = 16, face = "bold"),  # Increase plot title size and make it bold
    plot.subtitle = element_text(size = 14),  # Increase plot subtitle size
    strip.text = element_text(size = 14)  # Increase facet label size
  )

````
Plot patient box side-by-side

````{r}
# Create the plot with Healthy and RA patients side-by-side
ggplot(df_long, aes(x = Treatment, y = Intensity, fill = Treatment)) +
  geom_boxplot(aes(color = Disease), outlier.shape = NA, alpha = 0.8, position = position_dodge(width = 0.8)) +  # Boxplots with side-by-side grouping
  geom_jitter(aes(color = Disease), alpha = 0.5, size = 2, position = position_dodge(width = 0.8)) +  # Jittered points with side-by-side grouping (no width argument)
  scale_color_manual(values = c("Healthy" = "blue", "RA" = "red")) +  # Set colors for 'Disease'
  scale_fill_manual(values = c("Treatment1" = "lightgray", "Treatment2" = "darkgray")) +  # Set fill colors for treatments
  facet_wrap(~Feature, scales = "free_y") +  # Facet by feature
  labs(
    title = "Comparison of Intensity Features Across Treatments",
    subtitle = "Each dot represents one patient (Healthy in blue, RA in red)",
    x = "Treatment",
    y = "Intensity",
    fill = "Treatment",  # Legend for fill (boxplot)
    color = "Disease"    # Legend for color (dots)
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    text = element_text(size = 14),  # Increase general text size
    axis.text = element_text(size = 12),  # Increase axis text size
    axis.title = element_text(size = 14),  # Increase axis title size
    plot.title = element_text(size = 16, face = "bold"),  # Increase plot title size and make it bold
    plot.subtitle = element_text(size = 14),  # Increase plot subtitle size
    strip.text = element_text(size = 14)  # Increase facet label size
  )

````
# Handle Lipidtox data

````{r}
# Plot histogram for Intensity_IntegratedIntensity_CorrLipid
# to select cut off

ggplot(
    df_long[df_long$Feature == "Intensity_IntegratedIntensity_CorrLipid", ],
    aes(x = Intensity)
) +
    geom_histogram(bins = 100, fill = "steelblue", color = "black") +
    scale_x_continuous(
        limits = c(0, quantile(df_long$Intensity[df_long$Feature == "Intensity_IntegratedIntensity_CorrLipid"], 0.99)),
        expand = expansion(mult = c(0, 0.05))  # small padding on the right
    ) +
    labs(
        title = "Histogram of Intensity_IntegratedIntensity_CorrLipid",
        x = "Integrated Intensity",
        y = "Cell Count"
    ) +
    theme_minimal()

# Calculate the 75th percentile (Q3)
q75 <- quantile(
  df_long$Intensity[df_long$Feature == "Intensity_IntegratedIntensity_CorrLipid"],
  probs = 0.75,
  na.rm = TRUE
)

q75

# Define the cutoff value => according from 75%
cutoff <- 969


df_corrLipid <- df_long %>%
  filter(Feature == "Intensity_IntegratedIntensity_CorrLipid", !is.na(Intensity))

# Create the histogram with adjusted axes
ggplot(df_corrLipid, aes(x = Intensity)) +
  geom_histogram(
    binwidth = 20,   # reasonable based on your spread; adjust if needed
    fill = "blue",
    alpha = 0.7
  ) +
  geom_vline(
    xintercept = cutoff,  # define this if you haven’t already
    color = "red",
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Histogram of Intensity_IntegratedIntensity_CorrLipid",
    x = "Intensity",
    y = "Count"
  ) +
  facet_wrap(~ Disease + Treatment) +
  scale_x_continuous(
    limits = c(0, 3000),  # shows main data bulk; remove or increase to show full range
    breaks = seq(0, 20, 2)
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )


#Calculate % of cells above cutoff

df_summary <- df_corrLipid %>%
  group_by(Disease, Treatment) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

# Print the result
print(df_summary)

#Plot — compare Treatments, facet by Disease

ggplot(df_summary, aes(x = Treatment, y = percent_above, fill = Treatment)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    y = "% Cells Above Cutoff",
    x = "Treatment"
  ) +
  facet_wrap(~ Disease) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

#facet by both Disease and Treatment
##Calculate per-patient percent above cutoff
df_per_patient <- df_corrLipid %>%
  group_by(Disease, Treatment, Patient) %>%
  summarise(
    total_cells = n(),
    above_cutoff = sum(Intensity > cutoff),
    percent_above = (above_cutoff / total_cells) * 100
  ) %>%
  ungroup()

#Summarize for bars (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(
    mean_percent = mean(percent_above, na.rm = TRUE),
    .groups = "drop"
  )

#Plot with bar + dot (each patient) and facet grid
ggplot() +
  # Bar: average percentage per disease-treatment group
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_percent, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # Dots: individual patient values
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = percent_above),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +  # one panel per Treatment
  labs(
    title = paste("Percentage of Cells with Intensity >", cutoff),
    x = "Disease",
    y = "% Cells Above Cutoff"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

````
# Handle MitoSOX data

````{r}
# Plot histogram for Intensity_MeanIntensity_CorrMitosox
#or Intensity_IntegratedIntensity_CorrMitosox
# to select cut off

df_filtered <- df_long %>%
  filter(Feature == "Intensity_MeanIntensity_CorrMitosox")

#Summarize to get mean intensity per patient
df_per_patient <- df_filtered %>%
  group_by(Patient, Disease, Treatment) %>%
  summarise(mean_intensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


#Create summary data for bar plot (mean per group)
df_summary <- df_per_patient %>%
  group_by(Disease, Treatment) %>%
  summarise(mean_intensity = mean(mean_intensity, na.rm = TRUE), .groups = "drop")

# 4. Plot
ggplot() +
  # Bars: mean across patients
  geom_col(
    data = df_summary,
    aes(x = Disease, y = mean_intensity, fill = Disease),
    width = 0.6,
    alpha = 0.8
  ) +
  # One dot per patient
  geom_jitter(
    data = df_per_patient,
    aes(x = Disease, y = mean_intensity),
    width = 0.2,
    size = 2,
    shape = 21,
    fill = "black",
    color = "white"
  ) +
  facet_wrap(~ Treatment) +
  labs(
    title = "Mean CorrMitosox Intensity per Patient",  #change to Mean if use
    x = "Disease",
    y = "Mean Intensity per Patient"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold"),
    strip.text = element_text(size = 14)
  )

````
#Filter one function at a time

```{r}
#Mitosox <- mitochondrial ROS
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrMitosox"),  # keep only those rows
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Mitochondrial ROS pathway (MitoSOX)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))



```

Lipid TOX 

```{r}
#LipidTOX <- Neural lipid
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrLipid"),  # keep only those rows
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Neutral lipid pathway (LipidTOX)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

```

EdU

```{r}
#EdU <- proliferation
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrEdU"),  # keep only those rows
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Neutral lipid pathway (LipidTOX)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

````
#compare healthy, RA, OA

````{r}
#all features

sce_cont <- sce[, sce$Treatment == "control"]
 
result <- fitModel(sce_cont,
                   target = "Disease", 
                   interest_level = "RA", reference_level = "OA", 
                   group = "Patient", strata = "Disease",
                   n_folds = 20, n_threads = 8)


plotROC(result)

#for selected features

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]
sce_cont <- sce_select[, sce_select$Treatment == "control"]

result <- fitModel(sce_cont,
                   target = "Disease", 
                   interest_level = "OA", reference_level = "Healthy", 
                   group = "Patient", strata = "Disease",
                   n_folds = 20, n_threads = 8)


plotROC(result)

##only one function at a time

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrEdU"),  #CorrMitosox, CorrEdU,CorrLipid
  ]  
sce_cont <- sce_select[, sce_select$Treatment == "control"]

rownames(sce_select)

result <- fitModel(sce_cont,
                   target = "Disease", 
                   interest_level = "OA", reference_level = "Healthy", 
                   group = "Patient", strata = "Disease",
                   n_folds = 10, n_threads = 8)


plotROC(result)



````
#compare joint differences

````{r}

##only work for Joint that has in at least 2 diseases
sce_cont <- sce[, sce$Treatment == "control"]
 
#1
result <- fitModel(sce_cont,
                   target = "Joint", 
                   interest_level = "wrist", reference_level = "mcp", 
                   group = "Patient", strata = "Disease",
                   n_folds = 5, n_threads = 8)  #only work with 5 folds-> low sample

plotROC(result)

### can't do per function analysis error; No event observations were detected in `truth` with event level 'mcp'.

#2
result <- fitModel(sce_cont,
                   target = "Joint", 
                   interest_level = "shoulder", reference_level = "knee", 
                   group = "Patient", strata = "Disease",
                   n_folds = 10, n_threads = 8)


plotROC(result)

##only one function at a time

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrLipid"),  #CorrMitosox, CorrEdU,CorrLipid
  ]  
sce_cont <- sce_select[, sce_select$Treatment == "control"]

rownames(sce_select)

result <- fitModel(sce_cont,
                   target = "Joint", 
                   interest_level = "shoulder", reference_level = "knee", 
                   group = "Patient", strata = "Disease",
                   n_folds = 5, n_threads = 8)

plotROC(result)

#3
result <- fitModel(sce_cont,
                   target = "Joint", 
                   interest_level = "wrist", reference_level = "knee", 
                   group = "Patient", strata = "Disease",
                   n_folds = 5, n_threads = 8)


plotROC(result)

##only one function at a time

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrEdU"),  #CorrMitosox, CorrEdU,CorrLipid
  ]  
sce_cont <- sce_select[, sce_select$Treatment == "control"]

rownames(sce_select)

result <- fitModel(sce_cont,
                   target = "Joint", 
                   interest_level = "wrist", reference_level = "knee", 
                   group = "Patient", strata = "Disease",
                   n_folds = 5, n_threads = 8)

plotROC(result)


```
## Unsupervised Analysis PCA

````{r}
sce <- runPCA(sce, exprs_values = "tfmfeatures")
plotReducedDim(sce, dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce, dimred = "PCA", colour_by = "Plate")

sce_select <- sce[str_detect(names(sce), "Intensity"), ]
sce_select <- runPCA(sce_select, exprs_values = "tfmfeatures", ncomponents = 10)
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Plate")

```

What is driving PC1?

```{r}
plotPCACor(sce_select, filter_by = 1, top = 20)
```

