---
title: "Pre-Validation Analysis for CellProfiler"
author: "Christof Seiler"
date: now
format: html
editor: source
embed-resources: true
---

## Load Data

Load our package.

```{r load_pkgs, message=FALSE}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(stringr)
library(dplyr)
library(batchelor)
library(BiocParallel)
library(tidyr)
library(MultiAssayExperiment)
library(lme4)
library(lmerTest)
n_threads <- 16

```

Load `CellProfiler` output and create a `SingleCellExperiment` object.

```{r}
# path to files
# file_standard <- paste0("/shares/rheumatologie.usz/mim/250410 Cellpainting_RBPJ/MyExpt_Cells.csv")
# file_metabolism <- paste0("/shares/rheumatologie.usz/mim/250410 metabolism_RBPJ/MyExpt_Cells.csv")
# file_stress <- paste0("/shares/rheumatologie.usz/mim/250410 stress panel_RBPJ/MyExpt_Cells.csv")

# for CRISPR experiment
file_standard <- paste0("/shares/rheumatologie.usz/mim/250925 - RBPJ KO_Cellpainting/MyExpt_Cells.csv")
file_metabolism <- paste0("/shares/rheumatologie.usz/Ege/25092025 - RBPJ KO/metabolism/output/MyExpt_Cells.csv")
file_stress <- paste0("/shares/rheumatologie.usz/Ege/25092025 - RBPJ KO/stress/output/MyExpt_Cells.csv")

# load each plate into a separate bioconductor object
sce_standard <- loadData(file_standard)
sce_standard$Panel <- "standard"
sce_metabolism <- loadData(file_metabolism)
sce_metabolism$Panel <- "metabolism"
sce_stress <- loadData(file_stress)
sce_stress$Panel <- "stress"

# merge se objects on common features
sce_list <- list(standard = sce_standard, 
                 metabolism = sce_metabolism, 
                 stress = sce_stress)

```

## Data Transformation

Prepare the data for machine learning.

```{r}
data_transformation <- function(sce) {

  # remove cells with missing features
  sce <- removeNAs(sce)
    
  # remove outlier images
  sce <- removeOutliers(sce, min = 5, max = 300)
    
  # remove outlier cells
  stats <- perCellQCMetrics(sce, assay.type = "features")
  sce <- sce[, !isOutlier(stats$sum)]
    
  # remove features with zero spread
  sce <- removeLowVariance(sce, robust = TRUE)
    
  # remove zero-inflated features
  sce <- removeZeroInflation(sce)
    
  # transform non-negatives features and scale all features
  sce <- transformLogScale(sce, robust = TRUE)
  
  # fix factors
  sce$Treatment <- as.factor(sce$Treatment)
  sce$Treatment <- relevel(sce$Treatment, ref = "control")
  sce$Group <- as.factor(sce$Group)
  sce$Group <- relevel(sce$Group, ref = "Cas9")
  sce
  
}
sce_list <- lapply(sce_list, data_transformation)
lapply(sce_list, plotCellsPerImage)

```
#Pseudo-bulk per well.

```{r}
sce_list_aggr <- lapply(sce_list, function(sce) {
  
  aggregateAcrossCells(
    sce,
    id = colData(sce)[, c("ImageNumber", "Well", "Patient", "Plate")],
    use.assay.type = "tfmfeatures",
    statistics = "mean"
    )
  
})

```
## Unsupervised Analysis

Use `scater` for exploratory data analysis.

```{r}
# sce_list_aggr <- lapply(sce_list_aggr, runPCA, exprs_values = "tfmfeatures")
# lapply(sce_list_aggr, plotReducedDim, dimred = "PCA", colour_by = "Group")


sce_list <- lapply(sce_list, runPCA, exprs_values = "tfmfeatures")
lapply(sce_list, plotReducedDim, dimred = "PCA", colour_by = "Treatment")

```

What is driving PC1?

```{r}
# lapply(sce_list_aggr, plotPCACor, filter_by = 1, top = 20)

lapply(sce_list, plotPCACor, filter_by = 1, top = 20)

```
Check batch effect for just one feature.

```{r}
tb_sce <- makePerCellDF(sce_list$stress, 
                        assay.type = "tfmfeatures",
                        features = "Texture_Entropy_CorrTMRM_10_02_256"
                        )
tb_sce |> 
  ggplot(aes(Texture_Entropy_CorrTMRM_10_02_256, 
             group = Patient, color = Plate)) + 
  geom_density() + 
  facet_grid(Treatment ~ Disease)

```

Remove features.

```{r}
# feature_ids <- which(
#   rownames(sce_list_aggr$standard) %in% c("Correlation_Costes_CorrMito_CorrBlue")
#   )
# sce_list_aggr$standard <- sce_list_aggr$standard[-feature_ids, ]


# standard
feature_ids <- which(
  rownames(sce_list_aggr$standard) %in% c("Correlation_Manders_CorrCyto_CorrSYTO", "Correlation_Costes_CorrCyto_CorrBlue","Correlation_Costes_CorrMito_CorrBlue","Correlation_Costes_CorrER_CorrBlue","Correlation_Costes_CorrSYTO_CorrBlue","Correlation_Costes_CorrMito_CorrSYTO","Correlation_Costes_CorrMito_CorrER", "Correlation_Manders_CorrER_CorrMito")
  )
sce_list_aggr$standard <- sce_list_aggr$standard[-feature_ids, ]


# metabolism
feature_ids <- which(
  rownames(sce_list$metabolism) %in% c("Correlation_Costes_CorrMitosox_CorrBlue", "Correlation_Costes_CorrLipid_CorrBlue")
  )
sce_list$metabolism <- sce_list$metabolism[-feature_ids, ]


# stress

patterns <- c(
  "^Texture_SumEntropy_CorrTMRM_5",
  "^Texture_SumEntropy_CorrTMRM_3",
  "^Texture_Entropy_CorrTMRM_5",
  "^Texture_Entropy_CorrTMRM_3"
)

# Create one combined regex
pattern_regex <- paste(patterns, collapse = "|")

# Identify features to remove
feature_ids <- grep(pattern_regex, rownames(sce_list$stress))

# Remove them
sce_list$stress <- sce_list$stress[-feature_ids, ]


# feature_ids <- which(
#   rownames(sce_list$stress) %in% c("Texture_SumEntropy_CorrTMRM_5_03_256", "Texture_SumEntropy_CorrTMRM_5_02_256")
#   )
# sce_list$stress <- sce_list$stress[-feature_ids, ]

```

```{r}

```

## Pre-Validation Analysis

Standard panel.

```{r}
# filter to controls only
standard <- sce_list_aggr$standard
standard <- standard[, standard$Treatment == "control"]

# comparison
reference_level <- "Cas9"      #Cas9 cont
interest_level <- "KO"         #KO RBPJ

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrCyto", "CorrER", "CorrMito", "CorrSYTO", "CorrBlue")

standard <- predictLOO(
  standard, target = "Group",  #Group
  interest_level = interest_level, reference_level = reference_level,
  types = NULL, channels = channels,
  weights = "ncells", n_threads = n_threads
  )

#plotLOO(standard, meta_vars = c("Patient", "Group"), target = "Group")  #cas9 data change Group-> Group

plotLOO(standard, meta_vars = c("Patient", "Group"), target = "Group") + geom_line(aes(group = Patient), color = "darkgray")

plotROC(standard, meta_vars = c("Patient", "Group"), target = "Group")
plotAUC(standard, meta_vars = c("Patient", "Group"), target = "Group")
volcanoPlot(standard, 
            meta_vars = c("Patient", "Group"),
            target = "Group")

p_loo <- plotLOO(standard, meta_vars = c("Patient", "Group"), target = "Group")+ geom_line(aes(group = Patient), color = "darkgray")
p_roc <- plotROC(standard, meta_vars = c("Patient", "Group"), target = "Group")
p_auc <- plotAUC(standard, meta_vars = c("Patient", "Group"), target = "Group")
p_volcano <- volcanoPlot(
    standard,
    meta_vars = c("Patient", "Group"),
    target = "Group"
)

save_pub <- function(plot, filename, width = 6, height = 5) {
    ggsave(paste0(filename, ".png"), plot = plot, width = width, height = height, dpi = 300)
}
save_pub(p_loo, "LOO_plot_standard")
save_pub(p_roc, "ROC_plot_standard")
save_pub(p_auc, "AUC_plot_standard")
save_pub(p_volcano, "Volcano_plot_standard")

```

Metabolism panel.

```{r}
# filter to controls only
metabolism <- sce_list_aggr$metabolism
metabolism <- metabolism[, metabolism$Treatment == "control"]

# channels vary across panels
channels <- c("CorrBlue", "CorrLipid", "CorrEdU", "CorrMitosox")

metabolism <- predictLOO(
  metabolism, target = "Group",
  interest_level = interest_level, reference_level = reference_level,
  types = NULL, channels = channels,
  weights = "ncells", n_threads = n_threads
  )

plotLOO(metabolism, meta_vars = c("Patient", "Group"), target = "Group")
plotROC(metabolism, meta_vars = c("Patient", "Group"), target = "Group")
plotAUC(metabolism, meta_vars = c("Patient", "Group"), target = "Group")
volcanoPlot(metabolism, 
            meta_vars = c("Patient", "Group"),
            target = "Group")

p_loo <- plotLOO(metabolism, meta_vars = c("Patient", "Group"), target = "Group")+ geom_line(aes(group = Patient), color = "darkgray")
p_roc <- plotROC(metabolism, meta_vars = c("Patient", "Group"), target = "Group")
p_auc <- plotAUC(metabolism, meta_vars = c("Patient", "Group"), target = "Group")
p_volcano <- volcanoPlot(
    metabolism,
    meta_vars = c("Patient", "Group"),
    target = "Group"
)

save_pub <- function(plot, filename, width = 6, height = 5) {
    ggsave(paste0(filename, ".png"), plot = plot, width = width, height = height, dpi = 300)
}
save_pub(p_loo, "LOO_plot_metabolism")
save_pub(p_roc, "ROC_plot_metabolism")
save_pub(p_auc, "AUC_plot_metabolism")
save_pub(p_volcano, "Volcano_plot_metabolism")

```

Stress panel.

```{r}
# filter to controls only
stress <- sce_list_aggr$stress
stress <- stress[, stress$Treatment == "control"]

# channels vary across panels
channels <- c("CorrCytoID", "CorrBlue", "CorrTMRM", "CorrLyso")

stress <- predictLOO(
  stress, target = "Group",
  interest_level = interest_level, reference_level = reference_level,
  types = NULL, channels = channels,
  weights = "ncells", n_threads = n_threads
  )

plotLOO(stress, meta_vars = c("Patient", "Group"), target = "Group")
plotROC(stress, meta_vars = c("Patient", "Group"), target = "Group")
plotAUC(stress, meta_vars = c("Patient", "Group"), target = "Group")
volcanoPlot(stress, 
            meta_vars = c("Patient", "Group", "Panel"),
            target = "Group")

p_loo <- plotLOO(stress, meta_vars = c("Patient", "Group"), target = "Group")+ geom_line(aes(group = Patient), color = "darkgray")
p_roc <- plotROC(stress, meta_vars = c("Patient", "Group"), target = "Group")
p_auc <- plotAUC(stress, meta_vars = c("Patient", "Group"), target = "Group")
p_volcano <- volcanoPlot(
    stress,
    meta_vars = c("Patient", "Group"),
    target = "Group"
)

save_pub <- function(plot, filename, width = 6, height = 5) {
    ggsave(paste0(filename, ".png"), plot = plot, width = width, height = height, dpi = 300)
}
save_pub(p_loo, "LOO_plot_stress")
save_pub(p_roc, "ROC_plot_stress")
save_pub(p_auc, "AUC_plot_stress")
save_pub(p_volcano, "Volcano_plot_stress")

```

Combine volcano plot.

```{r}
named_list <- list(standard = standard, 
                   metabolism = metabolism, 
                   stress = stress)
p <- volcanoPlot(
  named_list,
  meta_vars = c("Patient", "Group"),
  target = "Group", 
  p_cutoff = 0.1, 
  fc_cutoff = 0.25
)

ggsave("volcano_plot.png", plot = p, width = 8, height = 6)

```

Calculate effect sizes (log-odds) and p-values (paired ranked test).

```{r}
stats <- calculateStats(sce_list)

stats$Panel <- factor(stats$Panel, c("standard", "metabolism", "stress"))
volcanoPlot(stats, facet = "Target", color = "Panel")
```

## Reproducibilty

```{r}
sessionInfo()
```
# EdU

```{r}
sce_met <- sce_list$metabolism
intensity_data <- as.data.frame(t(assay(sce_met, "tfmfeatures"))) 
metadata <- as.data.frame(colData(sce_met)[, c("Plate", "Well", "Patient", "Treatment", "Group")])

# Combine
df <- cbind(metadata, intensity_data)

#plot histogram
ggplot(df, aes(x = Intensity_MeanIntensity_CorrEdU)) +
  geom_histogram(
    bins = 50,
    fill = "steelblue",
    color = "black",
    alpha = 0.6
  ) +
  geom_vline(
    xintercept = 4, 
    color = "red", 
    linetype = "dashed", 
    linewidth = 1
  ) +
  theme_bw(base_size = 14) +
  labs(
    title = "Histogram of Intensity_MeanIntensity_CorrEdU",
    x = "Intensity_MeanIntensity_CorrEdU (a.u.)",
    y = "Cell count"
  )

# filter for control
df_control <- df %>% 
  filter(Treatment == "control")

# Compute % positive per patient
cutoff <- 4

df_patient <- df %>%
  mutate(Pos = Intensity_MeanIntensity_CorrEdU > cutoff) %>%
  group_by(Patient, Group, Treatment) %>%
  summarise(
    Total = n(),
    Positive = sum(Pos),
    Percent_Positive = (Positive / Total) * 100,
    .groups = "drop"
  )

# mean per Group × Treatment
df_avg <- df_patient %>%
  group_by(Group, Treatment) %>%
  summarise(
    Mean_Percent = mean(Percent_Positive),
    SD_Percent   = sd(Percent_Positive),
    .groups = "drop"
  )

p <- ggplot() +
  
  # bars
  geom_col(data = df_avg,
           aes(x = Group, y = Mean_Percent, fill = Group),
           color = "black", width = 0.65) +
  
  # dots
  geom_jitter(data = df_patient,
              aes(x = Group, y = Percent_Positive),
              width = 0.15, size = 3, alpha = 0.8) +
  
  # significance per facet (auto)
  stat_compare_means(
    data = df_patient,
    aes(x = Group, y = Percent_Positive),
    method = "wilcox.test",
    ref.group = "Healthy",
    label = "p.signif",
    label.y = max(df_patient$Percent_Positive) * 1.15
  ) +
  
  facet_wrap(~ Treatment) +
  
  scale_fill_brewer(palette = "Set2") +
  
  theme_bw(base_size = 18) +
  labs(
    title = paste0("Percentage of CorrEdU+ Cells (x > ", cutoff, ")"),
    x = "Group",
    y = "Percent Positive Cells (%)",
    fill = "Group"
  ) +
  theme(
    panel.grid = element_blank(),
    legend.position = "none"
  )

print(p)



# save plot
ggsave("CorrEdU_percent_positive_by groupXtreatment.png",
       plot = p, width = 6, height = 5, dpi = 300)
