---
title: "Stress"
author: "Phatthamon"
date: now
format: html
editor: source
embed-resources: true
---

## Load Data

```{r}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(stringr)
library(dplyr)
library(batchelor)
library(BiocParallel)
library(tidyr)
library(MultiAssayExperiment)
library(lme4)
library(lmerTest)
library(ggrepel)
n_threads <- 16

```

Path to files.

```{r}

# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# stress panel
files_stress <- c(
  paste0(base, "250709 cytokines_stress/MyExpt_Cells.csv"), 
  paste0(base, "250717 cytokines_stress lot2/MyExpt_Cells.csv")
  )
sce <- Reduce(cbind, lapply(files_stress, loadData))


```

merge with previous exp.

```{r}

# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# stress panel
files_stress <- c(
  paste0(base, "250709 cytokines_stress/MyExpt_Cells.csv"), 
  paste0(base, "250717 cytokines_stress lot2/MyExpt_Cells.csv"),
  paste0(base, "250320 stress panel_cytokine/MyExpt_Cells.csv")
  )
#in case some plate have different column name_plate from experiment 250320
sce_list <- lapply(files_stress, loadData)

# Add missing colData columns dynamically
all_cols <- unique(unlist(lapply(sce_list, function(x) colnames(colData(x)))))

sce_list_fixed <- lapply(sce_list, function(x) {
  missing_cols <- setdiff(all_cols, colnames(colData(x)))
  for (col in missing_cols) {
    colData(x)[[col]] <- NA
  }
  x
})

sce <- Reduce(cbind, sce_list_fixed)

```

Path to files..for panel development plate

```{r}
# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

#stress panel develope plate
files_stress <- c(paste0(base, "250226 stress panel_TMRM/MyExpt_Cells.csv"))

sce_list <- lapply(files_stress, loadData)

# If you only have one file, you can get the SCE directly
sce <- sce_list[[1]]

```

## Data Transformation

Prepare the data for machine learning.

```{r}
data_transformation <- function(sce) {

  # per plate
  plates <- unique(sce[["Plate"]])

  plate_list <- lapply(plates, function(plate_id) {

    # normalize per plate
    sce_plate <- sce[, sce[["Plate"]] == plate_id]

    # remove cells with missing features
    sce_plate <- removeNAs(sce_plate)
    
    # remove outlier images
    sce_plate <- removeOutliers(sce_plate, min = 30, max = 300)
    
    # remove outlier cells
    stats <- perCellQCMetrics(sce_plate, assay.type = "features")
    sce_plate <- sce_plate[, !isOutlier(stats$sum)]
    
    # remove features with zero spread
    sce_plate <- removeLowVariance(sce_plate, robust = FALSE)
    
    # transform non-negatives features and scale all features
    sce_plate <- transformLogScale(sce_plate, robust = TRUE)
  
    # fix factors
    sce_plate$Treatment <- as.factor(sce_plate$Treatment)
    sce_plate$Treatment <- relevel(sce_plate$Treatment, ref = "control")
    
    sce_plate
  
  })

  # combine plates
  feature_name_list <- lapply(plate_list,
                              function(sce_plate) rownames(sce_plate))
  include_features <- Reduce(intersect, feature_name_list)
  plate_list <- lapply(plate_list,
                       function(sce_plate) sce_plate[include_features, ])
  sce <- do.call(cbind, plate_list)
  sce$Plate <- as.factor(sce$Plate)
  sce
  
}
sce <- data_transformation(sce)

lapply(unique(sce[["Plate"]]), function(plate_id) {
  
  plotCellsPerImage(sce[, sce[["Plate"]] == plate_id])
  
})


#change treatment name -> these 2 plate have IFNg just like the ohter batch
# colData(sce)$Treatment[colData(sce)$Plate %in% c("140001", "150001") &
#                        colData(sce)$Treatment == "IFNa"] <- "IFN"
# 
# colData(sce)$Disease[colData(sce)$Plate %in% c("140001", "150001") &
#                        colData(sce)$Disease == "healthy"] <- "Healthy"

```

Pseudo-bulk over images.

```{r}
sce_aggr <- aggregateAcrossCells(
  sce,
  id = colData(sce)[, c("ImageNumber", "Well", "Patient", "Plate")], 
  use.assay.type = "tfmfeatures", 
  statistics = "mean"
  )

```

## Pre-Validation Analysis

Predict treatment.

```{r}
# comparison
reference_level <- "control"
interest_levels <- setdiff(levels(sce$Treatment), reference_level)    #change to sce_agg if uses

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrCytoID", "CorrBlue", "CorrTMRM", "CorrLyso")

sce_list <- lapply(interest_levels, function(interest_level) {
  
  # compute y_hat
  predictLOO(
    sce, 
    interest_level = interest_level, reference_level = reference_level,
    types = types, channels = channels,
    n_threads = n_threads
    )
  
})

target = "Treatment"
n_tests <- length(types)+length(channels)
meta_vars = c("Patient", "Plate", "Treatment")
lapply(sce_list, plotLOO)
lapply(sce_list, plotROC)
plotAUC(sce_list, meta_vars = meta_vars, target = target, batch = "Plate")
volcanoPlot(sce_list, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target)
```

Predict disease.

```{r}
# define new target
target = "Disease"

# set Healthy as reference level
sce_aggr$Disease <- as.factor(sce_aggr$Disease)
sce_aggr$Disease <- relevel(sce_aggr$Disease, ref = "Healthy")

# filter to control cells only and plate
sce_aggr <- sce_aggr[, sce_aggr$Treatment == "control"]

# comparison
reference_level <- "Healthy"
interest_levels <- setdiff(levels(sce_aggr$Disease), reference_level)

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrCytoID", "CorrBlue", "CorrTMRM", "CorrLyso")

# compare OA vs Healthy
sce_oa <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "OA", reference_level = "Healthy",
  types = types, channels = channels,
  n_threads = n_threads
  )

n_tests <- length(types)+length(channels)
meta_vars = c("Patient", "Disease", "Plate")
plotLOO(sce_oa, meta_vars = meta_vars, target = target)
plotROC(sce_oa, meta_vars = meta_vars, target = target)
plotAUC(sce_oa, meta_vars = meta_vars, target = target, batch = "Plate")
volcanoPlot(sce_oa, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target)

# compare RA vs Healthy
sce_ra <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "RA", reference_level = "Healthy",
  types = types, channels = channels,
  n_threads = n_threads
  )
plotLOO(sce_ra, meta_vars = meta_vars, target = target)
plotROC(sce_ra, meta_vars = meta_vars, target = target)
plotAUC(sce_ra, meta_vars = meta_vars, target = target, batch = "Plate")
volcanoPlot(sce_ra, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5, 
            meta_vars = meta_vars, target = target)

# combine OA and RA
plotAUC(list(sce_oa, sce_ra), meta_vars = meta_vars, target = target)
volcanoPlot(list(sce_oa, sce_ra), 
            p_cutoff = 0.1/(2*n_tests), 
            fc_cutoff = 0.5, 
            meta_vars = meta_vars, target = target)
```

#For Multiple model

```{r}

interest <- setdiff(unique(sce$Treatment), c("control"))
result_list <- lapply(interest, function(level) {
  fitModel(sce, 
           target = "Treatment", 
           interest_level = level, reference_level = "control", 
           group = "Patient", strata = c("Disease"), 
           n_folds = 10, n_threads = 8)
})

plotAUC(result_list)


```
Only selected features

````{r}

# overview of CellProfiler feature categories
#   include: Texture, Intensity, AreaShape, 
#   maybe include: RadialDistribution,  Neighbors, Location
#   exclude because many corner cases: Granularity, Correlation

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

interest <- setdiff(unique(sce_select$Treatment), c("control"))

result_list <- lapply(interest, function(level) {
  fitModel(sce_select,
                   target = "Treatment", 
                   interest_level = level, reference_level = "control", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 10, n_threads = 8)
  })

plotAUC(result_list)

````
#Filter one function at a time

```{r}
#TMRM <- mitochondrial health
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrTMRM"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Mitochondrial Health (TMRM)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

ggsave("AUC_TMRM-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)
````
#run each function with selected features

```{r}

#CorrCytoID <- autophagy
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrCytoID"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Autophagy (CytoID)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

ggsave("AUC_CytoID-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)



#CorrLyso <- Lysosome
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrLyso"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Lysosome"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))


ggsave("AUC_Lysosome-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)

````
#explore each markers in details (only Intensity data)

````{r}

feature_names <- rownames(sce)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrTMRM", 
                                    "Intensity_MeanIntensity_CorrCytoID", 
                                    "Intensity_MeanIntensity_CorrLyso", 
                                    "Intensity_IntegratedIntensity_CorrTMRM",
                                    "Intensity_IntegratedIntensity_CorrCytoID",
                                    "Intensity_IntegratedIntensity_CorrLyso" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Joint", "Plate")])
df <- cbind(metadata, intensity_data)  # Replace with correct column names

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Calculate mean per patient per condition
df_summary <- df_long %>%
  group_by(Patient, Treatment, Disease, Joint, Plate, Feature) %>%
  summarise(MeanIntensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


features <- unique(df_summary$Feature)

# Loop over each feature
for (f in features) {
  df_plot <- df_summary %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = MeanIntensity, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    geom_text_repel(
    aes(label = Joint, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
    ) +
    facet_wrap(~ Disease, scales = "fixed") +
    theme_bw(base_size = 14) +
    labs(
      title = paste("Robust Z-scored log-transformed feature intensities per Treatment_", f),
      y = "Z-scored log intensity", #Mean Intensity (a.u.)
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
   print(p) 
}

##split by plate

# Get list of features
features <- unique(df_summary$Feature)
plates <- unique(df_summary$Plate)

# Loop over each feature
for (f in features) {
  
  # Filter for feature
  df_feature <- df_summary %>% filter(Feature == f)
  
  # Loop over each plate
  for (pl in plates) {
    df_plot <- df_feature %>% filter(Plate == pl)
    
    p <- ggplot(df_plot, aes(x = Treatment, y = MeanIntensity, fill = Treatment)) +
      stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
      stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
      geom_point(aes(color = Treatment),
                 position = position_jitter(width = 0.1),
                 size = 3.5, stroke = 1, alpha = 0.9) +
      facet_wrap(~ Disease, scales = "free_y") +
      theme_bw(base_size = 14) +
      labs(
        title = paste("Mean", f, "by Treatment – Plate", pl),
        y = "Mean Intensity (a.u.)",
        x = "Treatment"
      ) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        strip.background = element_rect(fill = "gray90", color = "gray60"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right"
      )
    
    print(p)  # Show in RStudio viewer
    
  }
}

## plot differences from control

# Combine both plates and compute difference from control per patient
# (Assuming control treatment is named "Control" — change if needed)
control_label <- "control"

df_diff <- df_summary %>%
  group_by(Patient, Disease, Feature) %>%
  mutate(
    ControlValue = mean(MeanIntensity[Treatment == control_label], na.rm = TRUE),
    DiffFromControl = MeanIntensity - ControlValue
  ) %>%
  filter(Treatment != control_label) %>%
  ungroup()



# Loop through each feature
features <- unique(df_diff$Feature)

for (f in features) {
  df_plot <- df_diff %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = DiffFromControl, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    facet_wrap(~ Disease, scales = "fixed") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    theme_bw(base_size = 14) +
    labs(
      title = paste("Relative feature intensity compared to control -", f),
      y = "Difference from control (z-scored intensity)",
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  print(p)

}

````
#CytoID

````{r}
# Plot histogram for Intensity_MeanIntensity_CorrCytoID
# to select cut off
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # or use the assay name you used earlier
metadata <- as.data.frame(colData(sce)[, c("Plate", "Well", "Patient", "Treatment", "Disease")])

# Combine
df <- cbind(metadata, intensity_data)

# 3. Compute 75th percentile per Plate
percentiles <- df %>%
  group_by(Plate) %>%
  summarise(P75 = quantile(Intensity_IntegratedIntensity_CorrCytoID, 0.75, na.rm = TRUE)) #or Intensity_IntegratedIntensity_CorrCytoID, Intensity_MeanIntensity_CorrCytoID

print(percentiles)

# 4. Plot histogram with vertical line per plate
ggplot(df, aes(x = Intensity_IntegratedIntensity_CorrCytoID)) +
  geom_histogram(fill = "#2C7BB6", alpha = 0.5, bins = 40, color = "black") +
  geom_vline(
    data = percentiles,
    aes(xintercept = P75),
    color = "red", linetype = "dashed", size = 1
  ) +
  facet_wrap(~ Plate, scales = "free_y") +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Distribution of Intensity (CorrCytoID) per Plate",
    x = "Intensity (CorrCytoID)",
    y = "Cell Count",
    caption = "Red dashed line = 75th percentile"
  )

#plot positive cell >75th percentile

# 3. Assign positivity per cell
df <- df %>%
  left_join(percentiles, by = "Plate") %>%
  mutate(Positive = Intensity_MeanIntensity_CorrCytoID > P75)

# 4. Summarize percent positive per Patient x Treatment x Plate
df_percent <- df %>%
  group_by(Patient, Treatment, Plate, Disease) %>%
  summarise(PercentPositive = mean(Positive) * 100, .groups = "drop")

# 5. Plot percent positive cells, combine both plates
ggplot(df_percent, aes(x = Treatment, y = PercentPositive, fill = Treatment)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
  geom_point(aes(shape = Plate, color = Treatment),
             position = position_jitter(width = 0.15),
             size = 3.5, alpha = 0.9) +
  geom_text_repel(
    aes(label = Patient, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
  ) +
  facet_wrap(~ Disease, scales = "fixed") +
  theme_bw(base_size = 14) +
  labs(
    title = "Percent CytoID Positive Cells per Treatment (Cutoff = 75th percentile per Plate)",
    y = "Percent Positive Cells (%)",
    x = "Treatment"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = "gray60"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )


````
#compare by Disease

````{r}
# 1. Extract features + metadata
intensity_data <- as.data.frame(t(assay(sce, "features")))

# Select only the features you want
selected_features <- c(
  "Intensity_MeanIntensity_CorrTMRM", 
  "Intensity_MeanIntensity_CorrCytoID", 
  "Intensity_MeanIntensity_CorrLyso", 
  "Intensity_IntegratedIntensity_CorrTMRM",
  "Intensity_IntegratedIntensity_CorrCytoID",
  "Intensity_IntegratedIntensity_CorrLyso"
)
intensity_data <- intensity_data[, selected_features]

metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Plate")])

# Combine metadata and intensity data
df <- cbind(metadata, intensity_data)

# Filter only control samples
df_control <- df %>% filter(Treatment == "control")

# Loop through each selected feature
for (feature in selected_features) {
  
  # Summarize mean per patient x disease x plate
  df_summary <- df_control %>%
    group_by(Patient, Disease, Plate) %>%
    summarise(MeanIntensity = mean(.data[[feature]], na.rm = TRUE), .groups = "drop")
  
  # Plot
  p <- ggplot(df_summary, aes(x = Disease, y = MeanIntensity, fill = Disease)) +
    stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Disease),
               position = position_jitter(width = 0.1),
               size = 3.5) +
    geom_text_repel(
      aes(label = Patient, color = Disease),
      size = 3.5,
      position = position_jitter(width = 0.1),
      show.legend = FALSE
    ) +
    #facet_wrap(~ Plate) + 
    theme_bw(base_size = 14) +
    labs(
      title = paste("Mean", feature, "per Disease (Control Samples)"),
      y = "Mean Intensity (a.u.)",
      x = "Disease"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  # Print plot
  print(p)
}

````
## Unsupervised Analysis PCA

````{r}
sce <- runPCA(sce, exprs_values = "tfmfeatures")
plotReducedDim(sce, dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce, dimred = "PCA", colour_by = "Plate")

sce_select <- sce[str_detect(names(sce), "Intensity"), ]
sce_select <- runPCA(sce_select, exprs_values = "tfmfeatures", ncomponents = 10)
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Plate")

```

What is driving PC1?

```{r}
plotPCACor(sce_select, filter_by = 1, top = 20)
```
