---
title: "Stress"
author: "Phatthamon"
date: now
format: html
editor: source
embed-resources: true
---

## Load Data

```{r}
library(cellpaintr)
library(ggplot2)
library(scater)
library(scran)
library(bluster)
library(stringr)
library(dplyr)
library(batchelor)
library(BiocParallel)
library(tidyr)
library(MultiAssayExperiment)
library(lme4)
library(lmerTest)
library(ggrepel)
library(ggpubr)
library(viridis)
library(rstatix)
n_threads <- 16

```

Path to files.

```{r}

# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# stress panel
files_stress <- c(
  paste0(base, "250709 cytokines_stress/MyExpt_Cells.csv"), 
  paste0(base, "250717 cytokines_stress lot2/MyExpt_Cells.csv")
  )
sce <- Reduce(cbind, lapply(files_stress, loadData))


```

merge with previous exp.

```{r}

# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# stress panel
files_stress <- c(
  paste0(base, "250709 cytokines_stress/MyExpt_Cells.csv"), 
  paste0(base, "250717 cytokines_stress lot2/MyExpt_Cells.csv"),
  paste0(base, "250320 stress panel_cytokine/MyExpt_Cells.csv")
  )
#in case some plate have different column name_plate from experiment 250320
sce_list <- lapply(files_stress, loadData)

# Add missing colData columns dynamically
all_cols <- unique(unlist(lapply(sce_list, function(x) colnames(colData(x)))))

sce_list_fixed <- lapply(sce_list, function(x) {
  missing_cols <- setdiff(all_cols, colnames(colData(x)))
  for (col in missing_cols) {
    colData(x)[[col]] <- NA
  }
  x
})

sce <- Reduce(cbind, sce_list_fixed)

```

Path to files..for panel development plate

```{r}
# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

#stress panel develope plate
files_stress <- c(paste0(base, "250226 stress panel_TMRM/MyExpt_Cells.csv"))

sce_list <- lapply(files_stress, loadData)

# If you only have one file, you can get the SCE directly
sce <- sce_list[[1]]

```
Path to files.-> for experiment 250320

```{r}
# path to files
base <- paste0("/shares/rheumatologie.usz/mim/")

# stress panel
files_stress <- c(
  paste0(base, "250320 stress panel_cytokine/MyExpt_Cells.csv")
  )
sce <- Reduce(cbind, lapply(files_stress, loadData))

# this experiment has 2 plates -> need to merge them
# Select the two plates
sce_140 <- sce[, sce$Plate == "140001"]
sce_150 <- sce[, sce$Plate == "150001"]

# Combine them (bind cells/columns)
sce_merged <- cbind(sce_140, sce_150)

# Rename the Plate column to the same value for all
sce_merged$Plate <- factor("140001")

# Check the result
table(sce_merged$Patient, sce_merged$Plate)

#now can go for transform step
```
## Data Transformation

Prepare the data for machine learning.

```{r}
data_transformation <- function(sce) {

  # per plate
  plates <- unique(sce[["Plate"]])

  plate_list <- lapply(plates, function(plate_id) {

    # normalize per plate
    sce_plate <- sce[, sce[["Plate"]] == plate_id]

    # remove cells with missing features
    sce_plate <- removeNAs(sce_plate)
    
    # remove outlier images
    sce_plate <- removeOutliers(sce_plate, min = 30, max = 300)
    
    # remove outlier cells
    stats <- perCellQCMetrics(sce_plate, assay.type = "features")
    sce_plate <- sce_plate[, !isOutlier(stats$sum)]
    
    # remove features with zero spread
    sce_plate <- removeLowVariance(sce_plate, robust = TRUE)
    
    # remove zero-inflated features
    sce_plate <- removeZeroInflation(sce_plate)
    
    # transform non-negatives features and scale all features
    sce_plate <- transformLogScale(sce_plate, robust = TRUE)
  
    # fix factors
    sce_plate$Treatment <- as.factor(sce_plate$Treatment)
    sce_plate$Treatment <- relevel(sce_plate$Treatment, ref = "cont")
    
    sce_plate
  
  })

  # combine plates
  feature_name_list <- lapply(plate_list,
                              function(sce_plate) rownames(sce_plate))
  include_features <- Reduce(intersect, feature_name_list)
  plate_list <- lapply(plate_list,
                       function(sce_plate) sce_plate[include_features, ])
  sce <- do.call(cbind, plate_list)
  sce$Plate <- as.factor(sce$Plate)
  sce
  
}
sce <- data_transformation(sce)

lapply(unique(sce[["Plate"]]), function(plate_id) {
  
  plotCellsPerImage(sce[, sce[["Plate"]] == plate_id])
  
})


#change treatment name -> these 2 plate have IFNg just like the ohter batch
# colData(sce)$Treatment[colData(sce)$Plate %in% c("140001", "150001") &
#                        colData(sce)$Treatment == "IFNa"] <- "IFN"
# 
# colData(sce)$Disease[colData(sce)$Plate %in% c("140001", "150001") &
#                        colData(sce)$Disease == "healthy"] <- "Healthy"

```

Pseudo-bulk over images.

```{r}
sce_aggr <- aggregateAcrossCells(
  sce,
  id = colData(sce)[, c("ImageNumber", "Well", "Patient", "Plate")], 
  use.assay.type = "tfmfeatures", 
  statistics = "mean"
  )

```

## Pre-Validation Analysis

Predict treatment.

```{r}
# comparison
reference_level <- "control"
interest_levels <- setdiff(levels(sce$Treatment), reference_level)    #change to sce_agg if uses

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrCytoID", "CorrBlue", "CorrTMRM", "CorrLyso")

sce_list <- lapply(interest_levels, function(interest_level) {
  
  # compute y_hat
  predictLOO(
    sce, 
    interest_level = interest_level, reference_level = reference_level,
    types = types, channels = channels,
    n_threads = n_threads
    )
  
})

target = "Treatment"
n_tests <- length(types)+length(channels)
meta_vars = c("Patient", "Plate", "Treatment")
lapply(sce_list, plotLOO)
lapply(sce_list, plotROC)
plotAUC(sce_list, meta_vars = meta_vars, target = target, batch = "Plate")
volcanoPlot(sce_list, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target)
```

Predict disease.

```{r}
# define new target
target = "Disease"

# set Healthy as reference level
sce_aggr$Disease <- as.factor(sce_aggr$Disease)
sce_aggr$Disease <- relevel(sce_aggr$Disease, ref = "Healthy")

# filter to control cells only and plate
sce_aggr <- sce_aggr[, sce_aggr$Treatment == "control"]

# comparison
reference_level <- "Healthy"
interest_levels <- setdiff(levels(sce_aggr$Disease), reference_level)

# same across panels
types <- c("AreaShape", "Intensity", "Texture", "RadialDistribution", 
           "Granularity")

# channels vary across panels
channels <- c("CorrCytoID", "CorrBlue", "CorrTMRM", "CorrLyso")

# compare OA vs Healthy
sce_oa <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "OA", reference_level = "Healthy",
  types = types, channels = channels,
  n_threads = n_threads
  )

n_tests <- length(types)+length(channels)
meta_vars = c("Patient", "Disease", "Plate")
plotLOO(sce_oa, meta_vars = meta_vars, target = target)
plotROC(sce_oa, meta_vars = meta_vars, target = target)
plotAUC(sce_oa, meta_vars = meta_vars, target = target, batch = "Plate")
volcanoPlot(sce_oa, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5,
            meta_vars = meta_vars, target = target)

# compare RA vs Healthy
sce_ra <- predictLOO(
  sce_aggr, 
  target = target,
  interest_level = "RA", reference_level = "Healthy",
  types = types, channels = channels,
  n_threads = n_threads
  )
plotLOO(sce_ra, meta_vars = meta_vars, target = target)
plotROC(sce_ra, meta_vars = meta_vars, target = target)
plotAUC(sce_ra, meta_vars = meta_vars, target = target, batch = "Plate")
volcanoPlot(sce_ra, p_cutoff = 0.05/n_tests, fc_cutoff = 0.5, 
            meta_vars = meta_vars, target = target)

# combine OA and RA
plotAUC(list(sce_oa, sce_ra), meta_vars = meta_vars, target = target)
volcanoPlot(list(sce_oa, sce_ra), 
            p_cutoff = 0.1/(2*n_tests), 
            fc_cutoff = 0.5, 
            meta_vars = meta_vars, target = target)
```

#For Multiple model

```{r}

interest <- setdiff(unique(sce$Treatment), c("control"))
result_list <- lapply(interest, function(level) {
  fitModel(sce, 
           target = "Treatment", 
           interest_level = level, reference_level = "control", 
           group = "Patient", strata = c("Disease"), 
           n_folds = 10, n_threads = 8)
})

plotAUC(result_list)


```
Only selected features

````{r}

# overview of CellProfiler feature categories
#   include: Texture, Intensity, AreaShape, 
#   maybe include: RadialDistribution,  Neighbors, Location
#   exclude because many corner cases: Granularity, Correlation

sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

interest <- setdiff(unique(sce_select$Treatment), c("control"))

result_list <- lapply(interest, function(level) {
  fitModel(sce_select,
                   target = "Treatment", 
                   interest_level = level, reference_level = "control", 
                   group = "Patient", strata =  "Disease", 
                   n_folds = 10, n_threads = 8)
  })

plotAUC(result_list)

````
#Filter one function at a time

```{r}
#TMRM <- mitochondrial health
#from only some features Texture|Intensity|AreaShape

#input data again
sce_select <- sce[str_detect(names(sce), "Texture|Intensity|AreaShape"), ]

sce_select <- sce_select[
  str_detect(rownames(sce_select), "CorrTMRM"),  #change between CorrTMRM, CorrCytoID, CorrLyso
  ]  

#check if only have selected channel
rownames(sce_select)


interest <- setdiff(unique(sce_select$Treatment), c("control"))
sce_select$Treatment <- as.factor(sce_select$Treatment)
sce_select$Treatment <- droplevels(sce_select$Treatment)

result_list <- lapply(interest, function(level) {
   fitModel(sce_select,
            target = "Treatment",
            interest_level = level, reference_level = "control",
            group = "Patient", strata = c("Disease"),
            n_folds = 10, n_threads = 8)
 })

plotAUC(result_list)

#customise plot
aucs <- dplyr::bind_rows(lapply(result_list, function(result) {
    level <- stringr::str_remove(names(result)[3], ".pred_")
    mutate(yardstick::roc_auc(group_by(result, fold), names(result)[1], 
                              names(result)[3]), predict = level)
}))

#Define treatment order and colors
treatment_order <- aucs %>%
  group_by(predict) %>%
  summarise(median_auc = median(.estimate)) %>%
  arrange(desc(median_auc)) %>%
  pull(predict)

color_palette <- c("IL1" = "#E69F00",  # Orange for IL1
                   "PIC" = "#D55E00",  # Blue for others
                   "LPS" = "#009E73",
                   "IFN" = "#F0E442",
                   "TNF" = "#0072B2")

ggplot(aucs %>%
         mutate(predict = factor(predict, levels = treatment_order)),
       aes(x = predict, y = .estimate)) +
  geom_boxplot(aes(fill = predict), alpha = 0.8, outlier.shape = NA) +
  geom_jitter(width = 0.1, aes(color = predict), size = 2.5, alpha = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_color_manual(values = color_palette) +
  theme_bw() +
  theme(
    panel.background = element_rect(fill = "white"),
    plot.background = element_rect(fill = "white"),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank(),
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14, face = "bold"),
    axis.title.x = element_blank()
  ) +
  labs(
    y = "AUC",
    title = "Area under the AUC curves for Mitochondrial Health (TMRM)"
  ) +
  scale_y_continuous(limits = c(0.5, 1.0), breaks = seq(0.5, 1.0, by = 0.1))

ggsave("AUC_TMRM-selected features.png",
       plot = last_plot(),
       width = 8, height = 6, dpi = 300)
````

#explore each markers in details (only Intensity data)

````{r}

feature_names <- rownames(sce)

#see all intensity feature name
intensity_features <- grep("Intensity", feature_names, value = TRUE)
print(intensity_features)


# Extract intensity features from the assay slot
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # change back to "tfmfeatures" if work with logtransform in above step

# Filter for columns that start with "Intensity"
intensity_features <- grep("^Intensity", colnames(intensity_data), value = TRUE)
#intensity_data <- intensity_data[, intensity_features, drop = FALSE] #if want all Intensity data

intensity_data <- intensity_data[, c("Intensity_MeanIntensity_CorrTMRM", 
                                    "Intensity_MeanIntensity_CorrCytoID", 
                                    "Intensity_MeanIntensity_CorrLyso", 
                                    "Intensity_IntegratedIntensity_CorrTMRM",
                                    "Intensity_IntegratedIntensity_CorrCytoID",
                                    "Intensity_IntegratedIntensity_CorrLyso",
                                    "Intensity_MeanIntensityEdge_CorrTMRM" )]


# Extract metadata from colData
metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Joint", "Plate")])
df <- cbind(metadata, intensity_data)  # Replace with correct column names

df_long <- pivot_longer(
  df,
  cols = starts_with("Intensity"),
  names_to = "Feature",
  values_to = "Intensity"
)

# Calculate mean per patient per condition
df_summary <- df_long %>%
  group_by(Patient, Treatment, Disease, Joint, Plate, Feature) %>%
  summarise(MeanIntensity = mean(Intensity, na.rm = TRUE), .groups = "drop")


features <- unique(df_summary$Feature)

# Loop over each feature
for (f in features) {
  df_plot <- df_summary %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = MeanIntensity, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    geom_text_repel(
    aes(label = Joint, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
    ) +
    facet_wrap(~ Disease, scales = "fixed") +
    theme_bw(base_size = 14) +
    labs(
      title = paste("Z-scored log-transformed feature intensities per Treatment_", f),
      y = "Z-scored log intensity", #Mean Intensity (a.u.)
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
   print(p) 
}

##split by plate

# Get list of features
features <- unique(df_summary$Feature)
plates <- unique(df_summary$Plate)

# Loop over each feature
for (f in features) {
  
  # Filter for feature
  df_feature <- df_summary %>% filter(Feature == f)
  
  # Loop over each plate
  for (pl in plates) {
    df_plot <- df_feature %>% filter(Plate == pl)
    
    p <- ggplot(df_plot, aes(x = Treatment, y = MeanIntensity, fill = Treatment)) +
      stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
      stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
      geom_point(aes(color = Treatment),
                 position = position_jitter(width = 0.1),
                 size = 3.5, stroke = 1, alpha = 0.9) +
      facet_wrap(~ Disease, scales = "free_y") +
      theme_bw(base_size = 14) +
      labs(
        title = paste("Mean", f, "by Treatment – Plate", pl),
        y = "Mean Intensity (a.u.)",
        x = "Treatment"
      ) +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold"),
        strip.background = element_rect(fill = "gray90", color = "gray60"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right"
      )
    
    print(p)  # Show in RStudio viewer
    
  }
}

## plot differences from control

# Combine both plates and compute difference from control per patient
# (Assuming control treatment is named "Control" — change if needed)
control_label <- "control"

df_diff <- df_summary %>%
  group_by(Patient, Disease, Feature) %>%
  mutate(
    ControlValue = mean(MeanIntensity[Treatment == control_label], na.rm = TRUE),
    DiffFromControl = MeanIntensity - ControlValue
  ) %>%
  filter(Treatment != control_label) %>%
  ungroup()



# Loop through each feature
features <- unique(df_diff$Feature)

for (f in features) {
  df_plot <- df_diff %>% filter(Feature == f)
  
  p <- ggplot(df_plot, aes(x = Treatment, y = DiffFromControl, fill = Treatment)) +
    stat_summary(fun = mean, geom = "bar", color = "black", alpha = 0.4, width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(color = Treatment),
               position = position_jitter(width = 0.1),
               size = 3.5, stroke = 1, alpha = 0.9) +
    facet_wrap(~ Disease, scales = "fixed") +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    theme_bw(base_size = 14) +
    labs(
      title = paste("Relative feature intensity compared to control -", f),
      y = "Difference from control (z-scored intensity)",
      x = "Treatment"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.background = element_rect(fill = "gray90", color = "gray60"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  print(p)

}

````
# for develpoment panel -> single cell plot

````{r}

intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))
intensity_features <- c(
  "Intensity_MeanIntensity_CorrTMRM", 
  "Intensity_MeanIntensity_CorrCytoID", 
  "Intensity_MeanIntensity_CorrLyso", 
  "Intensity_IntegratedIntensity_CorrTMRM",
  "Intensity_IntegratedIntensity_CorrCytoID",
  "Intensity_IntegratedIntensity_CorrLyso",
  "Intensity_MeanIntensity_CorrBlue"
)
intensity_data <- intensity_data[, intensity_features, drop = FALSE]

metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Joint", "Plate")])
df <- cbind(metadata, intensity_data)

df <- df %>% filter(Disease == "healthy")

df_long <- df %>%
  pivot_longer(
    cols = all_of(intensity_features),
    names_to = "Feature",
    values_to = "Intensity"
  )

ref_treatment <- "cont"
treatment_colors <- RColorBrewer::brewer.pal(n = length(unique(df_long$Treatment)), name = "Set2")
names(treatment_colors) <- unique(df_long$Treatment)

# 7. Loop through all features and plot
for (f in unique(df_long$Feature)) {
  df_plot <- df_long %>% filter(Feature == f)

  # Statistical test (Wilcoxon, Bonferroni)
  stat.test <- ggpubr::compare_means(
    Intensity ~ Treatment,
    data = df_plot,
    method = "wilcox.test",
    ref.group = ref_treatment,
    p.adjust.method = "bonferroni"
  )

  # Assign y positions for stars
  stat.test <- stat.test %>%
    mutate(
      y.position = max(df_plot$Intensity, na.rm = TRUE) + 
        (1:n()) * (0.05 * diff(range(df_plot$Intensity, na.rm = TRUE)))
    )

  # Plot
  p <- ggplot(df_plot, aes(x = Treatment, y = Intensity, fill = Treatment)) +
    geom_violin(trim = FALSE, alpha = 0.35, color = "gray40", width = 0.9) +
    geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7, color = "black") +
    geom_jitter(aes(color = Treatment), size = 1.2, width = 0.2, alpha = 0.4) +
    scale_fill_manual(values = treatment_colors) +
    scale_color_manual(values = treatment_colors) +
    labs(
      title = paste("Single-cell intensity (Healthy) —", f),
      y = "Z-scored log intensity",
      x = "Treatment",
      color = "Treatment",
      fill = "Treatment"
    ) +
    theme_bw(base_size = 18) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 20),
      axis.title = element_text(size = 18, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 16),
      axis.text.y = element_text(size = 14),
      legend.position = "right",
      panel.grid = element_blank()
    ) +
    stat_pvalue_manual(
      stat.test,
      label = "p.signif",  # show stars instead of p values
      size = 6,
      tip.length = 0.01
    )

  print(p)
}

````
#CytoID

````{r}
# Plot histogram for Intensity_MeanIntensity_CorrCytoID
# to select cut off
intensity_data <- as.data.frame(t(assay(sce, "tfmfeatures")))  # or use the assay name you used earlier
metadata <- as.data.frame(colData(sce)[, c("Plate", "Well", "Patient", "Treatment", "Disease")])

# Combine
df <- cbind(metadata, intensity_data)

# 3. Compute 75th percentile per Plate
percentiles <- df %>%
  group_by(Plate) %>%
  summarise(P75 = quantile(Intensity_IntegratedIntensity_CorrCytoID, 0.75, na.rm = TRUE)) #or Intensity_IntegratedIntensity_CorrCytoID, Intensity_MeanIntensity_CorrCytoID

print(percentiles)

# 4. Plot histogram with vertical line per plate
ggplot(df, aes(x = Intensity_IntegratedIntensity_CorrCytoID)) +
  geom_histogram(fill = "#2C7BB6", alpha = 0.5, bins = 40, color = "black") +
  geom_vline(
    data = percentiles,
    aes(xintercept = P75),
    color = "red", linetype = "dashed", size = 1
  ) +
  facet_wrap(~ Plate, scales = "free_y") +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5)
  ) +
  labs(
    title = "Distribution of Intensity (CorrCytoID) per Plate",
    x = "Intensity (CorrCytoID)",
    y = "Cell Count",
    caption = "Red dashed line = 75th percentile"
  )

#plot positive cell >75th percentile

# 3. Assign positivity per cell
df <- df %>%
  left_join(percentiles, by = "Plate") %>%
  mutate(Positive = Intensity_MeanIntensity_CorrCytoID > P75)

# 4. Summarize percent positive per Patient x Treatment x Plate
df_percent <- df %>%
  group_by(Patient, Treatment, Plate, Disease) %>%
  summarise(PercentPositive = mean(Positive) * 100, .groups = "drop")

# 5. Plot percent positive cells, combine both plates
ggplot(df_percent, aes(x = Treatment, y = PercentPositive, fill = Treatment)) +
  stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
  geom_point(aes(shape = Plate, color = Treatment),
             position = position_jitter(width = 0.15),
             size = 3.5, alpha = 0.9) +
  geom_text_repel(
    aes(label = Patient, color = Treatment),
    size = 3.5,
    position = position_jitter(width = 0.15),
    show.legend = FALSE
  ) +
  facet_wrap(~ Disease, scales = "fixed") +
  theme_bw(base_size = 14) +
  labs(
    title = "Percent CytoID Positive Cells per Treatment (Cutoff = 75th percentile per Plate)",
    y = "Percent Positive Cells (%)",
    x = "Treatment"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "gray90", color = "gray60"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )


````
#compare by Disease

````{r}
# 1. Extract features + metadata
intensity_data <- as.data.frame(t(assay(sce, "features")))

# Select only the features you want
selected_features <- c(
  "Intensity_MeanIntensity_CorrTMRM", 
  "Intensity_MeanIntensity_CorrCytoID", 
  "Intensity_MeanIntensity_CorrLyso", 
  "Intensity_IntegratedIntensity_CorrTMRM",
  "Intensity_IntegratedIntensity_CorrCytoID",
  "Intensity_IntegratedIntensity_CorrLyso"
)
intensity_data <- intensity_data[, selected_features]

metadata <- as.data.frame(colData(sce)[, c("Patient", "Treatment", "Disease", "Plate")])

# Combine metadata and intensity data
df <- cbind(metadata, intensity_data)

# Filter only control samples
df_control <- df %>% filter(Treatment == "control")

# Loop through each selected feature
for (feature in selected_features) {
  
  # Summarize mean per patient x disease x plate
  df_summary <- df_control %>%
    group_by(Patient, Disease, Plate) %>%
    summarise(MeanIntensity = mean(.data[[feature]], na.rm = TRUE), .groups = "drop")
  
  # Plot
  p <- ggplot(df_summary, aes(x = Disease, y = MeanIntensity, fill = Disease)) +
    stat_summary(fun = mean, geom = "bar", alpha = 0.4, color = "black", width = 0.6) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.3, color = "black") +
    geom_point(aes(shape = Plate, color = Disease),
               position = position_jitter(width = 0.1),
               size = 3.5) +
    geom_text_repel(
      aes(label = Patient, color = Disease),
      size = 3.5,
      position = position_jitter(width = 0.1),
      show.legend = FALSE
    ) +
    #facet_wrap(~ Plate) + 
    theme_bw(base_size = 14) +
    labs(
      title = paste("Mean", feature, "per Disease (Control Samples)"),
      y = "Mean Intensity (a.u.)",
      x = "Disease"
    ) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold"),
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "right"
    )
  
  # Print plot
  print(p)
}

````
## Unsupervised Analysis PCA

````{r}
sce <- runPCA(sce, exprs_values = "tfmfeatures")
plotReducedDim(sce, dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce, dimred = "PCA", colour_by = "Plate")

sce_select <- sce[str_detect(names(sce), "Intensity"), ]
sce_select <- runPCA(sce_select, exprs_values = "tfmfeatures", ncomponents = 10)
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Treatment")
plotReducedDim(sce_select , dimred = "PCA", colour_by = "Plate")

```

What is driving PC1?

```{r}
plotPCACor(sce_select, filter_by = 1, top = 20)
```

# reduce redundancy in CellProfiler feature
Try with Mitochondria

```{r}

mat <- assay(sce)
corr_features <- mat[grep("CorrTMRM", rownames(mat)), , drop = FALSE]
dim(corr_features)

#Compute feature–feature correlation matrix
# Transpose: we want correlation across features, not cells
cor_mat <- cor(t(corr_features), method = "pearson", use = "pairwise.complete.obs")


#Hierarchical clustering of correlated features
# Hierarchical clustering
hc <- hclust(as.dist(1 - cor_mat), method = "average")

# Visualize
plot(hc, labels = FALSE, main = "Hierarchical clustering of features")

#Cut the dendrogram at a correlation height
# Example: height = 0.3 (usually between 0.3-0.7)

# Cut the dendrogram at correlation 0.8 → distance = 1 - 0.8 = 0.2
clusters <- cutree(hc, h = 0.6)

table(clusters)

#Choose a representative feature per cluster
# Compute variance per feature
feat_var <- apply(corr_features, 1, var)

# Select representative feature per cluster
rep_features <- sapply(unique(clusters), function(cl){
  feats <- names(clusters[clusters == cl])
  feats[which.max(feat_var[feats])]
})

rep_features <- unname(rep_features)

#Check for multicollinearity using VIF
library(car)

# Create data frame of selected features
df_rep <- t(corr_features[rep_features, ])

# Fit a simple model to compute VIF
vif_model <- lm(df_rep[,1] ~ ., data = as.data.frame(df_rep))
vif_values <- vif(vif_model)

# Filter out any with VIF > 10
low_vif_features <- names(vif_values[vif_values < 10])

#subset into sce
sce_reduced <- sce[rownames(sce) %in% rep_features, ]

#Visualise clustered heatmap

library(pheatmap)
pheatmap(
  cor_mat,
  clustering_distance_rows = as.dist(1 - cor_mat),
  clustering_distance_cols = as.dist(1 - cor_mat),
  clustering_method = "average",
  color = colorRampPalette(c("blue", "white", "red"))(100),
  main = "Correlation heatmap of CorrTMRM features",
  show_rownames = FALSE,
  show_colnames = FALSE,
  border_color = NA
)


# corr_features: features x cells
# clusters: cluster assignment for each feature

rep_features <- sapply(unique(clusters), function(cl){
  
  # Get features in this cluster
  feats <- names(clusters[clusters == cl])
  
  # Subset matrix
  sub_mat <- corr_features[feats, , drop = FALSE]
  
  # Compute median across features per cell
  # Then find the feature closest to this median profile
  cluster_median <- apply(sub_mat, 2, median)
  
  # Compute Euclidean distance of each feature to cluster median
  dists <- apply(sub_mat, 1, function(x) sqrt(sum((x - cluster_median)^2)))
  
  # Select feature closest to median
  feats[which.min(dists)]
})

# Unname if needed
rep_features <- unname(rep_features)

# Check how many representative features
length(rep_features)
rep_features

#save file as csv table
rep_features_df <- data.frame(
  Feature = rep_features,
  Cluster = unique(clusters)
)

write.csv(rep_features_df, file = "rep_features_with_clusters_TMRM cut 0.7.csv", row.names = FALSE, quote = FALSE)

```
#Feature importance analysis

```{r}
#Step1:  Subset representative features
mat_rep <- t(assay(sce)[rep_features, , drop = FALSE])

# Get patient labels
labels <- as.factor(colData(sce)$Treatment) 

# Make sure rows and labels match
stopifnot(nrow(mat_rep) == length(labels))

#Step2: Fit logistic regression for each binary task
# List of disease subgroups
subgroups <- setdiff(levels(labels), "control")

# Store results
feature_importance <- list()

library(glmnet)  # optional for regularized logistic regression

for (sg in subgroups) {
  
  # Binary labels: 1 = subgroup, 0 = Control
  binary_labels <- ifelse(labels == sg, 1, 0)
  
  # Only keep cells for Control or this subgroup
  idx <- labels %in% c("control", sg)
  X <- mat_rep[idx, ]
  y <- binary_labels[idx]
  
  # Fit logistic regression (can add L2 penalty if high-dimensional)
  model <- glm(y ~ ., data = as.data.frame(X), family = binomial)
  
  # Extract absolute coefficients (ignore intercept)
  coefs <- abs(coef(model)[-1])
  
  # Store ranked features
  feature_importance[[sg]] <- sort(coefs, decreasing = TRUE)
}

#Step 3: Optional permutation importance
library(pROC)

perm_importance <- list()
set.seed(123)

for (sg in subgroups) {
  
  # Binary labels
  binary_labels <- ifelse(labels == sg, 1, 0)
  idx <- labels %in% c("control", sg)
  X <- mat_rep[idx, ]
  y <- binary_labels[idx]
  
  # Fit logistic regression
  model <- glm(y ~ ., data = as.data.frame(X), family = binomial)
  
  # Baseline AUC
  pred <- predict(model, type = "response")
  auc_base <- auc(y, pred)
  
  # Permutation importance
  imp <- sapply(1:ncol(X), function(j){
    X_perm <- X
    X_perm[, j] <- sample(X_perm[, j])  # shuffle feature
    pred_perm <- predict(model, newdata = as.data.frame(X_perm), type = "response")
    auc(y, pred_perm) - auc_base  # negative drop = importance
  })
  
  perm_importance[[sg]] <- sort(abs(imp), decreasing = TRUE)
}

#Step 4: Visualize top features
# Example for one subgroup
sg <- subgroups[2]

top_coef <- names(feature_importance[[sg]])[1:10]
top_perm <- names(perm_importance[[sg]])[1:10]

df_plot <- data.frame(
  Feature = c(top_coef, top_perm),
  Importance = c(feature_importance[[sg]][top_coef],
                 perm_importance[[sg]][top_perm]),
  Method = rep(c("Coefficient", "Permutation"), each = 10)
)

ggplot(df_plot, aes(x = reorder(Feature, Importance), y = Importance, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge") +
  coord_flip() +
  theme_classic() +
  ggtitle(paste("Top features for", sg, "vs Control"))

````
# Plot to see main repeated feature

```{r}
# Required libraries
library(pROC)
library(cowplot)    # for arranging plots

# PARAMETERS
label_col <- "Treatment"   # change if your metadata column has a different name
control_label <- "control" # assumed control name; script will fall back to first level if not present
nperm <- 30                # number of permutations per feature (increase for stability; slower)
top_n <- 10                # top features to display per method
min_auc <- 0.6             # only show subgroup if AUC > min_auc

# Prepare data matrix: cells x features
mat_rep <- t(assay(sce)[rep_features, , drop = FALSE])   # rows = cells, cols = features
labels <- factor(colData(sce)[[label_col]])

# determine control
if (!control_label %in% levels(labels)) {
  control_label <- levels(labels)[1]
  message("Control label not found; using first level as control: '", control_label, "'")
}
subgroups <- setdiff(levels(labels), control_label)

# Storage for plots
row_plots <- list()

for (sg in subgroups) {
  idx <- which(labels %in% c(control_label, sg))
  X <- as.data.frame(mat_rep[idx, , drop = FALSE])
  y_binary <- ifelse(labels[idx] == sg, 1, 0)

  # fit logistic regression (no regularization). Add tryCatch to handle collinearity issues.
  df_fit <- cbind(y = y_binary, X)
  glm_fit <- tryCatch(
    glm(y ~ ., data = df_fit, family = binomial),
    error = function(e) {
      warning("GLM failed for subgroup ", sg, ": ", e$message)
      return(NULL)
    }
  )
  if (is.null(glm_fit)) next

  # baseline AUC
  pred_base <- predict(glm_fit, type = "response")
  auc_base <- as.numeric(auc(y_binary, pred_base))

  # skip if AUC too low
  if (is.na(auc_base) || auc_base <= min_auc) {
    message("Skipping ", sg, " (AUC = ", round(auc_base, 3), " <= ", min_auc, ")")
    next
  }

  # absolute coefficients (ignore intercept)
  coefs <- coef(glm_fit)[-1]
  abs_coefs <- abs(coefs)
  names(abs_coefs) <- names(coefs)

  # permutation importance: for each feature, compute distribution of AUC drop across nperm shuffles
  set.seed(123)
  perm_results <- lapply(seq_len(ncol(X)), function(j) {
    fname <- colnames(X)[j]
    drops <- numeric(nperm)
    for (p in seq_len(nperm)) {
      Xp <- X
      Xp[[j]] <- sample(Xp[[j]], size = nrow(Xp), replace = FALSE)
      pred_p <- predict(glm_fit, newdata = Xp, type = "response")
      auc_p <- as.numeric(auc(y_binary, pred_p))
      drops[p] <- auc_base - auc_p   # positive drop = feature important
    }
    data.frame(Feature = fname, Drop = drops)
  })
  perm_df <- do.call(rbind, perm_results)

  # Get top features by coefficient and by median permutation importance
  top_coef_feats <- names(sort(abs_coefs, decreasing = TRUE))[1:min(top_n, length(abs_coefs))]
  perm_medians <- perm_df %>% group_by(Feature) %>% summarize(MedianDrop = median(Drop, na.rm=TRUE))
  top_perm_feats <- perm_medians %>% arrange(desc(MedianDrop)) %>% slice(1:min(top_n, n())) %>% pull(Feature)

  # Prepare left plot (coefficients)
  coef_df <- data.frame(Feature = top_coef_feats,
                        Coef = coefs[top_coef_feats]) %>%
    mutate(AbsCoef = abs(Coef)) %>%
    arrange(AbsCoef)
  p_left <- ggplot(coef_df, aes(x = AbsCoef, y = reorder(Feature, AbsCoef))) +
    geom_col() +
    labs(x = "Absolute coefficient", y = NULL, title = "Logistic regression") +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 8), plot.title = element_text(hjust = 0.5))

  # Prepare right plot (permutation importance boxplots)
  perm_plot_df <- perm_df %>% filter(Feature %in% top_perm_feats) %>%
    mutate(Feature = factor(Feature, levels = rev(top_perm_feats)))
  p_right <- ggplot(perm_plot_df, aes(x = Drop, y = Feature)) +
    geom_boxplot(outlier.size = 1) +
    labs(x = "AUC drop (baseline - permuted)", y = NULL, title = "Permutation importance") +
    theme_minimal() +
    theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(), plot.title = element_text(hjust = 0.5))

  # Combine left+right into a single row for this subgroup
  subtitle <- paste0(control_label, " vs ", sg, " (AUC = ", round(auc_base,3), ")")
  row_combined <- plot_grid(
    p_left, p_right, ncol = 2, rel_widths = c(0.5, 0.5),
    labels = c("", ""), align = "h"
  )
  row_with_label <- plot_grid(NULL, ggdraw() + draw_label(subtitle, fontface = "bold"), row_combined, ncol = 1, rel_heights = c(0.06, 0.06, 1))
  row_plots[[sg]] <- row_with_label
}

# Arrange all rows vertically
if (length(row_plots) == 0) {
  stop("No subgroup rows to plot (maybe all AUCs <= min_auc).")
}
final_plot <- plot_grid(plotlist = row_plots, ncol = 1)

# Save the final figure
ggsave("FeatureImportance_grid_TMRM cut0.6.png", final_plot, width = 12, height = 4 * length(row_plots), dpi = 200)
message("Saved plot to FeatureImportance_grid.png")

````
Code for violin plots of top features.

```{r}


# --- 1. Extract the features and metadata from your SCE ---
# Replace "tfmfeatures" with your assay name if needed
mat <- assay(sce, "tfmfeatures")

# Select the features of interest
selected_features <- c(
  "Correlation_Manders_CorrTMRM_CorrLyso",
  "Intensity_MeanIntensityEdge_CorrTMRM"
)

# Check they exist
selected_features <- selected_features[selected_features %in% rownames(mat)]

# Combine metadata and feature matrix
df <- data.frame(
  as.data.frame(colData(sce)),
  t(mat[selected_features, , drop = FALSE])
)

# Ensure metadata columns exist
stopifnot(all(c("Patient", "Treatment", "Disease") %in% colnames(df)))

# --- 2. Aggregate per Patient, Treatment, Disease ---
df_patient <- df %>%
  group_by(Patient, Treatment, Disease) %>%
  summarise(across(all_of(selected_features), median, na.rm = TRUE), .groups = "drop")

# --- 3. Long format for ggplot ---
df_long <- df_patient %>%
  pivot_longer(
    cols = all_of(selected_features),
    names_to = "Feature",
    values_to = "Value"
  )

# --- 4. Plot per patient ---
ggplot(df_long, aes(x = Disease, y = Value, fill = Disease)) +
  geom_violin(scale = "width", trim = FALSE, alpha = 0.7) +
  geom_boxplot(width = 0.1, outlier.shape = NA, color = "black", alpha = 0.5) +
  geom_point(
    aes(color = Patient),
    position = position_jitter(width = 0.1),
    size = 1.5,
    alpha = 0.8
  ) +
  facet_grid(Feature ~ Treatment, scales = "free_y") +
  theme_bw(base_size = 12) +
  labs(
    title = "Median feature values per Patient across Disease and Treatment",
    x = "Disease group",
    y = "Median feature value per patient"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text.y = element_text(face = "bold"),
    strip.text.x = element_text(face = "bold"),
    legend.position = "none"
  )

````
